### 2022-07-09 paper/main.tex
start

% To do:
% ------
% - Write a title and abstract.
% - Write an outline.
% - Fill in the outline.
% - Publish.
% - Wait for the call from Stockholm.

% Style notes:
% ------------
% - Uhhh.

\documentclass[modern]{aastex631}
\usepackage[utf8]{inputenc}

% Hogg typesetting issues
\addtolength{\textheight}{0.8in}
\addtolength{\topmargin}{-0.4in}
\setlength{\parindent}{1.2\baselineskip} % seriously
\frenchspacing\raggedbottom\sloppy\sloppypar

% Math issues
\newcommand{\teff}{T_{\mathrm{eff}}}
\newcommand{\logg}{\log g}
\newcommand{\feh}{[\mathrm{Fe}/\mathrm{H}]}

\shorttitle{machine learning and science}
\shortauthors{hogg}

\begin{document}

\title{Is machine learning good or bad for science?}

\author{David W. Hogg}

\begin{abstract}\noindent % seriously!
  Machine-learning methods are having a huge impact across the natural sciences.
  Advances have come from classification, regression, dimensionality reduction, clustering, and many other things.
  However, machine learning has a strong ontology---for machine-learning methods, only the data exist---and a strong epistemology---models are generally considered good if they perform well on held-out training data---and these philosophies are in strong conflict from standard practice and philosophy in the natural sciences.
  Here I identify uses for machine learning in the natural sciences where the ontology and epistemology are valuable, and will help scientific projects.
  I also call out uses for machine learning that are not a good idea.
  In particular, there are a lot of reasons to be concerned about the use of complex machine-learning models to emulate physical (or first-principles) simulations.
  Finally, I conclude by noting that there are places in scientific investigations in which the choice to use a machine-learning method is the most conservative possible choice.
  These places are generally in projects that look like causal inferences, where the machine-learning method is used to model the possible effects of confounders.
  I illustrate the above points with examples from astrophysics.
\end{abstract}

\section{Introduction}\label{sec:intro}

Hello World!

\section{What is machine learning?}\label{sec:what}

\section{Stuff in between}\label{sec:stuff}

\section{Discussion}\label{sec:discussion}

Hello World!

\bibliography{sample631}{}
\bibliographystyle{aasjournal}

\end{document}

### 2022-12-05 paper/main.tex
change language to astro; introduction text

@@
 
-\shorttitle{machine learning and science}
+\shorttitle{machine learning and astrophysics}
 \shortauthors{hogg}
@@
 
-\title{Is machine learning good or bad for science?}
+\title{Is machine learning good or bad for astrophysics?}
 
@@
 \begin{abstract}\noindent % seriously!
-  Machine-learning methods are having a huge impact across the natural sciences.
+  Machine-learning methods are having a huge impact across the sciences.
   Advances have come from classification, regression, dimensionality reduction, clustering, and many other things.
-  However, machine learning has a strong ontology---for machine-learning methods, only the data exist---and a strong epistemology---models are generally considered good if they perform well on held-out training data---and these philosophies are in strong conflict from standard practice and philosophy in the natural sciences.
-  Here I identify uses for machine learning in the natural sciences where the ontology and epistemology are valuable, and will help scientific projects.
+  However, machine learning has a strong ontology---for machine-learning methods, only the data exist---and a strong epistemology---models are generally considered good if they perform well on held-out training data---and these philosophies are in strong conflict from standard practice and philosophy in the natural sciences, and astrophysics in particular.
+  Here I identify uses for machine learning in astrophysics where the ontology and epistemology (and computational performance) are valuable.
   I also call out uses for machine learning that are not a good idea.
-  In particular, there are a lot of reasons to be concerned about the use of complex machine-learning models to emulate physical (or first-principles) simulations.
-  Finally, I conclude by noting that there are places in scientific investigations in which the choice to use a machine-learning method is the most conservative possible choice.
-  These places are generally in projects that look like causal inferences, where the machine-learning method is used to model the possible effects of confounders.
-  I illustrate the above points with examples from astrophysics.
+  In particular, there are a lot of reasons to be concerned about the use of complex machine-learning models to emulate physical (or first-principles) simulations, for epistemological and confirmation-bias reasons.
+  I conclude by noting that there are places in which the choice to use a machine-learning method is (oddly) the most conservative possible choice.
+  These places are generally in projects that look like causal inferences, where the machine-learning method is used to model the possible effects of confounders, such as foregrounds, backgrounds, or tellurics.
 \end{abstract}
@@
 
-Hello World!
+It is an understatement to say that machine learning---and in particular deep learning---is having a big impact across the sciences.
+There are hundreds of papers per year on machine-learning methods in astrophysics alone.
+However, when we ask what astrophysics breakthroughs have been enabled by this influx of new tools and methods, there isn't a long list.
+The success of the AlphaFold projects in protein structure are often raised.
+But these are successes in a biological field, not astrophysics, and in an area where performance is valued over understanding.
+Here in astrophysics we care almost exclusively about understanding.
+
+That is, astrophysics is an academic discipline, concerned with understanding the physical world, and naturally occurring mechanisms in play in that world.
+We make progress by discovering new kinds of objects and phenomena, and explaining (and, even better, predicting) new kinds of objects and phenomena.
+Our most successful investigations are judged in terms of the questions they answer, or the new questions they raise, or both.
+How can machine learning contribute to this mission?
+
+I mentioned the word ``performance'' above.
+One of the key ideas of machine-learning methods is that they are usually judged in terms of their performance, or their ability to predict or explain new data.
+That's different, philosophically, from what's traditionally important in astrophysics.
+When the expansion of the Universe was discovered (\citealt{expansion}), the result was not (primarily) judged on the basis of its ability to predict new data.
+That is, the investigators were not mainly trying to predict the Doppler shifts of held-out galaxies!
+They were trying to measure a property of the Universe, traced by galaxies.
+And their result answered some important questions, and raised new ones.
+Of course the expansion result has lasted in part because it \emph{does} make new predictions for new data, and very accurate predictions!
+So it isn't irrelevant that the performance of the hypothesis is good.
+However, there is some disconnect between how we assess a machine-learning method and how we assess a scientific contribution.
+
+I don't want to sound anti-technological however.
+Machine learning is a new and productive technology.
+We should use it!
+Astrophysics moves forward through technological progress, starting with Galileo (CITE).
+In my own lifetime, the most remarkable technological advance was (probably) the change from photographic plates to charge-coupled devices (CCDs).
+In just a few years, almost every telescope and spectrograph in the world switched from a photographic focal plane to a CCD focal plane, and immediately there were new discoveries.
+That's a case in which new technology immediately translated into new discoveries.
+That hasn't been nearly so true with machine learning, and (I note with respect), it is way easier to incorporate machine learning into your workflow than to incorporate CCDs into your optical path.
+
+I am both a developer of machine-learning methods, and a skeptic who thinks they should only rarely be used.
+That said, there are reasons that we absolutely must use machine learning in the near future of astrophysics.
+And there are indeed past successes of machine learning that might be invisible to many.
+I'll try to speak to both of these points below.
 
@@
 
-\section{Stuff in between}\label{sec:stuff}
+There is a vague definition of machine learning in terms of data and capacity:
+A method is a machine-learning method if its capability increases (hopefully dramatically) as the machine sees more data (HOGG GET THIS MORE PRECISE AND CITE).
+In some sense this is true of any measurement process:
+The measurement improves as more data come in!
+So if we are going to be more specific, we'd like the model capacity or capability to improve faster (in some sense) than something like the square root of the increment in data.
+
+There is a more specific definition of machine learning in terms of specific methods:
+Whenever a neural network or a Gaussian process is in play, machine learning is in play.
+But when the former definition is too broad, the latter is almost always too narrow.
+For example, principal components analysis (PCA) is a workhorse in astronomy since way back (for example, \citealt{pcaredshift}), and PCA is an old, and simple, but definitely machine-learning, method.
+
+Supervised and unsupervised.
+
+Classification, regression.
+
+Dimensionality reduction, clustering, density estimation.
+
+\section{Why do we need machine learning?}
+
+Label transfer.
+
+Speeding up decisions.
+
+Modeling nuisances.
+
+Classification?
+
+Speeding up simulations?
+
+The short summary is that we need machine learning; we can't live without it.
+
+\section{When is machine learning a bad idea?}
+
+Adversarial attacks argument.
+
+\section{When is machine learning the conservative choice?}
 

### 2022-12-12 paper/main.tex
start bib

@@
 \setlength{\parindent}{1.2\baselineskip} % seriously
+\renewcommand{\twocolumngrid}{} % omg
 \frenchspacing\raggedbottom\sloppy\sloppypar
@@
 
-\author{David W. Hogg}
+\author[0000-0003-2866-9403]{David W. Hogg}
+\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726~Broadway, New~York,~NY 10003, USA}
+\affiliation{Max-Planck-Institut f{\"u}r Astronomie, K{\"o}nigstuhl 17, D-69117 Heidelberg, Germany}
+\affiliation{Flatiron Institute, 162 Fifth Avenue, New~York,~NY 10010, USA}
 
@@
   Advances have come from classification, regression, dimensionality reduction, clustering, and many other things.
-  However, machine learning has a strong ontology---for machine-learning methods, only the data exist---and a strong epistemology---models are generally considered good if they perform well on held-out training data---and these philosophies are in strong conflict from standard practice and philosophy in the natural sciences, and astrophysics in particular.
+  However, machine learning has a strong ontology, in which only the data exist, and a strong epistemology, in which a model is considered good if it performs well on held-out training data.
+  These philosophies are in strong conflict from standard practice and philosophy in the natural sciences, and astrophysics in particular.
   Here I identify uses for machine learning in astrophysics where the ontology and epistemology (and computational performance) are valuable.
@@
   In particular, there are a lot of reasons to be concerned about the use of complex machine-learning models to emulate physical (or first-principles) simulations, for epistemological and confirmation-bias reasons.
-  I conclude by noting that there are places in which the choice to use a machine-learning method is (oddly) the most conservative possible choice.
-  These places are generally in projects that look like causal inferences, where the machine-learning method is used to model the possible effects of confounders, such as foregrounds, backgrounds, or tellurics.
+  I conclude by noting that there are places in which the choice to use a machine-learning method is (oddly) the most conservative possible choice:
+  When a project makes use of a machine-learning method to model the possible effects of nuisances, such as foregrounds, backgrounds, or tellurics, the immense flexibility and loose philosophy in fact make the project more trustworthy.
 \end{abstract}
@@
 There are hundreds of papers per year on machine-learning methods in astrophysics alone.
-However, when we ask what astrophysics breakthroughs have been enabled by this influx of new tools and methods, there isn't a long list.
-The success of the AlphaFold projects in protein structure are often raised.
-But these are successes in a biological field, not astrophysics, and in an area where performance is valued over understanding.
+However, when we ask what breakthroughs have been enabled by this influx of new tools and methods, there isn't a long list.
+The success of the AlphaFold projects in protein structure (\citealt{alphafold}) are often raised.
+But these are successes in a biological or medical field, not astrophysics, and in a very specific challenge-problem setting in which performance is valued over understanding.
 Here in astrophysics we care almost exclusively about understanding.
@@
 That's different, philosophically, from what's traditionally important in astrophysics.
-When the expansion of the Universe was discovered (\citealt{expansion}), the result was not (primarily) judged on the basis of its ability to predict new data.
+When the expansion of the Universe was discovered (\citealt{expansion, expansion2}), the result was not (primarily) judged on the basis of its ability to predict new data.
 That is, the investigators were not mainly trying to predict the Doppler shifts of held-out galaxies!
@@
 
+HOGG: NEED to make the point, here or below, that I respect the ML point of view.
+It could be helpful for astrophysics if we moved towards it in some ways.
+Oh and the statistical care is of interest too.
+Maybe this is a section about the underlying philosophies of ML.
+
 I don't want to sound anti-technological however.
@@
 We should use it!
-Astrophysics moves forward through technological progress, starting with Galileo (CITE).
-In my own lifetime, the most remarkable technological advance was (probably) the change from photographic plates to charge-coupled devices (CCDs).
+Astrophysics has moved forward through technological progress, starting with Galileo (\citealt{galileo}).
+In my own lifetime, the most remarkable technological advance was (probably) the change from photographic plates to charge-coupled devices (CCDs; see, for an early review, \citealt{ccd}).
 In just a few years, almost every telescope and spectrograph in the world switched from a photographic focal plane to a CCD focal plane, and immediately there were new discoveries.
@@
 That said, there are reasons that we absolutely must use machine learning in the near future of astrophysics.
-And there are indeed past successes of machine learning that might be invisible to many.
+And there are indeed successes of machine learning that might seem invisible.
 I'll try to speak to both of these points below.
@@
 
+\section{What does machine learning do well?}
+
 \section{When is machine learning a bad idea?}
 
-Adversarial attacks argument.
+Adversarial attacks argument. Confirmation bias argument. Planet search argument.
 
 \section{When is machine learning the conservative choice?}
+
+Come back to the planet-search example.
 
@@
 
-\bibliography{sample631}{}
+\bibliography{ml_in_astro}
 \bibliographystyle{aasjournal}

### 2023-01-13 paper/main.tex
update

@@
 
-\title{Is machine learning good or bad for astrophysics?}
+\title{Is deep learning good or bad for astrophysics?}
 
@@
 \begin{abstract}\noindent % seriously!
-  Machine-learning methods are having a huge impact across the sciences.
-  Advances have come from classification, regression, dimensionality reduction, clustering, and many other things.
-  However, machine learning has a strong ontology, in which only the data exist, and a strong epistemology, in which a model is considered good if it performs well on held-out training data.
+  Deep-learning methods are having a huge impact across the sciences.
+  Advances have come from classification, regression, dimensionality reduction, clustering(?), and many other things.
+  However, deep learning (and machine learning more generally) has a strong ontology, in which only the data exist, and a strong epistemology, in which a model is considered good if (and only if) it performs well on held-out data that is similar to the training data.
   These philosophies are in strong conflict from standard practice and philosophy in the natural sciences, and astrophysics in particular.
-  Here I identify uses for machine learning in astrophysics where the ontology and epistemology (and computational performance) are valuable.
-  I also call out uses for machine learning that are not a good idea.
+  Here I identify uses for deep learning in astrophysics where the ontology and epistemology (and computational performance) are valuable.
+  I also call out uses for deep learning that are not a good idea.
   In particular, there are a lot of reasons to be concerned about the use of complex machine-learning models to emulate physical (or first-principles) simulations, for epistemological and confirmation-bias reasons.
-  I conclude by noting that there are places in which the choice to use a machine-learning method is (oddly) the most conservative possible choice:
-  When a project makes use of a machine-learning method to model the possible effects of nuisances, such as foregrounds, backgrounds, or tellurics, the immense flexibility and loose philosophy in fact make the project more trustworthy.
+  On the other hand, there are places in which the introduction of deep learning can make a project more scientifically conservative:
+  When a flexible machine-learning model is used to represent the effects of nuisances, such as foregrounds, backgrounds, or tellurics, the model capacity and loose philosophy of the method can in fact make the project more trustworthy.
 \end{abstract}
@@
 
-It is an understatement to say that machine learning---and in particular deep learning---is having a big impact across the sciences.
-There are hundreds of papers per year on machine-learning methods in astrophysics alone.
-However, when we ask what breakthroughs have been enabled by this influx of new tools and methods, there isn't a long list.
+It is an understatement to say that deep learning is having a big impact across the sciences.
+There are hundreds of papers per year now on deep-learning methods in astrophysics alone.
+However, when we ask what scientific breakthroughs have been enabled by this influx of new tools and methods, there isn't a long list.
 The success of the AlphaFold projects in protein structure (\citealt{alphafold}) are often raised.
@@
 
-That is, astrophysics is an academic discipline, concerned with understanding the physical world, and naturally occurring mechanisms in play in that world.
+If we broaden our scope from deep learning to machine learning more generally, we can point to very important contributions of machine learning to astrophysics.
+Kepler discoveries
+PCA and, eg, SDSS redshift system
+
+Astrophysics is an academic discipline, concerned with understanding the physical world, and naturally occurring mechanisms in play in that world.
 We make progress by discovering new kinds of objects and phenomena, and explaining (and, even better, predicting) new kinds of objects and phenomena.
 Our most successful investigations are judged in terms of the questions they answer, or the new questions they raise, or both.
-How can machine learning contribute to this mission?
+The question here is: How can deep learning contribute to this mission?
 
@@
 
+Something about how we will use the word ``model'' in what follows.
+
 \section{Why do we need machine learning?}
@@
 
-\section{What does machine learning do well?}
+\section{The ontology and epistemology of machine learning}
+
+HOGG: Definitions of latent variables and so on; what are latent variables and why are they important?
+
+Contemporary machine-learning models---and especially deep-learning models---have explicit unbroken, combinatorically large degeneracies.
+That is, the weights and nodes in various layers of the network have permutation symmetries such that a very large space of permutation operators can be applied to the latent space without making any changes whatsoever to the input--output relationships of the model.
+That shows that there is no belief in the meaning of any internal latent variables.
+
+Furthermore, stochastic gradient, early stopping and so on:
+There isn't even a belief that the latents should be the actual argmin of anything.
+The latent values are of absolutely no importance or meaning whatsoever.
+
+If there is any stability to a model, it is the stability of its predictions, or its behavior on held-out data.
+
+Also maybe mention here the statistical philosophy or care.
 

### 2023-12-02 paper/main.tex
new abstract; start to reformat for ICML Position Paper

@@
 
-\documentclass[modern]{aastex631}
+\documentclass[11pt]{article}
 \usepackage[utf8]{inputenc}
+\usepackage{microtype}
+\usepackage[letterpaper]{geometry}
 
 % Hogg typesetting issues
-\addtolength{\textheight}{0.8in}
-\addtolength{\topmargin}{-0.4in}
+\setlength{\textwidth}{6.50in}
+\setlength{\oddsidemargin}{0.00in}
+\setlength{\textheight}{9.00in}
+\setlength{\topmargin}{-0.50in}
 \setlength{\parindent}{1.2\baselineskip} % seriously
-\renewcommand{\twocolumngrid}{} % omg
+\linespread{1.08}
 \frenchspacing\raggedbottom\sloppy\sloppypar
+\renewcommand{\author}[1]{\medskip\par\noindent\textbf{#1}}
+\newcommand{\affil}[1]{{\footnotesize\par\noindent #1 \par}}
+
+% Gross
+\newcommand{\apj}{The Astrophysical Journal}
+\newcommand{\araa}{Annual Reviews in Astronomy and Astrophysics}
 
@@
 
-\shorttitle{machine learning and astrophysics}
-\shortauthors{hogg}
-
 \begin{document}
 
-\title{Is deep learning good or bad for astrophysics?}
+\section*{Is machine learning good or bad for the natural sciences?}
 
-\author[0000-0003-2866-9403]{David W. Hogg}
-\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726~Broadway, New~York,~NY 10003, USA}
-\affiliation{Max-Planck-Institut f{\"u}r Astronomie, K{\"o}nigstuhl 17, D-69117 Heidelberg, Germany}
-\affiliation{Flatiron Institute, 162 Fifth Avenue, New~York,~NY 10010, USA}
+\author{David W. Hogg}
+\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726~Broadway, New~York,~NY 10003, USA}
+\affil{Center for Data Science, New York University, 60 Fifth Avenue, New York, NY 10011, USA}
+\affil{Max-Planck-Institut f{\"u}r Astronomie, K{\"o}nigstuhl 17, D-69117 Heidelberg, Germany}
+\affil{Flatiron Institute, a division of the Simons Foundation, 162 Fifth Avenue, New~York,~NY 10010, USA}
 
-\begin{abstract}\noindent % seriously!
-  Deep-learning methods are having a huge impact across the sciences.
-  Advances have come from classification, regression, dimensionality reduction, clustering(?), and many other things.
-  However, deep learning (and machine learning more generally) has a strong ontology, in which only the data exist, and a strong epistemology, in which a model is considered good if (and only if) it performs well on held-out data that is similar to the training data.
-  These philosophies are in strong conflict from standard practice and philosophy in the natural sciences, and astrophysics in particular.
-  Here I identify uses for deep learning in astrophysics where the ontology and epistemology (and computational performance) are valuable.
-  I also call out uses for deep learning that are not a good idea.
-  In particular, there are a lot of reasons to be concerned about the use of complex machine-learning models to emulate physical (or first-principles) simulations, for epistemological and confirmation-bias reasons.
-  On the other hand, there are places in which the introduction of deep learning can make a project more scientifically conservative:
-  When a flexible machine-learning model is used to represent the effects of nuisances, such as foregrounds, backgrounds, or tellurics, the model capacity and loose philosophy of the method can in fact make the project more trustworthy.
-\end{abstract}
+\paragraph{Abstract}
+  Machine learning (ML) methods are having a huge impact across all of the sciences.
+  Advances have come from classification, regression, dimensionality reduction, clustering, emulation, and many other things.
+  However, ML has a strong ontology---in which only the data exist---and a strong epistemology---in which a model is considered good if it performs well on held-out training data.
+  These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
+  Here I identify locations for ML in the natural sciences where the ontology and epistemology (and computational performance) are valuable.
+  I also call out uses for ML that are not a good idea.
+  In particular, there are a lot of reasons to be concerned about the use of complex ML models to emulate physical (or first-principles) simulations, for epistemological and confirmation-bias reasons.
+  On the other hand, there are also places where the introduction of ML can in fact make a project more scientifically conservative:
+  When a flexible machine-learning model is used to represent the effects of nuisances, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of the method can sometimes make the results more trustworthy.
+  These remarks are intended to apply to all of the natural sciences, but most of the specific examples are in the astrophysics domain.
 
@@
 However, when we ask what scientific breakthroughs have been enabled by this influx of new tools and methods, there isn't a long list.
-The success of the AlphaFold projects in protein structure (\citealt{alphafold}) are often raised.
+The success of the AlphaFold projects in protein structure (\cite{alphafold}) are often raised.
 But these are successes in a biological or medical field, not astrophysics, and in a very specific challenge-problem setting in which performance is valued over understanding.
@@
 That's different, philosophically, from what's traditionally important in astrophysics.
-When the expansion of the Universe was discovered (\citealt{expansion, expansion2}), the result was not (primarily) judged on the basis of its ability to predict new data.
+When the expansion of the Universe was discovered (\cite{expansion, expansion2}), the result was not (primarily) judged on the basis of its ability to predict new data.
 That is, the investigators were not mainly trying to predict the Doppler shifts of held-out galaxies!
@@
 We should use it!
-Astrophysics has moved forward through technological progress, starting with Galileo (\citealt{galileo}).
-In my own lifetime, the most remarkable technological advance was (probably) the change from photographic plates to charge-coupled devices (CCDs; see, for an early review, \citealt{ccd}).
+Astrophysics has moved forward through technological progress, starting with Galileo (\cite{galileo}).
+In my own lifetime, the most remarkable technological advance was (probably) the change from photographic plates to charge-coupled devices (CCDs; see, for an early review, \cite{ccd}).
 In just a few years, almost every telescope and spectrograph in the world switched from a photographic focal plane to a CCD focal plane, and immediately there were new discoveries.
@@
 But when the former definition is too broad, the latter is almost always too narrow.
-For example, principal components analysis (PCA) is a workhorse in astronomy since way back (for example, \citealt{pcaredshift}), and PCA is an old, and simple, but definitely machine-learning, method.
+For example, principal components analysis (PCA) is a workhorse in astronomy since way back (for example, \cite{pcaredshift}), and PCA is an old, and simple, but definitely machine-learning, method.
 
@@
 \bibliography{ml_in_astro}
-\bibliographystyle{aasjournal}
+\bibliographystyle{plain}
 

### 2023-12-16 paper/example_paper.tex
Updates from Overleaf

%%%%%%%% ICML 2024 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2024} with \usepackage[nohyperref]{icml2024} above.
\usepackage{hyperref}

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
\usepackage{icml2024}

% If accepted, instead use the following line for the camera-ready submission:
% \usepackage[accepted]{icml2024}

% For theorems and such
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}

% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Submission and Formatting Instructions for ICML 2024}

\begin{document}

\twocolumn[
\icmltitle{Submission and Formatting Instructions for \\
           International Conference on Machine Learning (ICML 2024)}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2024
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Firstname1 Lastname1}{equal,yyy}
\icmlauthor{Firstname2 Lastname2}{equal,yyy,comp}
\icmlauthor{Firstname3 Lastname3}{comp}
\icmlauthor{Firstname4 Lastname4}{sch}
\icmlauthor{Firstname5 Lastname5}{yyy}
\icmlauthor{Firstname6 Lastname6}{sch,yyy,comp}
\icmlauthor{Firstname7 Lastname7}{comp}
%\icmlauthor{}{sch}
\icmlauthor{Firstname8 Lastname8}{sch}
\icmlauthor{Firstname8 Lastname8}{yyy,comp}
%\icmlauthor{}{sch}
%\icmlauthor{}{sch}
\end{icmlauthorlist}

\icmlaffiliation{yyy}{Department of XXX, University of YYY, Location, Country}
\icmlaffiliation{comp}{Company Name, Location, Country}
\icmlaffiliation{sch}{School of ZZZ, Institute of WWW, Location, Country}

\icmlcorrespondingauthor{Firstname1 Lastname1}{first1.last1@xxx.edu}
\icmlcorrespondingauthor{Firstname2 Lastname2}{first2.last2@www.uk}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Machine Learning, ICML}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

\begin{abstract}
This document provides a basic paper template and submission guidelines.
Abstracts must be a single paragraph, ideally between 4--6 sentences long.
Gross violations will trigger corrections at the camera-ready phase.
\end{abstract}

\section{Electronic Submission}
\label{submission}

Submission to ICML 2024 will be entirely electronic, via a web site
(not email). Information about the submission process and \LaTeX\ templates
are available on the conference web site at:
\begin{center}
\textbf{\texttt{http://icml.cc/}}
\end{center}

The guidelines below will be enforced for initial submissions and
camera-ready copies. Here is a brief summary:
\begin{itemize}
\item Submissions must be in PDF\@. 
\item \textbf{New to this year}: If your paper has appendices, submit the appendix together with the main body and the references \textbf{as a single file}. Reviewers will not look for appendices as a separate PDF file. So if you submit such an extra file, reviewers will very likely miss it.
\item Page limit: The main body of the paper has to be fitted to 8 pages, excluding references and appendices; the space for the latter two is not limited. For the final version of the paper, authors can add one extra page to the main body.
\item \textbf{Do not include author information or acknowledgements} in your
    initial submission.
\item Your paper should be in \textbf{10 point Times font}.
\item Make sure your PDF file only uses Type-1 fonts.
\item Place figure captions \emph{under} the figure (and omit titles from inside
    the graphic file itself). Place table captions \emph{over} the table.
\item References must include page numbers whenever possible and be as complete
    as possible. Place multiple citations in chronological order.
\item Do not alter the style template; in particular, do not compress the paper
    format by reducing the vertical spaces.
\item Keep your abstract brief and self-contained, one paragraph and roughly
    4--6 sentences. Gross violations will require correction at the
    camera-ready phase. The title should have content words capitalized.
\end{itemize}

\subsection{Submitting Papers}

\textbf{Paper Deadline:} The deadline for paper submission that is
advertised on the conference website is strict. If your full,
anonymized, submission does not reach us on time, it will not be
considered for publication.

\textbf{Anonymous Submission:} ICML uses double-blind review: no identifying
author information may appear on the title page or in the paper
itself. \cref{author info} gives further details.

\textbf{Simultaneous Submission:} ICML will not accept any paper which,
at the time of submission, is under review for another conference or
has already been published. This policy also applies to papers that
overlap substantially in technical content with conference papers
under review or previously published. ICML submissions must not be
submitted to other conferences and journals during ICML's review
period.
%Authors may submit to ICML substantially different versions of journal papers
%that are currently under review by the journal, but not yet accepted
%at the time of submission.
Informal publications, such as technical
reports or papers in workshop proceedings which do not appear in
print, do not fall under these restrictions.

\medskip

Authors must provide their manuscripts in \textbf{PDF} format.
Furthermore, please make sure that files contain only embedded Type-1 fonts
(e.g.,~using the program \texttt{pdffonts} in linux or using
File/DocumentProperties/Fonts in Acrobat). Other fonts (like Type-3)
might come from graphics files imported into the document.

Authors using \textbf{Word} must convert their document to PDF\@. Most
of the latest versions of Word have the facility to do this
automatically. Submissions will not be accepted in Word format or any
format other than PDF\@. Really. We're not joking. Don't send Word.

Those who use \textbf{\LaTeX} should avoid including Type-3 fonts.
Those using \texttt{latex} and \texttt{dvips} may need the following
two commands:

{\footnotesize
\begin{verbatim}
dvips -Ppdf -tletter -G0 -o paper.ps paper.dvi
ps2pdf paper.ps
\end{verbatim}}
It is a zero following the ``-G'', which tells dvips to use
the config.pdf file. Newer \TeX\ distributions don't always need this
option.

Using \texttt{pdflatex} rather than \texttt{latex}, often gives better
results. This program avoids the Type-3 font problem, and supports more
advanced features in the \texttt{microtype} package.

\textbf{Graphics files} should be a reasonable size, and included from
an appropriate format. Use vector formats (.eps/.pdf) for plots,
lossless bitmap formats (.png) for raster graphics with sharp lines, and
jpeg for photo-like images.

The style file uses the \texttt{hyperref} package to make clickable
links in documents. If this causes problems for you, add
\texttt{nohyperref} as one of the options to the \texttt{icml2024}
usepackage statement.

\subsection{Submitting Final Camera-Ready Copy}

The final versions of papers accepted for publication should follow the
same format and naming convention as initial submissions, except that
author information (names and affiliations) should be given. See
\cref{final author} for formatting instructions.

The footnote, ``Preliminary work. Under review by the International
Conference on Machine Learning (ICML). Do not distribute.'' must be
modified to ``\textit{Proceedings of the
$\mathit{41}^{st}$ International Conference on Machine Learning},
Vienna, Austria, PMLR 235, 2024.
Copyright 2024 by the author(s).''

For those using the \textbf{\LaTeX} style file, this change (and others) is
handled automatically by simply changing
$\mathtt{\backslash usepackage\{icml2024\}}$ to
$$\mathtt{\backslash usepackage[accepted]\{icml2024\}}$$
Authors using \textbf{Word} must edit the
footnote on the first page of the document themselves.

Camera-ready copies should have the title of the paper as running head
on each page except the first one. The running title consists of a
single line centered above a horizontal rule which is $1$~point thick.
The running head should be centered, bold and in $9$~point type. The
rule should be $10$~points above the main text. For those using the
\textbf{\LaTeX} style file, the original title is automatically set as running
head using the \texttt{fancyhdr} package which is included in the ICML
2024 style file package. In case that the original title exceeds the
size restrictions, a shorter form can be supplied by using

\verb|\icmltitlerunning{...}|

just before $\mathtt{\backslash begin\{document\}}$.
Authors using \textbf{Word} must edit the header of the document themselves.

\section{Format of the Paper}

All submissions must follow the specified format.

\subsection{Dimensions}

The text of the paper should be formatted in two columns, with an
overall width of 6.75~inches, height of 9.0~inches, and 0.25~inches
between the columns. The left margin should be 0.75~inches and the top
margin 1.0~inch (2.54~cm). The right and bottom margins will depend on
whether you print on US letter or A4 paper, but all final versions
must be produced for US letter size.
Do not write anything on the margins.

The paper body should be set in 10~point type with a vertical spacing
of 11~points. Please use Times typeface throughout the text.

\subsection{Title}

The paper title should be set in 14~point bold type and centered
between two horizontal rules that are 1~point thick, with 1.0~inch
between the top rule and the top edge of the page. Capitalize the
first letter of content words and put the rest of the title in lower
case.

\subsection{Author Information for Submission}
\label{author info}

ICML uses double-blind review, so author information must not appear. If
you are using \LaTeX\/ and the \texttt{icml2024.sty} file, use
\verb+\icmlauthor{...}+ to specify authors and \verb+\icmlaffiliation{...}+ to specify affiliations. (Read the TeX code used to produce this document for an example usage.) The author information
will not be printed unless \texttt{accepted} is passed as an argument to the
style file.
Submissions that include the author information will not
be reviewed.

\subsubsection{Self-Citations}

If you are citing published papers for which you are an author, refer
to yourself in the third person. In particular, do not use phrases
that reveal your identity (e.g., ``in previous work \cite{langley00}, we
have shown \ldots'').

Do not anonymize citations in the reference section. The only exception are manuscripts that are
not yet published (e.g., under submission). If you choose to refer to
such unpublished manuscripts \cite{anonymous}, anonymized copies have
to be submitted
as Supplementary Material via OpenReview\@. However, keep in mind that an ICML
paper should be self contained and should contain sufficient detail
for the reviewers to evaluate the work. In particular, reviewers are
not required to look at the Supplementary Material when writing their
review (they are not required to look at more than the first $8$ pages of the submitted document).

\subsubsection{Camera-Ready Author Information}
\label{final author}

If a paper is accepted, a final camera-ready copy must be prepared.
%
For camera-ready papers, author information should start 0.3~inches below the
bottom rule surrounding the title. The authors' names should appear in 10~point
bold type, in a row, separated by white space, and centered. Author names should
not be broken across lines. Unbolded superscripted numbers, starting 1, should
be used to refer to affiliations.

Affiliations should be numbered in the order of appearance. A single footnote
block of text should be used to list all the affiliations. (Academic
affiliations should list Department, University, City, State/Region, Country.
Similarly for industrial affiliations.)

Each distinct affiliations should be listed once. If an author has multiple
affiliations, multiple superscripts should be placed after the name, separated
by thin spaces. If the authors would like to highlight equal contribution by
multiple first authors, those authors should have an asterisk placed after their
name in superscript, and the term ``\textsuperscript{*}Equal contribution"
should be placed in the footnote block ahead of the list of affiliations. A
list of corresponding authors and their emails (in the format Full Name
\textless{}email@domain.com\textgreater{}) can follow the list of affiliations.
Ideally only one or two names should be listed.

A sample file with author names is included in the ICML2024 style file
package. Turn on the \texttt{[accepted]} option to the stylefile to
see the names rendered. All of the guidelines above are implemented
by the \LaTeX\ style file.

\subsection{Abstract}

The paper abstract should begin in the left column, 0.4~inches below the final
address. The heading `Abstract' should be centered, bold, and in 11~point type.
The abstract body should use 10~point type, with a vertical spacing of
11~points, and should be indented 0.25~inches more than normal on left-hand and
right-hand margins. Insert 0.4~inches of blank space after the body. Keep your
abstract brief and self-contained, limiting it to one paragraph and roughly 4--6
sentences. Gross violations will require correction at the camera-ready phase.

\subsection{Partitioning the Text}

You should organize your paper into sections and paragraphs to help
readers place a structure on the material and understand its
contributions.

\subsubsection{Sections and Subsections}

Section headings should be numbered, flush left, and set in 11~pt bold
type with the content words capitalized. Leave 0.25~inches of space
before the heading and 0.15~inches after the heading.

Similarly, subsection headings should be numbered, flush left, and set
in 10~pt bold type with the content words capitalized. Leave
0.2~inches of space before the heading and 0.13~inches afterward.

Finally, subsubsection headings should be numbered, flush left, and
set in 10~pt small caps with the content words capitalized. Leave
0.18~inches of space before the heading and 0.1~inches after the
heading.

Please use no more than three levels of headings.

\subsubsection{Paragraphs and Footnotes}

Within each section or subsection, you should further partition the
paper into paragraphs. Do not indent the first line of a given
paragraph, but insert a blank line between succeeding ones.

You can use footnotes\footnote{Footnotes
should be complete sentences.} to provide readers with additional
information about a topic without interrupting the flow of the paper.
Indicate footnotes with a number in the text where the point is most
relevant. Place the footnote in 9~point type at the bottom of the
column in which it appears. Precede the first footnote in a column
with a horizontal rule of 0.8~inches.\footnote{Multiple footnotes can
appear in each column, in the same order as they appear in the text,
but spread them across columns and pages if possible.}

\begin{figure}[ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{icml_numpapers}}
\caption{Historical locations and number of accepted papers for International
Machine Learning Conferences (ICML 1993 -- ICML 2008) and International
Workshops on Machine Learning (ML 1988 -- ML 1992). At the time this figure was
produced, the number of accepted papers for ICML 2008 was unknown and instead
estimated.}
\label{icml-historical}
\end{center}
\vskip -0.2in
\end{figure}

\subsection{Figures}

You may want to include figures in the paper to illustrate
your approach and results. Such artwork should be centered,
legible, and separated from the text. Lines should be dark and at
least 0.5~points thick for purposes of reproduction, and text should
not appear on a gray background.

Label all distinct components of each figure. If the figure takes the
form of a graph, then give a name for each axis and include a legend
that briefly describes each curve. Do not include a title inside the
figure; instead, the caption should serve this function.

Number figures sequentially, placing the figure number and caption
\emph{after} the graphics, with at least 0.1~inches of space before
the caption and 0.1~inches after it, as in
\cref{icml-historical}. The figure caption should be set in
9~point type and centered unless it runs two or more lines, in which
case it should be flush left. You may float figures to the top or
bottom of a column, and you may set wide figures across both columns
(use the environment \texttt{figure*} in \LaTeX). Always place
two-column figures at the top or bottom of the page.

\subsection{Algorithms}

If you are using \LaTeX, please use the ``algorithm'' and ``algorithmic''
environments to format pseudocode. These require
the corresponding stylefiles, algorithm.sty and
algorithmic.sty, which are supplied with this package.
\cref{alg:example} shows an example.

\begin{algorithm}[tb]
   \caption{Bubble Sort}
   \label{alg:example}
\begin{algorithmic}
   \STATE {\bfseries Input:} data $x_i$, size $m$
   \REPEAT
   \STATE Initialize $noChange = true$.
   \FOR{$i=1$ {\bfseries to} $m-1$}
   \IF{$x_i > x_{i+1}$}
   \STATE Swap $x_i$ and $x_{i+1}$
   \STATE $noChange = false$
   \ENDIF
   \ENDFOR
   \UNTIL{$noChange$ is $true$}
\end{algorithmic}
\end{algorithm}

\subsection{Tables}

You may also want to include tables that summarize material. Like
figures, these should be centered, legible, and numbered consecutively.
However, place the title \emph{above} the table with at least
0.1~inches of space before the title and the same after it, as in
\cref{sample-table}. The table title should be set in 9~point
type and centered unless it runs two or more lines, in which case it
should be flush left.

% Note use of \abovespace and \belowspace to get reasonable spacing
% above and below tabular lines.

\begin{table}[t]
\caption{Classification accuracies for naive Bayes and flexible
Bayes on various data sets.}
\label{sample-table}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lcccr}
\toprule
Data set & Naive & Flexible & Better? \\
\midrule
Breast    & 95.9$\pm$ 0.2& 96.7$\pm$ 0.2& $\surd$ \\
Cleveland & 83.3$\pm$ 0.6& 80.0$\pm$ 0.6& $\times$\\
Glass2    & 61.9$\pm$ 1.4& 83.8$\pm$ 0.7& $\surd$ \\
Credit    & 74.8$\pm$ 0.5& 78.3$\pm$ 0.6&         \\
Horse     & 73.3$\pm$ 0.9& 69.7$\pm$ 1.0& $\times$\\
Meta      & 67.1$\pm$ 0.6& 76.5$\pm$ 0.5& $\surd$ \\
Pima      & 75.1$\pm$ 0.6& 73.9$\pm$ 0.5&         \\
Vehicle   & 44.9$\pm$ 0.6& 61.5$\pm$ 0.4& $\surd$ \\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

Tables contain textual material, whereas figures contain graphical material.
Specify the contents of each row and column in the table's topmost
row. Again, you may float tables to a column's top or bottom, and set
wide tables across both columns. Place two-column tables at the
top or bottom of the page.

\subsection{Theorems and such}
The preferred way is to number definitions, propositions, lemmas, etc. consecutively, within sections, as shown below.
\begin{definition}
\label{def:inj}
A function $f:X \to Y$ is injective if for any $x,y\in X$ different, $f(x)\ne f(y)$.
\end{definition}
Using \cref{def:inj} we immediate get the following result:
\begin{proposition}
If $f$ is injective mapping a set $X$ to another set $Y$, 
the cardinality of $Y$ is at least as large as that of $X$
\end{proposition}
\begin{proof} 
Left as an exercise to the reader. 
\end{proof}
\cref{lem:usefullemma} stated next will prove to be useful.
\begin{lemma}
\label{lem:usefullemma}
For any $f:X \to Y$ and $g:Y\to Z$ injective functions, $f \circ g$ is injective.
\end{lemma}
\begin{theorem}
\label{thm:bigtheorem}
If $f:X\to Y$ is bijective, the cardinality of $X$ and $Y$ are the same.
\end{theorem}
An easy corollary of \cref{thm:bigtheorem} is the following:
\begin{corollary}
If $f:X\to Y$ is bijective, 
the cardinality of $X$ is at least as large as that of $Y$.
\end{corollary}
\begin{assumption}
The set $X$ is finite.
\label{ass:xfinite}
\end{assumption}
\begin{remark}
According to some, it is only the finite case (cf. \cref{ass:xfinite}) that is interesting.
\end{remark}
%restatable

\subsection{Citations and References}

Please use APA reference format regardless of your formatter
or word processor. If you rely on the \LaTeX\/ bibliographic
facility, use \texttt{natbib.sty} and \texttt{icml2024.bst}
included in the style-file package to obtain this format.

Citations within the text should include the authors' last names and
year. If the authors' names are included in the sentence, place only
the year in parentheses, for example when referencing Arthur Samuel's
pioneering work \yrcite{Samuel59}. Otherwise place the entire
reference in parentheses with the authors and year separated by a
comma \cite{Samuel59}. List multiple references separated by
semicolons \cite{kearns89,Samuel59,mitchell80}. Use the `et~al.'
construct only for citations with three or more authors or after
listing all authors to a publication in an earlier reference \cite{MachineLearningI}.

Authors should cite their own work in the third person
in the initial version of their paper submitted for blind review.
Please refer to \cref{author info} for detailed instructions on how to
cite your own papers.

Use an unnumbered first-level section heading for the references, and use a
hanging indent style, with the first line of the reference flush against the
left margin and subsequent lines indented by 10 points. The references at the
end of this document give examples for journal articles \cite{Samuel59},
conference publications \cite{langley00}, book chapters \cite{Newell81}, books
\cite{DudaHart2nd}, edited volumes \cite{MachineLearningI}, technical reports
\cite{mitchell80}, and dissertations \cite{kearns89}.

Alphabetize references by the surnames of the first authors, with
single author entries preceding multiple author entries. Order
references for the same authors by year of publication, with the
earliest first. Make sure that each reference includes all relevant
information (e.g., page numbers).

Please put some effort into making references complete, presentable, and
consistent, e.g. use the actual current name of authors.
If using bibtex, please protect capital letters of names and
abbreviations in titles, for example, use \{B\}ayesian or \{L\}ipschitz
in your .bib file.

\section*{Accessibility}
Authors are kindly asked to make their submissions as accessible as possible for everyone including people with disabilities and sensory or neurological differences.
Tips of how to achieve this and what to pay attention to will be provided on the conference website \url{http://icml.cc/}.

\section*{Software and Data}

If a paper is accepted, we strongly encourage the publication of software and data with the
camera-ready version of the paper whenever appropriate. This can be
done by including a URL in the camera-ready copy. However, \textbf{do not}
include URLs that reveal your institution or identity in your
submission for review. Instead, provide an anonymous URL or upload
the material as ``Supplementary Material'' into the OpenReview reviewing
system. Note that reviewers are not required to look at this material
when writing their review.

% Acknowledgements should only appear in the accepted version.
\section*{Acknowledgements}

\textbf{Do not} include acknowledgements in the initial version of
the paper submitted for blind review.

If a paper is accepted, the final camera-ready version can (and
probably should) include acknowledgements. In this case, please
place such acknowledgements in an unnumbered section at the
end of the paper. Typically, this will include thanks to reviewers
who gave useful comments, to colleagues who contributed to the ideas,
and to funding agencies and corporate sponsors that provided financial
support.

% In the unusual situation where you want a paper to appear in the
% references without citing it in the main text, use \nocite
\nocite{langley00}

\bibliography{example_paper}
\bibliographystyle{icml2024}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn
\section{You \emph{can} have an appendix here.}

You can have as much text here as you want. The main body must be at most $8$ pages long.
For the final version, one more page can be added.
If you want, you can use an appendix like this one.

The $\mathtt{\backslash onecolumn}$ command above can be kept in place if you prefer a one-column appendix, or can be removed if you prefer a two-column appendix.  Apart from this possible change, the style (font size, spacing, margins, page numbering, etc.) should be kept the same as the main body.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018, and modified by Alexandre Bouchard in
% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
% Modified again in 2023 and 2024 by Sivan Sabato and Jonathan Scarlett.
% Previous contributors include Dan Roy, Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2023-12-16 paper/main.tex
makefile
Updates from Overleaf

@@
 % - Fill in the outline.
-% - Publish.
+% - Publish on arXiv.
+% - Format as ICML Position Paper (including title change).
 % - Wait for the call from Stockholm.
@@
 \setlength{\parindent}{1.2\baselineskip} % seriously
+\renewenvironment{abstract}{\paragraph{Abstract:}}{}
 \linespread{1.08}
@@
 
-\section*{Is machine learning good or bad for the natural sciences?}
+\section*{\raggedright %Position Paper:
+Is machine learning good or bad for the natural sciences?}
 
@@
 
-\paragraph{Abstract}
+\begin{abstract}
   Machine learning (ML) methods are having a huge impact across all of the sciences.
@@
   Here I identify locations for ML in the natural sciences where the ontology and epistemology (and computational performance) are valuable.
-  I also call out uses for ML that are not a good idea.
-  In particular, there are a lot of reasons to be concerned about the use of complex ML models to emulate physical (or first-principles) simulations, for epistemological and confirmation-bias reasons.
+  I also call out places in current practice where the introduction of ML introduces strong, unwanted statistical biases.
+  For one, when ML models are used to emulate physical (or first-principles) simulations, they bring strong confirmation biases.
+  For another, when flexible regressions are used to make datasets, the elements of those datasets cannot be used in joint analyses without taking on uncontrolled biases.
   On the other hand, there are also places where the introduction of ML can in fact make a project more scientifically conservative:
@@
   These remarks are intended to apply to all of the natural sciences, but most of the specific examples are in the astrophysics domain.
+\end{abstract}
 
@@
 
-It is an understatement to say that deep learning is having a big impact across the sciences.
+HOGG FIX: It is an understatement to say that deep learning is having a big impact across the sciences.
 There are hundreds of papers per year now on deep-learning methods in astrophysics alone.
@@
 
-If we broaden our scope from deep learning to machine learning more generally, we can point to very important contributions of machine learning to astrophysics.
-Kepler discoveries
-PCA and, eg, SDSS redshift system
-
-Astrophysics is an academic discipline, concerned with understanding the physical world, and naturally occurring mechanisms in play in that world.
+HOGG FIX: Astrophysics is an academic discipline, concerned with understanding the physical world, and naturally occurring mechanisms in play in that world.
 We make progress by discovering new kinds of objects and phenomena, and explaining (and, even better, predicting) new kinds of objects and phenomena.
@@
 
-I mentioned the word ``performance'' above.
+HOGG FIX: I mentioned the word ``performance'' above.
 One of the key ideas of machine-learning methods is that they are usually judged in terms of their performance, or their ability to predict or explain new data.
@@
 
-I am both a developer of machine-learning methods, and a skeptic who thinks they should only rarely be used.
-That said, there are reasons that we absolutely must use machine learning in the near future of astrophysics.
-And there are indeed successes of machine learning that might seem invisible.
-I'll try to speak to both of these points below.
+\paragraph{Our contributions:}
+\begin{itemize}
+  \item We deliver a description of the fundamental \emph{ontology} and \emph{epistemology} of machine learning, and contrast these with the ontologies and epistemologies of the natural sciences.
+  \item We elucidate two important and strong statistical biases that are being introduced to the natural sciences by some of the uses of ML in those disciplines.
+One is a confirmation bias that arises when simulations are replaced or augmented by emulators.
+Another is a more standard estimator bias that is (possibly enormously) amplified when elements of datasets produced by ML regressions are used in combination or joint analyses.
+  \item We show that there are many safe places for the use of ML methods in current natural-science practices, and places where (in the contemporary context) the use of ML methods is effectively \emph{required}.
+Many of these places are in the operational parts of scientific projects.
+  \item Further, we demonstrate that there are contexts inside a natural-scientific analysis in which using the most flexible ML method delivers \emph{the most conservative approach} to the scientific problem.
+These contexts are causal contexts, where providing immense flexibility to (say) a confounder model can strengthen conclusions about the causal channels of interest.
+\end{itemize}
 

### 2023-12-16 paper/main.tex
what the FUCK

@@
 % ------
-% - Write a title and abstract.
 % - Write an outline.
 % - Fill in the outline.
-% - Publish on arXiv.
-% - Format as ICML Position Paper (including title change).
+% - Publish.
 % - Wait for the call from Stockholm.
@@
 \setlength{\parindent}{1.2\baselineskip} % seriously
-\renewenvironment{abstract}{\paragraph{Abstract:}}{}
 \linespread{1.08}
@@
 
-\section*{\raggedright %Position Paper:
-Is machine learning good or bad for the natural sciences?}
+\section*{Is machine learning good or bad for the natural sciences?}
 
@@
   These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
-  Here I identify locations for ML in the natural sciences where the ontology and epistemology (and computational performance) are valuable.
-  I also call out places in current practice where the introduction of ML introduces strong, unwanted statistical biases.
-  For one, when ML models are used to emulate physical (or first-principles) simulations, they bring strong confirmation biases.
+  Here we identify locations for ML in the natural sciences where the ontology and epistemology (and computational performance) are valuable.
+  We also show that the introduction of ML introduces strong, unwanted statistical biases.
+  For one, when flexible models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
   For another, when flexible regressions are used to make datasets, the elements of those datasets cannot be used in joint analyses without taking on uncontrolled biases.
-  On the other hand, there are also places where the introduction of ML can in fact make a project more scientifically conservative:
-  When a flexible machine-learning model is used to represent the effects of nuisances, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of the method can sometimes make the results more trustworthy.
+  On the other hand, there are also places where the introduction of ML can in fact make a project more scientifically conservative; when a flexible machine-learning model is used to represent the effects of nuisances, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy can sometimes make the results more trustworthy.
   These remarks are intended to apply to all of the natural sciences, but most of the specific examples are in the astrophysics domain.
@@
 
-HOGG FIX: It is an understatement to say that deep learning is having a big impact across the sciences.
+It is an understatement to say that deep learning is having a big impact across the sciences.
 There are hundreds of papers per year now on deep-learning methods in astrophysics alone.
@@
 
-HOGG FIX: Astrophysics is an academic discipline, concerned with understanding the physical world, and naturally occurring mechanisms in play in that world.
+If we broaden our scope from deep learning to machine learning more generally, we can point to very important contributions of machine learning to astrophysics.
+Kepler discoveries
+PCA and, eg, SDSS redshift system
+
+Astrophysics is an academic discipline, concerned with understanding the physical world, and naturally occurring mechanisms in play in that world.
 We make progress by discovering new kinds of objects and phenomena, and explaining (and, even better, predicting) new kinds of objects and phenomena.
@@
 
-HOGG FIX: I mentioned the word ``performance'' above.
+I mentioned the word ``performance'' above.
 One of the key ideas of machine-learning methods is that they are usually judged in terms of their performance, or their ability to predict or explain new data.
@@
 That hasn't been nearly so true with machine learning, and (I note with respect), it is way easier to incorporate machine learning into your workflow than to incorporate CCDs into your optical path.
+
+I am both a developer of machine-learning methods, and a skeptic who thinks they should only rarely be used.
+That said, there are reasons that we absolutely must use machine learning in the near future of astrophysics.
+And there are indeed successes of machine learning that might seem invisible.
+I'll try to speak to both of these points below.
 
@@
 
-\section{Why do we need machine learning?}
+\section{Why do we need machine learning in the natural sciences?}
 

### 2023-12-16 paper/main.tex
trying to undo the fuckups of github-Overleaf

@@
 % - Fill in the outline.
-% - Publish.
+% - Publish on the arXiv.
+% - format for ICML Position Paper and submit.
 % - Wait for the call from Stockholm.
@@
 \setlength{\parindent}{1.2\baselineskip} % seriously
+\renewenvironment{abstract}{\paragraph{Abstract:}}{}
+\renewcommand{\author}[1]{\medskip\par\noindent\textbf{#1}}
+\newcommand{\affil}[1]{{\footnotesize\par\noindent #1 \par}}
 \linespread{1.08}
 \frenchspacing\raggedbottom\sloppy\sloppypar
-\renewcommand{\author}[1]{\medskip\par\noindent\textbf{#1}}
-\newcommand{\affil}[1]{{\footnotesize\par\noindent #1 \par}}
 
@@
   Machine learning (ML) methods are having a huge impact across all of the sciences.
-  Advances have come from classification, regression, dimensionality reduction, clustering, emulation, and many other things.
   However, ML has a strong ontology---in which only the data exist---and a strong epistemology---in which a model is considered good if it performs well on held-out training data.
   These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
-  Here we identify locations for ML in the natural sciences where the ontology and epistemology (and computational performance) are valuable.
-  We also show that the introduction of ML introduces strong, unwanted statistical biases.
+  Here we identify locations for ML in the natural sciences where the ontology and epistemology are valuable.
+  For example, when a flexible machine-learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can sometimes make the results more trustworthy.
+  We also show that the there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
   For one, when flexible models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
   For another, when flexible regressions are used to make datasets, the elements of those datasets cannot be used in joint analyses without taking on uncontrolled biases.
-  On the other hand, there are also places where the introduction of ML can in fact make a project more scientifically conservative; when a flexible machine-learning model is used to represent the effects of nuisances, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy can sometimes make the results more trustworthy.
   These remarks are intended to apply to all of the natural sciences, but most of the specific examples are in the astrophysics domain.

### 2023-12-16 paper/main.tex
intro snark

@@
 
-It is an understatement to say that deep learning is having a big impact across the sciences.
-There are hundreds of papers per year now on deep-learning methods in astrophysics alone.
+It is an understatement to say that machine learning (ML) is having a big impact across the sciences.
+A significant fraction of all scientific papers in the natural sciences now employ ML in part (or all) of their analyses.
 However, when we ask what scientific breakthroughs have been enabled by this influx of new tools and methods, there isn't a long list.
-The success of the AlphaFold projects in protein structure (\cite{alphafold}) are often raised.
-But these are successes in a biological or medical field, not astrophysics, and in a very specific challenge-problem setting in which performance is valued over understanding.
-Here in astrophysics we care almost exclusively about understanding.
+The success of the AlphaFold projects in protein structure \cite{alphafold} are often raised.
+But these are successes in a very specific challenge-problem setting in which \emph{performance} is valued over \emph{understanding}.
+In the natural sciences we almost exclusively care about understanding, in the long run.
 
-If we broaden our scope from deep learning to machine learning more generally, we can point to very important contributions of machine learning to astrophysics.
-Kepler discoveries
-PCA and, eg, SDSS redshift system
+The natural sciences are concerned with understanding the world, and naturally occurring mechanisms in play in that world.
+We make progress by discovering new kinds of objects and phenomena, and explaining (and, even better, predicting) qualitatively new kinds of objects and phenomena.
+Our most successful investigations are judged in terms of the questions they answer, or the new questions they raise, or both.
+The question here is: How will ML contribute to this mission?
 
-Astrophysics is an academic discipline, concerned with understanding the physical world, and naturally occurring mechanisms in play in that world.
-We make progress by discovering new kinds of objects and phenomena, and explaining (and, even better, predicting) new kinds of objects and phenomena.
-Our most successful investigations are judged in terms of the questions they answer, or the new questions they raise, or both.
-The question here is: How can deep learning contribute to this mission?
+In contrast to natural science, ML research and ML methods are concerned with making accurate predictions for, or descriptions of, \emph{data}.
+A ML method is considered successful if it performs well on held-out training data, even if the latent structure of the model is generic and the internals are impossible to interpret.
+In ML, the considerations are all at the level of the data, and we are happy to use models the parameters of which we don't understand in detail.
+In natural science, on the other hand, the considerations are all at the level of the \emph{latents}:
+We use data to learn about the latent structure of the world or of the system we are studying.
+The things we care about are almost never directly observable; they are latent objects in a physical (or chemical or biological) model that predicts the observables.
 
-I mentioned the word ``performance'' above.
-One of the key ideas of machine-learning methods is that they are usually judged in terms of their performance, or their ability to predict or explain new data.
-That's different, philosophically, from what's traditionally important in astrophysics.
-When the expansion of the Universe was discovered (\cite{expansion, expansion2}), the result was not (primarily) judged on the basis of its ability to predict new data.
-That is, the investigators were not mainly trying to predict the Doppler shifts of held-out galaxies!
-They were trying to measure a property of the Universe, traced by galaxies.
-And their result answered some important questions, and raised new ones.
-Of course the expansion result has lasted in part because it \emph{does} make new predictions for new data, and very accurate predictions!
-So it isn't irrelevant that the performance of the hypothesis is good.
-However, there is some disconnect between how we assess a machine-learning method and how we assess a scientific contribution.
+For a concrete example, when the expansion of the Universe was discovered \cite{expansion, expansion2}, the discovery was important, but not because it permitted us to predict the values of the redshifts of previously unobserved galaxies (though it did indeed permit that).
+The discovery was important because it told us previously unknown things about the age and evolution of the Universe, and it confirmed a prediction of general relativity, which is a theory of the latent structure of space and time.
+The discovery would not have been seen as important if Hubble and Humason had instead announced that they had trained a deep multilayer perceptron that could predict the Doppler shifts of held-out extragalactic nebulae.
 
-HOGG: NEED to make the point, here or below, that I respect the ML point of view.
-It could be helpful for astrophysics if we moved towards it in some ways.
+HOGG: NEED to make the point, here or below, that we respect the ML point of view.
+It could be helpful for some areas in natural science if we moved towards it in some ways.
 Oh and the statistical care is of interest too.
-Maybe this is a section about the underlying philosophies of ML.
+Maybe this should move to the section about the underlying philosophies of ML.
+
+HOGG: Make sure we say: One way to state the question raised in this position paper is:
+\emph{Where, in the natural sciences, can you use a model that you don't understand?}
 
@@
 
+\paragraph{Acknowledgements:}
+Hill, Roweis, Sch\"olkopf, Villar.
+
 \bibliography{ml_in_astro}

### 2023-12-18 paper/main.tex
more philosophy

@@
 % ------------
-% - Uhhh.
+% - Be very careful about the SUBJECT of sentences. Is it ML that is bad? Is it the practice that is bad? Is it the practitioner that is bad?
+% - [more here]
 
@@
 \usepackage[letterpaper]{geometry}
+\usepackage{xcolor}
 
@@
 \setlength{\parindent}{1.2\baselineskip} % seriously
-\renewenvironment{abstract}{\paragraph{Abstract:}}{}
+\renewcommand{\title}[1]{\section*{\raggedright #1}}
 \renewcommand{\author}[1]{\medskip\par\noindent\textbf{#1}}
 \newcommand{\affil}[1]{{\footnotesize\par\noindent #1 \par}}
+\renewenvironment{abstract}{\paragraph{Abstract:}}{}
+\newcommand{\documentname}{\textsl{Position Paper}}
+\pagestyle{myheadings}
+\markboth{foo}{\textcolor{gray}{\textsf{Hogg / Is machine learning good or bad for the natural sciences?}}}
 \linespread{1.08}
@@
 
-\begin{document}
+\begin{document}\thispagestyle{empty}
 
-\section*{Is machine learning good or bad for the natural sciences?}
+\title{\documentname: Is machine learning good or bad for the natural sciences?}
 
@@
 A ML method is considered successful if it performs well on held-out training data, even if the latent structure of the model is generic and the internals are impossible to interpret.
-In ML, the considerations are all at the level of the data, and we are happy to use models the parameters of which we don't understand in detail.
-In natural science, on the other hand, the considerations are all at the level of the \emph{latents}:
+In ML, the considerations are all at the level of the data, and we are happy to use models where we have litlle or no understanding of the meanings or values of the latent parameters or weights.
+Another way to ask our question is:
+\emph{Where, in the natural sciences, can you use a model that you don't understand?}
+
+In natural science the most important contributions and results are all at the level of the \emph{latents}:
 We use data to learn about the latent structure of the world or of the system we are studying.
-The things we care about are almost never directly observable; they are latent objects in a physical (or chemical or biological) model that predicts the observables.
+In astrophysics, this could be the interior structure of the Sun, or the statistics of planets around other stars, or the map of the dark matter surrounding the Milky Way.
+The things we care about are almost never directly observable; they are parameters (or hyper-parameters) of a physical (or chemical or biological) model that predicts the observables.
 
-For a concrete example, when the expansion of the Universe was discovered \cite{expansion, expansion2}, the discovery was important, but not because it permitted us to predict the values of the redshifts of previously unobserved galaxies (though it did indeed permit that).
+For a concrete example, when the expansion of the Universe was discovered \cite{expansion, expansion2}, the discovery was important, but not because it permitted us to predict the values of the redshifts of new galaxies (though it did indeed permit that).
 The discovery was important because it told us previously unknown things about the age and evolution of the Universe, and it confirmed a prediction of general relativity, which is a theory of the latent structure of space and time.
@@
 
-HOGG: NEED to make the point, here or below, that we respect the ML point of view.
-It could be helpful for some areas in natural science if we moved towards it in some ways.
-Oh and the statistical care is of interest too.
-Maybe this should move to the section about the underlying philosophies of ML.
-
-HOGG: Make sure we say: One way to state the question raised in this position paper is:
-\emph{Where, in the natural sciences, can you use a model that you don't understand?}
-
-I don't want to sound anti-technological however.
-Machine learning is a new and productive technology.
-We should use it!
-Astrophysics has moved forward through technological progress, starting with Galileo (\cite{galileo}).
-In my own lifetime, the most remarkable technological advance was (probably) the change from photographic plates to charge-coupled devices (CCDs; see, for an early review, \cite{ccd}).
-In just a few years, almost every telescope and spectrograph in the world switched from a photographic focal plane to a CCD focal plane, and immediately there were new discoveries.
-That's a case in which new technology immediately translated into new discoveries.
-That hasn't been nearly so true with machine learning, and (I note with respect), it is way easier to incorporate machine learning into your workflow than to incorporate CCDs into your optical path.
-
-I am both a developer of machine-learning methods, and a skeptic who thinks they should only rarely be used.
-That said, there are reasons that we absolutely must use machine learning in the near future of astrophysics.
-And there are indeed successes of machine learning that might seem invisible.
-I'll try to speak to both of these points below.
+We are both ML \emph{skeptics} and ML \emph{practitioners}.
+We are writing this for an audience that is either developing an ML method for science applications or else bringing ML into a scientific domain.
+The main points of this \documentname{} are \textsl{(1)}~that ML has places where it is very valuable in the contemporary practice of science, and \textsl{(2)}~that it also has places where it will create problems for science.
+The answer to the question in the title is ``Both.''
 
@@
 
-\section{What is machine learning?}\label{sec:what}
+\section{The ontology and epistemology of machine learning}
 
-There is a vague definition of machine learning in terms of data and capacity:
-A method is a machine-learning method if its capability increases (hopefully dramatically) as the machine sees more data (HOGG GET THIS MORE PRECISE AND CITE).
-In some sense this is true of any measurement process:
-The measurement improves as more data come in!
-So if we are going to be more specific, we'd like the model capacity or capability to improve faster (in some sense) than something like the square root of the increment in data.
+\paragraph{What is machine learning?}
+For the purposes of this \documentname, we will employ an expansive or inclusive definition of ML.
+For us, a method is ML if its \emph{capability increases substantially as it sees more data} \cite{ml_definition}.
+In some sense this definition could be seen as true of any measurement process, since measurements improve as the data improve. 
+If we are going to be more specific, we'd like the model precision or capability to improve faster (in some sense) than (something like) the square root of the increment in data.
 
-There is a more specific definition of machine learning in terms of specific methods:
-Whenever a neural network or a Gaussian process is in play, machine learning is in play.
-But when the former definition is too broad, the latter is almost always too narrow.
-For example, principal components analysis (PCA) is a workhorse in astronomy since way back (for example, \cite{pcaredshift}), and PCA is an old, and simple, but definitely machine-learning, method.
+This definition is broad:
+In addition to methods like convolutional neural networks \cite{cnn}, multi-layer perceptrons \cite{mlp}, and transformers \cite{transformer}, it includes large linear regressions \cite{linearregression}, gaussian processes \cite{gp}, support-vector machines \cite{svm}, principal components analysis \cite{pca}, kernel density estimates \cite{kde}, and even some kinds of multilayer or hierarchical models \cite{multilevel}.
+Feel free to personally take any more restrictive definition.
+Our comments will apply to everything in the larger class.
 
-Supervised and unsupervised.
+\paragraph{Ontology:}
+Unsupervised ML methods deliver descriptions or compressions of data.
+Supervised ML methods find relationships between data (features) and data (labels).
+In both cases, the methods are judged on their capability to accurately describe the data.
+They are not judged on the details of their latent structure.
+In a very important sense, the ML ontology is that \emph{only the data exist}.
 
-Classification, regression.
+In support of this point of view, here are some comments:
+Deep-learning models have enormously large internal degeneracies (combinatorial degeneracies even).
+These degeneracies are not seen as a problem, since all they do is make the latent weights less interpretable.
+Contemporary optimization schemes are stochastic \cite{stochastic} and most models are non-convex.
+The fact that it is impossible to find the global optimum---and the fact that in most contemporary models that isn't even the goal \cite{not_global}---shows that the latent parameters are not important.
+The use of early stopping \cite{early_stop} and drop-outs \cite{dropout} to stabilize optimization strengthens this point.
+Any practitioner using a 40-layer network can trivially (and without analysis) switch to a 42-layer, despite the fact that (from a functional point of view), this substantially changes the model capacity.
 
-Dimensionality reduction, clustering, density estimation.
+Models are deemed stably optimized and useful not if the latent space is stable, but rather if the predictions for held-out data are stable.
+Thus it is the data that exist in ML contexts, not the latent parameters.
 
-Something about how we will use the word ``model'' in what follows.
+\paragraph{Epistemology:}
+A trained ML model is deemed successful or correct if \emph{it performs well on held-out training data}.
+This epistemological position is related strongly to the ontological position:
+If only the data exist, then the success of a model is judged only in terms of the data it describes.
+
+In support of this position we could point to the literature on adversarial attacks \cite{adversarial1, adversarial2}.
+This literature shows us, in dramatic ways, that ML methods are not doing what we (na\"ively) believe that humans or scientists or scientific theories do in comparable circumstances.
+And yet, these attacks do not suggest (to most practitioners of ML) that the methods are wrong or require revision.
+Even the responses to attacks have responded in ways that involve augmentation of data \cite{adversarial_training}, such that the vulnerability to attack is lessened without compromising the fundamental epistemological point that performance on data is the primary standard of truth.
+
+A critic of our epistemological claim could point to the large literature on out-of-sample generalization, transfer learning \cite{transfer}, or the more grand contemporary idea of foundation models \cite{foundation}.
+But even in these contexts, models are deemed successful when they explain new or held-out data; the latent structure of the models is not a primary consideration when the validity of the model is in question.
+
+\paragraph{Natural sciences:}
+Ontologies and epistemologies are somewhat different in different scientific fields---and among different practitioners---but in almost no natural-scientific context do the ontology and epistemology match those of ML.
+
+In physics, for example, the ontology is much broader than it is in ML.
+Not only do the data exist, but so do forces, energies, momenta, charges, spacetime, wave functions, virtual particles, and much more.
+These entities are judged to exist in part because they are involved in the latent structure of the successful theories; almost none of them are direct observables.
+
+In physics the epistemology is much more restrictive than it is in ML.
+A physics model---which, as we note, is almost always a model of latent structure---is judged to be good or strongly confirmed if it explains some aspects of different data in multiple domains, and if it connects in natural ways to other theories or principles (such as conservation laws and invariances) that are strongly confirmed themselves.
+It is worthy of noting here that a successful theory is usually a combination of a simple theory of the latents, and a less simple observation model, which explains how the latents appear in the data space.
+
+Differences aside, there is an argument to make that maybe the natural sciences could be better off if they moved at least slightly closer to the ML points of view on ontology and epistemology.
+In the long run, the data are more stable than the fundamental theories of physics:
+Newtonian gravity was replaced with general relativity, and we expect general relativity to be replaced by some kind of quantum gravitational theory.
+At each change, the ontology changes, while the data are stable.
+We are not recommending this, but noting that a robust theory of physics might be more positively oriented towards the data than standard practice.
 
@@
 
+Outlier detection.
+
 Speeding up simulations?
@@
 
-\section{The ontology and epistemology of machine learning}
+Engineering point: Science requires execution of projects. Execution requires engineering. ML is often great for helping with that engineering.
 
-HOGG: Definitions of latent variables and so on; what are latent variables and why are they important?
+\section{When is machine learning bad for natural science?}
 
-Contemporary machine-learning models---and especially deep-learning models---have explicit unbroken, combinatorically large degeneracies.
-That is, the weights and nodes in various layers of the network have permutation symmetries such that a very large space of permutation operators can be applied to the latent space without making any changes whatsoever to the input--output relationships of the model.
-That shows that there is no belief in the meaning of any internal latent variables.
+Multiplication of priors argument. Confirmation bias argument.
 
-Furthermore, stochastic gradient, early stopping and so on:
-There isn't even a belief that the latents should be the actual argmin of anything.
-The latent values are of absolutely no importance or meaning whatsoever.
+\section{When is machine learning good for natural science?}
 
-If there is any stability to a model, it is the stability of its predictions, or its behavior on held-out data.
-
-Also maybe mention here the statistical philosophy or care.
-
-\section{When is machine learning a bad idea?}
-
-Adversarial attacks argument. Confirmation bias argument. Planet search argument.
-
-\section{When is machine learning the conservative choice?}
-
-Come back to the planet-search example.
+Foreground, background, and confounder models.
 
@@
 \paragraph{Acknowledgements:}
-Hill, Roweis, Sch\"olkopf, Villar.
+It is a pleasure to thank
+  Gaby Contardo (SISSA),
+  Jennifer Hill (NYU),
+  Adrian Price-Whelan (Flatiron),
+  Sam Roweis (deceased),
+  Bernhard Sch\"olkopf (MPI-IS), and
+  Soledad Villar (JHU)
+for conversations over the years related to these arguments.
 
+\raggedright
 \bibliography{ml_in_astro}

### 2023-12-18 paper/main.tex
philosophy and biases

@@
 \setlength{\oddsidemargin}{0.00in}
-\setlength{\textheight}{9.00in}
+\setlength{\textheight}{9.25in}
 \setlength{\topmargin}{-0.50in}
@@
 \section{Introduction}\label{sec:intro}
-
 It is an understatement to say that machine learning (ML) is having a big impact across the sciences.
@@
 
-\section{The ontology and epistemology of machine learning}
+\section{The ontology and epistemology of machine learning}\label{sec:philosophy}
 
@@
 
+HOGG: DROP THIS?
 Differences aside, there is an argument to make that maybe the natural sciences could be better off if they moved at least slightly closer to the ML points of view on ontology and epistemology.
@@
 At each change, the ontology changes, while the data are stable.
-We are not recommending this, but noting that a robust theory of physics might be more positively oriented towards the data than standard practice.
+We are not recommending this, but noting that a robust philosophy of science might be more positively oriented towards the data than standard practice.
 
@@
 
-\section{When is machine learning bad for natural science?}
+\section{When is machine learning bad for natural science?}\label{sec:bad}
+In this section we elucidate two statistical biases that can be introduced when ML methods are introduced into a natural-science project.
+Neither of these biases can be easily corrected or removed.
 
-Multiplication of priors argument. Confirmation bias argument.
+\paragraph{Multiplications of uncontrolled priors:}
+The outcomes of ML regressions are label estimates that are conditioned on the input features \emph{and also} on the totality of the training set used to train the regression.
+That conditioning on the training set can lead to strong biases when outputs of regressions are used jointly or in combination to measure a population or sub-population property.
 
-\section{When is machine learning good for natural science?}
+Imagine that you have taken data on millions of stars, and you know the precise ages of only a small fraction of them.
+You can train a ML regression on those for which you know the ages, and label all the other stars with the ages delivered by the trained regression model.
+This will leave you with a million stellar ages, which span lots of ages.
+These ages are not precise, because the age information in the data is small; any age estimates will be very noisy.
+Nonetheless, you train this regression, validate it will held-out training data, and produce a publicly available catalog of stellar ages.
+
+Now any investigator can download the catalog, take subsets of stars and ask ``What is the mean age of this subset?''
+For example, they can ask for the mean age for stars at Galactocentric radii $>10$\,kpc, which they expect to be on the younger side (though well within the distribution of ages in your training set).
+The answer they get for this mean age will be \emph{very strongly biased}.
+
+Why? When the information in the data is low (as it is by construction in this example), the prior information coming from the training set is comparatively more important.
+And when you combine many measurements into the mean age, you will generally be compounding (rather than cancelling out) the bias.
+One way to see it is in this toy example: HOGG TOY WITH GAUSSIANS.
+
+HOGG: Summarize the source of the bias. Counter-act this bias? Use LF-only methods if such can be constructed? Train a joint-star model instead!?
+
+\paragraph{Emulator-induced confirmation bias:}
+In many fields in the natural sciences, the fundamental theoretical model is \emph{computational}, meaning that theoretical predictions are made with large computer simulations.
+These simulations are usually limited in the range of length scales and time scales which they can simultaneously compute.
+Comparison of data with simulations usually involves running enormous numbers of simulations \cite{sbi}, because simulations have to span ranges of fundamental parameters and initial conditions and so on.
+Thus these simulation requirements are getting very large for contemporary research projects.
+For example, in cosmology contemporary experiments could easily consume the total computing capacity of the United States.
+Hence we turn to emulators---trained ML regressions that have learned the input-output relationship of a simulation, or the relationship between an inexpensive low-resolution simulation and an expensive high-resolution experiment.
+Emulators can make computationally impossible projects possible.
+
+Recall that because these emulators are ML methods, they generally have uninterpretable internal parameters and weights, and they generally have been validated by comparison with held-out training data.
+Like all high-capacity ML methods, they generally are vulnerable to adversarial attacks.
+
+Now imagine that two cosmology experiments proceed to do enormous emulator-based inferences on large data sets.
+Imagine that these inferences are so large that it would have been exceedingly expensive to have done them with the original first-principles simulations instead of the emulators.
+Experiment A finds cosmological parameters very much in line with our expectations going in.
+Experiment B finds something extremely surprising about the mass matrix for the neutrinos, in conflict with other measurements and our expectations.
+The comment is made, when the results of Experiment B come out, that perhaps the anomaly comes from something very slightly wrong or off in the trained ML emulator.
+
+By construction, both experiments are extremely expensive to re-analyze with first-principles simulations.
+Which one would we fund for reproduction?
+The motivation to re-analyze Experiment B is far, far higher than the motivation to re-analyze Experiment A.
+That is a classic example of \emph{confirmation bias}.
+
+Since the money involved is astronomical, there is no simple fix for this unless we plan to re-analyze all experiments using first-principles simulations, no matter what they find.
+If that is what's required to avoid the bias, then we should just analyze everything with simulations in the first place, and never use the emulators at any stage.
+
+\section{When is machine learning good for natural science?}\label{sec:good}
 

### 2023-12-20 paper/main.tex
start appendix, other changes

@@
 \newcommand{\documentname}{\textsl{Position Paper}}
+\newcommand{\sectionname}{Section}
 \pagestyle{myheadings}
@@
 A physics model---which, as we note, is almost always a model of latent structure---is judged to be good or strongly confirmed if it explains some aspects of different data in multiple domains, and if it connects in natural ways to other theories or principles (such as conservation laws and invariances) that are strongly confirmed themselves.
-It is worthy of noting here that a successful theory is usually a combination of a simple theory of the latents, and a less simple observation model, which explains how the latents appear in the data space.
+It is worthy of note here that a successful theory is usually a combination of a simple theory of the latents, and a less simple observation model, which explains how the latents appear in the data space.
 
@@
 \section{Why do we need machine learning in the natural sciences?}
-
-Label transfer.
-
-Speeding up decisions.
-
-Modeling nuisances.
-
-Classification?
-
-Outlier detection.
-
-Speeding up simulations?
-
-The short summary is that we need machine learning; we can't live without it.
-
-Engineering point: Science requires execution of projects. Execution requires engineering. ML is often great for helping with that engineering.
+In this \sectionname{} we give some idea of how and why ML has had such a big impact in the natural sciences, and how and why many scientific projects require ML technologies.
+In most cases, the ML is required on the engineering or execution side of projects that are large-scale in some sense.
+\begin{description}
+  \item[Label transfer:] Sometimes a project has informative data on a very large number of objects, but precise labels for only a few, maybe obtained through very careful analysis or external data.
+  In this case, if it is extremely expensive to label more objects, a regression can be trained on the labeled data points and then the trained regression used to label
+  \item[Speeding up decisions:]
+  \item[Modeling nuisances:]
+  \item[Classification:]
+  \item[Outlier detection:]
+  \item[Speeding up simulations:]
+\end{description}
+The short summary is that we need ML in the natural sciences; we can't live without it.
+The question here is how we will use it safely?
 
 \section{When is machine learning bad for natural science?}\label{sec:bad}
-In this section we elucidate two statistical biases that can be introduced when ML methods are introduced into a natural-science project.
+In this \sectionname{} we elucidate two statistical biases that can be introduced when ML methods are introduced into a natural-science project.
 Neither of these biases can be easily corrected or removed.
@@
 
-\raggedright
+{\raggedright
 \bibliography{ml_in_astro}
 \bibliographystyle{plain}
+}
+
+\clearpage\appendix
+\section{\raggedright Population-level biases in regression outputs}
+A machine-learning (ML) regression learns the best set of parameters or weights $\hat{\theta}$ of a flexible function $f(x;\theta)$ from a training set of labeled data.
+The learned function $f(x;\hat{\theta})$ delivers a predicted or estimated label $\hat{y}$ for any data point with a complete set of features $x$.
+Thus an estimated label $\hat{y}_\ast$ for a data point with features $x_\ast$ has taken information \emph{both} from the features $x_\ast$ and \emph{also} from the features and labels of all the data points in the training set used to set the weights $\hat{\theta}$.
+In applications in which regression-estimated labels are used, it matters \emph{how much information} is coming from the individual data-point's features and how much from the original training set.
+
+To illustrate these ideas, we construct a toy data analysis.
+This toy is very general, but it has a clearly astronomical feel to it.
+
+Hello world
 

### 2023-12-20 paper/main.tex
messing around

@@
 \clearpage\appendix
-\section{\raggedright Population-level biases in regression outputs}
+\section{Population-level biases in regression outputs}
 A machine-learning (ML) regression learns the best set of parameters or weights $\hat{\theta}$ of a flexible function $f(x;\theta)$ from a training set of labeled data.

### 2023-12-21 paper/main.tex
Updates from Overleaf
toy data creation, figure and figure caption

@@
 \usepackage{xcolor}
+\usepackage{graphicx}
 
 % Hogg typesetting issues
-\setlength{\textwidth}{6.50in}
+\setlength{\textwidth}{6.00in}
+\setlength{\oddsidemargin}{0.25in}
 \setlength{\oddsidemargin}{0.00in}
@@
 
-Foreground, background, and confounder models.
+\paragraph{Foreground, background, and confounder models:}
+
+\paragraph{Outlier detection:}
 
@@
 \section{Population-level biases in regression outputs}
-A machine-learning (ML) regression learns the best set of parameters or weights $\hat{\theta}$ of a flexible function $f(x;\theta)$ from a training set of labeled data.
+A machine-learning (ML) regression learns the best set of parameters (weights and thresholds, for example) $\hat{\theta}$ of a flexible function $f(x;\theta)$ from a training set of labeled data.
 The learned function $f(x;\hat{\theta})$ delivers a predicted or estimated label $\hat{y}$ for any data point with a complete set of features $x$.
-Thus an estimated label $\hat{y}_\ast$ for a data point with features $x_\ast$ has taken information \emph{both} from the features $x_\ast$ and \emph{also} from the features and labels of all the data points in the training set used to set the weights $\hat{\theta}$.
+Thus an estimated label $\hat{y}_\ast$ for a data point with features $x_\ast$ has taken information \emph{both} from the features $x_\ast$ and \emph{also} from the features and labels of all the data points in the training set used to set the parameters $\hat{\theta}$.
 In applications in which regression-estimated labels are used, it matters \emph{how much information} is coming from the individual data-point's features and how much from the original training set.
@@
 To illustrate these ideas, we construct a toy data analysis.
+The complete toy data set is generated with a process with the following elements:
+\begin{enumerate}\setlength{\itemsep}{0in}
+    \item A floating-point value of a known parameter called ``guiding radius'' $r$ is generated in the range $0<r<14$.
+    \item A floating-point value of a latent parameter called ``age'' $\eta$ is generated from a Gaussian with unit variance [HOGG CHECK] and mean $13 - (13/14)\,r$
+    \item The point is discarded if the age is outside the range $0<\eta<13$. If the point is not discarded:
+    \item A $K$-dimensional latent vector $\zeta$ with $K=14$ [HOGG CHECK] is generated from a zero-mean, unit-variance Gaussian.
+    \item A $M$-dimensional latent vector $\xi$ with $M=110$ is created by $\xi = \xi_\eta\,\eta + \xi_\zeta\cdot\zeta$, where $\xi_\eta$ is a $M$-vector and $\xi_\zeta$ is a $M\times K$ matrix.
+The elements of $\xi_\eta$ and $\xi_\zeta$ were drawn (before the start) from a Gaussian with zero mean and variance $0.00046$ and $0.071$ respectively [HOGG CHECK].
+    \item A label $y$ called ``measured age'' is generated from a Gaussian with unit variance [HOGG CHECK] and mean $\eta$.
+    \item A $M$-dimensional feature vector $x$ called ``the data'' is generated from a Gaussian with variance $0.0025$ and a mean set by taking $1 + \xi$ and rectifying all pixels to lie in the range between zero and unity.
+\end{enumerate}
+The data generated this way are shown in \figurename~\ref{fig:regression}.
 This toy is very general, but it has a clearly astronomical feel to it.
+The data contain a non-trivial relationship between measured age $y$ and guiding radius $r$.
+This relationship is shown in \figurename~\ref{fig:regression}.
+It isn't precisely linear because of the censoring of data points with latent ages outside the range $0<\eta<13$.
 
 Hello world
+\begin{figure}[p!]
+\includegraphics[width=0.49\textwidth]{notebooks/data_examples.png}
+\includegraphics[width=0.49\textwidth]{notebooks/training_data.png} \\
+\includegraphics[width=0.49\textwidth]{notebooks/validation.png}
+\includegraphics[width=0.49\textwidth]{notebooks/test_data_results.png}
+\caption{\sffamily%
+Visualization of the toy regression.
+\textsl{Top-left:} Random examples of the data vectors $x$, which are one-dimensional images generated from a linear model plus a nonlinearity created by two rectifications.
+Details of the data generation are given in the text.
+Each example $x$ is labeled on the right side by the value of its label $y$.
+\textsl{Top-right:} The training-set labels $y$, plotted against the known parameter $r$, which is not used in the regression (only the vectors $x$ are used).
+Also shown are solid red circles showing the true mean relationship between $y$ and $r$ in the toy data.
+Open black squares show the empirical mean relationship measured in bins in the training-set data.
+\textsl{Bottom-left:} Validation of the trained regression in the validation set, showing that the label estimates $\hat{y}$ are noisy (as expected given the problem set-up) but not strongly biased.
+\textsl{Bottom-right:} The regression estimates $\hat{y}$ in the very large test set, plotted against the known parameter $r$.
+Also shown are the same solid red circles and open black squares as in the top-right plot.
+Black exes show the mean relationship between $\hat{y}$ and $r$.
+The relationship shown by the exes is very precise but biased far away from the true relationship, unlike the relationship shown by the open squares (measured in the training set alone).\label{fig:regression}}
+\end{figure}
 

### 2023-12-21 paper/main.tex
moar experiment and result description

@@
 \end{enumerate}
-The data generated this way are shown in \figurename~\ref{fig:regression}.
-This toy is very general, but it has a clearly astronomical feel to it.
-The data contain a non-trivial relationship between measured age $y$ and guiding radius $r$.
+The data generated this way are shown in the top panels of \figurename~\ref{fig:regression}.
+This toy will make a very general point, but it has a decidedly astronomical feel to it.
+The data contain a non-trivial relationship between label $y$ (measured age) $y$ and known parameter $r$ (guiding radius).
 This relationship is shown in \figurename~\ref{fig:regression}.
@@
 
-Hello world
+We construct a three-layer multilayer perceptron (MLP) (using the scikit-learn implementation) with layer sizes of 64, 32, and 16 neurons.
+The model is trained on a training set of 4096 data points with features $x$ and labels $y$.
+The trained model takes as input features $x$ and outputs estimated labels $\hat{y}$.
+The model is validated on a validation set of 2048 data points with features $x$ and labels $y$.
+The test set is a much larger set of 105,864 data points with features $x$ and no labels.
+The model is used to make estimated labels $\hat{y}$ in the full test set.
+The validation and test labels are shown in the bottom panels of \figurename~\ref{fig:regression}.
+Importantly, \emph{the training, validation, and test samples are drawn from exactly the same process}, with the same distribution in all properties.
 \begin{figure}[p!]
@@
 
+The regression-estimated labels $\hat{y}$ are related to the measured labels $y$ in the validation set with a linear relationship with unit slope and zero intercept.
+They are not precisely estimated, but there is no evidence for bias at the individual-object level.
+
+The true relationship between label $y$ (measured age) and known parameter $r$ (guiding radius) is estimated in a large generated data set by taking means of $y$ in bins of $r$.
+This relationship is shown in \figurename~\ref{fig:regression} with solid red circles.
+The empirical relationship between label $y$ and known parameter $r$ in the training-set data is shown in \figurename~\ref{fig:regression} with open black squares.
+These two relationships are statistically consistent.
+The empirical relationship in the test-set data between the estimated label $\hat{y}$ and the known parameter $r$ in the test-set data is shown in \figurename~\ref{fig:regression} with black exes.
+The test-set relationship is strongly biased, deviating from the true relationship at the edges of the $r$ range.
+Given the large size of the data set, precision is high: This bias is substantially larger than 20-sigma in the worst data points.
+
 \end{document}

### 2023-12-22 paper/main.tex
text-appendix interactions

@@
   \item Further, we demonstrate that there are contexts inside a natural-scientific analysis in which using the most flexible ML method delivers \emph{the most conservative approach} to the scientific problem.
-These contexts are causal contexts, where providing immense flexibility to (say) a confounder model can strengthen conclusions about the causal channels of interest.
+An example is in causal contexts, where providing immense flexibility to (say) a confounder model can strengthen conclusions about the causal channels of interest.
 \end{itemize}
@@
 
-\paragraph{Multiplications of uncontrolled priors:}
+\paragraph{Amplifications of uncontrolled biases:}
 The outcomes of ML regressions are label estimates that are conditioned on the input features \emph{and also} on the totality of the training set used to train the regression.
-That conditioning on the training set can lead to strong biases when outputs of regressions are used jointly or in combination to measure a population or sub-population property.
+This is good; the individual-data-point label estimates from the regression are the lowest ``risk'' (in the statistical sense) when they use non-zero bias in the bias--variance trade-off.
+However, the biases that are weak or manageable on an individual data-point basis become strong or unacceptable biases when outputs of regressions are used jointly or in combination to measure a population or sub-population property.
 
-Imagine that you have taken data on millions of stars, and you know the precise ages of only a small fraction of them.
-You can train a ML regression on those for which you know the ages, and label all the other stars with the ages delivered by the trained regression model.
-This will leave you with a million stellar ages, which span lots of ages.
-These ages are not precise, because the age information in the data is small; any age estimates will be very noisy.
-Nonetheless, you train this regression, validate it will held-out training data, and produce a publicly available catalog of stellar ages.
+This problem is demonstrated with a toy example in \appendixname~\ref{app:toy}.
+It is worst when the regression is used to label a large data set or catalog or sample of data,
+and when elements of that data set are used in populations or joint analyses.
+In general, when multiple data point estimates are used jointly, the variances of the estimators average out but the biases remain.
+This is a straightforward point of statistics, but it isn't currently informing most of the practice of ML in the natural sciences.
+For example, HOGG CITES.
 
-Now any investigator can download the catalog, take subsets of stars and ask ``What is the mean age of this subset?''
-For example, they can ask for the mean age for stars at Galactocentric radii $>10$\,kpc, which they expect to be on the younger side (though well within the distribution of ages in your training set).
-The answer they get for this mean age will be \emph{very strongly biased}.
+In Bayesian language, this problem is closely related to the point that when data are to be combined, they should be combined at the likelihood level, not the posterior level.
+Likelihoods are not affected by population-level biases, and potentially produce unbiased estimates.
+Likelihood-based estimators are not usually the lowest-risk estimators, but they can be low- or zero-bias estimates.
+The outputs of regressions are more like posterior-based estimators, affected by the implicit prior set by the training set.
+When they are combined, the implicit prior gets amplified.
 
-Why? When the information in the data is low (as it is by construction in this example), the prior information coming from the training set is comparatively more important.
-And when you combine many measurements into the mean age, you will generally be compounding (rather than cancelling out) the bias.
-One way to see it is in this toy example: HOGG TOY WITH GAUSSIANS.
+In principle, there are fixes for this problem that involve de-biasing the estimates.
+This is not possible in general, because there is not usually enough validation data to accurately assess the bias.
 
-HOGG: Summarize the source of the bias. Counter-act this bias? Use LF-only methods if such can be constructed? Train a joint-star model instead!?
+Finally we note---and demonstrate in \appendixname~\ref{app:toy}---that none of these problems arise from out-of-sample generalization or distribution-shift problems.
+Distribution shifts just tend to make these kinds of problems worse.
 
@@
 \clearpage\appendix
-\section{Population-level biases in regression outputs}
+\section{Population-level biases in regression outputs}\label{app:toy}
 A machine-learning (ML) regression learns the best set of parameters (weights and thresholds, for example) $\hat{\theta}$ of a flexible function $f(x;\theta)$ from a training set of labeled data.
@@
 The complete toy data set is generated with a process with the following elements:
+HOGG REFORMAT THIS AS AN ALGORITHM WITH THE algorithm PACKAGE.
 \begin{enumerate}\setlength{\itemsep}{0in}
@@
 The test-set relationship is strongly biased, deviating from the true relationship at the edges of the $r$ range.
-Given the large size of the data set, precision is high: This bias is substantially larger than 20-sigma in the worst data points.
+Given the large size of the test set, the precision of the result is high: This bias in the estimated relationship is substantially larger than 20-sigma in the worst data points.
+
+HOGG: What to do? Just use the training set!
+
+HOGG: Note that none of these problems come from bad data, out-of-distribution, or complexity.
 

### 2023-12-22 paper/main.tex

@@
 \newcommand{\sectionname}{Section}
+\newenvironment{hoggnumerate}
+  {\begin{list}
+    {\arabic{enumii}.}
+    {\usecounter{enumii}
+     %\setlength{\labelwidth}{3em}
+     %\setlength{\labelsep}{0em}
+     \setlength{\itemsep}{0in}
+     \setlength{\parsep}{0in}
+     \setlength{\parskip}{0in}
+     %\setlength{\leftmargin}{1.5cm}
+     %\setlength{\rightmargin}{2cm}
+     %\setlength{\itemindent}{0em} 
+     %\let\makelabel=\makeboxlabel
+    }
+  }
+{\end{list}}
 \pagestyle{myheadings}
@@
   In this case, if it is extremely expensive to label more objects, a regression can be trained on the labeled data points and then the trained regression used to label
+  \item[Classification:]
   \item[Speeding up decisions:]
+  \item[Speeding up simulations:]
   \item[Modeling nuisances:]
-  \item[Classification:]
   \item[Outlier detection:]
-  \item[Speeding up simulations:]
+  \item[Making discoveries?]
 \end{description}
@@
 
-\paragraph{Amplifications of uncontrolled biases:}
+\paragraph{Amplifications of training-set biases:}
 The outcomes of ML regressions are label estimates that are conditioned on the input features \emph{and also} on the totality of the training set used to train the regression.
@@
 and when elements of that data set are used in populations or joint analyses.
-In general, when multiple data point estimates are used jointly, the variances of the estimators average out but the biases remain.
-This is a straightforward point of statistics, but it isn't currently informing most of the practice of ML in the natural sciences.
+In general, when multiple data point estimates are used jointly, the variance-induced offsets of the estimators average out but the bias-induced offsets remain fixed.
+This is a straightforward point of statistics---this is not news---but it isn't currently informing most of the practice of ML in the natural sciences.
 For example, HOGG CITES.
+
+In principle, there are fixes for this problem that involve de-biasing the estimates.
+This is not possible in general, because there is not usually enough validation data to accurately assess the bias.
 
@@
 When they are combined, the implicit prior gets amplified.
-
-In principle, there are fixes for this problem that involve de-biasing the estimates.
-This is not possible in general, because there is not usually enough validation data to accurately assess the bias.
+Most ML regressions are not capable of generating or emulating likelihoods.
+There are interesting approaches in development to replace regressions with generative models that can produce likelihoods HOGG CITE THINGS.
 
@@
 To illustrate these ideas, we construct a toy data analysis.
-The complete toy data set is generated with a process with the following elements:
-HOGG REFORMAT THIS AS AN ALGORITHM WITH THE algorithm PACKAGE.
-\begin{enumerate}\setlength{\itemsep}{0in}
+This toy will make a general point, but it has a decidedly astronomical feel to it.
+The complete toy data set is generated by a process with the algorithm below.
+This algorithm makes references to objects $\xi_\eta$ and $\xi_\zeta$ which are generated prior to the start.
+\begin{hoggnumerate}
     \item A floating-point value of a known parameter called ``guiding radius'' $r$ is generated in the range $0<r<14$.
-    \item A floating-point value of a latent parameter called ``age'' $\eta$ is generated from a Gaussian with unit variance [HOGG CHECK] and mean $13 - (13/14)\,r$
+    \item A floating-point value of a latent parameter called ``age'' $\eta$ is generated from a Gaussian with mean $14 - r$ and variance 4.
     \item The point is discarded if the age is outside the range $0<\eta<13$. If the point is not discarded:
-    \item A $K$-dimensional latent vector $\zeta$ with $K=14$ [HOGG CHECK] is generated from a zero-mean, unit-variance Gaussian.
+    \item A $K$-dimensional latent vector $\zeta$ with $K=14$ is generated from a Gaussian with zero mean and variance 1.
     \item A $M$-dimensional latent vector $\xi$ with $M=110$ is created by $\xi = \xi_\eta\,\eta + \xi_\zeta\cdot\zeta$, where $\xi_\eta$ is a $M$-vector and $\xi_\zeta$ is a $M\times K$ matrix.
-The elements of $\xi_\eta$ and $\xi_\zeta$ were drawn (before the start) from a Gaussian with zero mean and variance $0.00046$ and $0.071$ respectively [HOGG CHECK].
-    \item A label $y$ called ``measured age'' is generated from a Gaussian with unit variance [HOGG CHECK] and mean $\eta$.
-    \item A $M$-dimensional feature vector $x$ called ``the data'' is generated from a Gaussian with variance $0.0025$ and a mean set by taking $1 + \xi$ and rectifying all pixels to lie in the range between zero and unity.
-\end{enumerate}
+    The elements of $\xi_\eta$ and $\xi_\zeta$ were drawn (before the start) from Gaussians with zero mean and variances $0.01/K$ and $1/K$ respectively.
+    \item A rectified latent vector $\tilde{\xi}$ is created by taking $1 + \xi$ and rectifying all pixels to lie in the range $[0, 1]$.
+    \item A $M$-dimensional feature vector $x$ called ``the data'' is generated from a Gaussian with mean $\tilde{\xi}$ and variance $0.0025$.
+    \item A label $y$ called ``measured age'' is generated from a Gaussian with mean $\eta$ and variance 1.
+\end{hoggnumerate}
 The data generated this way are shown in the top panels of \figurename~\ref{fig:regression}.
-This toy will make a very general point, but it has a decidedly astronomical feel to it.
 The data contain a non-trivial relationship between label $y$ (measured age) $y$ and known parameter $r$ (guiding radius).
@@
 The test-set relationship is strongly biased, deviating from the true relationship at the edges of the $r$ range.
-Given the large size of the test set, the precision of the result is high: This bias in the estimated relationship is substantially larger than 20-sigma in the worst data points.
+Given the large size of the test set, the precision of the result is high: This bias in the estimated relationship is larger than 30-sigma in the worst data points.
 
@@
 
+HOGG: Link to notebook / github.
+
 \end{document}

### 2023-12-22 paper/main.tex
Appendix zeroth draft done

@@
 \usepackage{graphicx}
+\usepackage{url}
 
@@
 
-HOGG: DROP THIS?
-Differences aside, there is an argument to make that maybe the natural sciences could be better off if they moved at least slightly closer to the ML points of view on ontology and epistemology.
-In the long run, the data are more stable than the fundamental theories of physics:
-Newtonian gravity was replaced with general relativity, and we expect general relativity to be replaced by some kind of quantum gravitational theory.
-At each change, the ontology changes, while the data are stable.
-We are not recommending this, but noting that a robust philosophy of science might be more positively oriented towards the data than standard practice.
-
 \section{Why do we need machine learning in the natural sciences?}
@@
 
+HOGG: DROP THIS?
+All differences aside, there is an argument to make that maybe the natural sciences could be better off if they moved at least slightly closer to the ML points of view on ontology and epistemology.
+In the long run, the data are more stable than the fundamental theories of physics:
+Newtonian gravity was replaced with general relativity, and we expect general relativity to be replaced by some kind of quantum gravitational theory.
+At each change, the ontology changes, while the data are stable.
+We are not recommending this, but noting that a robust philosophy of science might be more positively oriented towards the data than standard practice.
+
 \paragraph{Acknowledgements:}
@@
 Also shown are the same solid red circles and open black squares as in the top-right plot.
-Black exes show the mean relationship between $\hat{y}$ and $r$.
-The relationship shown by the exes is very precise but biased far away from the true relationship, unlike the relationship shown by the open squares (measured in the training set alone).\label{fig:regression}}
+Black X-shaped symbols show the mean relationship between $\hat{y}$ and $r$.
+The relationship shown by the Xs is very precise but biased far away from the true relationship, unlike the relationship shown by the open squares (measured in the training set alone).\label{fig:regression}}
 \end{figure}
 
-The regression-estimated labels $\hat{y}$ are related to the measured labels $y$ in the validation set with a linear relationship with unit slope and zero intercept.
-They are not precisely estimated, but there is no evidence for bias at the individual-object level.
+The regression-estimated labels $\hat{y}$ are related to the measured labels $y$ in the held-out validation set with a linear relationship with unit slope and zero intercept.
+They are not precisely estimated to be sure, but there is no evidence for strong bias at the individual-object level.
 
@@
 These two relationships are statistically consistent.
-The empirical relationship in the test-set data between the estimated label $\hat{y}$ and the known parameter $r$ in the test-set data is shown in \figurename~\ref{fig:regression} with black exes.
+The empirical relationship in the test-set data between the estimated label $\hat{y}$ and the known parameter $r$ in the test-set data is shown in \figurename~\ref{fig:regression} with black X-shaped symbols.
 The test-set relationship is strongly biased, deviating from the true relationship at the edges of the $r$ range.
@@
 
-HOGG: What to do? Just use the training set!
+In this case, what should the investigator do---the investigator who wants to know the relationship between $y$ (age) and $r$ (guiding radius)?
+The investigator should \emph{just use the training set}.
+Transferring labels to the test set did not help, even though the test set is far larger than the training set.
+The test-set results are very precise (error bars on the X-shaped symbols cannot be visibly shown in \figurename~\ref{fig:regression}), but they are very wrong.
+The training-set results are less precise but unbiased; they are much closer to the truth.
 
-HOGG: Note that none of these problems come from bad data, out-of-distribution, or complexity.
+Indeed, none of the issues here come from out-of-distribution problems or distribution shifts.
+All three data sets (train, validate, and test) are drawn from an identical parent population.
+When distribution shifts are involved (usually the labeled data are the best data, or the earliest, or the easiest to obtain), these kinds of problems only get worse.
 
-HOGG: Link to notebook / github.
+Related to that: The most interesting aspect of this problem is that there is no \emph{wrong} bias:
+The bias introduced from the training set is a correct bias; the training set represents a prior that accords with our beliefs about the test set.
+And yet the population-level inference is biased.
+The averaging of points shown by the X-shaped symbols in \figurename~\ref{fig:regression} is a simple misuse of the regression outputs; that point is trivial.
+However, it is a misuse that is occurring repeatedly in astrophysics with catalogs generated by ML regressions HOGG CITE.
+
+All code (including the data-generating code) used in this toy are available under an open-source license at \url{https://github.com/davidwhogg/BadForScience}.
 

### 2023-12-26 paper/main.tex
uses of ML in science

@@
 % ------
-% - Write an outline.
+% - Change it to be about PHYSICS rather than natural science?
 % - Fill in the outline.
@@
   Here we identify locations for ML in the natural sciences where the ontology and epistemology are valuable.
-  For example, when a flexible machine-learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can sometimes make the results more trustworthy.
+  For example, when an expressive machine-learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can sometimes make the results more trustworthy.
   We also show that the there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
-  For one, when flexible models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
-  For another, when flexible regressions are used to make datasets, the elements of those datasets cannot be used in joint analyses without taking on uncontrolled biases.
+  For one, when expressive models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
+  For another, when expressive regressions are used to label datasets, those labels cannot be used in downstream joint analyses without taking on uncontrolled biases.
   These remarks are intended to apply to all of the natural sciences, but most of the specific examples are in the astrophysics domain.
@@
 Many of these places are in the operational parts of scientific projects.
-  \item Further, we demonstrate that there are contexts inside a natural-scientific analysis in which using the most flexible ML method delivers \emph{the most conservative approach} to the scientific problem.
+  \item Further, we demonstrate that there are contexts inside a natural-scientific analysis in which using the most expressive ML method delivers \emph{the most conservative approach} to the scientific problem.
 An example is in causal contexts, where providing immense flexibility to (say) a confounder model can strengthen conclusions about the causal channels of interest.
@@
   \item[Label transfer:] Sometimes a project has informative data on a very large number of objects, but precise labels for only a few, maybe obtained through very careful analysis or external data.
-  In this case, if it is extremely expensive to label more objects, a regression can be trained on the labeled data points and then the trained regression used to label
+  In this case, if it is extremely expensive to label more objects, a regression can be trained on the few labeled data points and then the trained regression used to label all the rest.
+  Below (\sectionname~\ref{sec:bad}) we argue that this is not an advisable use of ML in the natural sciences if the expectation is that the regression-generated labels are going to be used in downstream scientific projects.
   \item[Classification:] 
-  \item[Speeding up decisions:]
-  \item[Speeding up simulations:]
-  \item[Modeling nuisances:]
-  \item[Outlier detection:]
-  \item[Making discoveries?]
+  \item[Speeding up decisions:] Many scientific projects must make decisions very fast in real time.
+  The most extreme examples of this are in particle physics, where detectors (like \textsl{Atlas} \cite{atlas}) must decide whether to trigger a data-saving event in a tiny fraction of second.
+  It is often the case that trained ML classifiers can reproduce the selection boundaries as well---or nearly as well---as first-principles models, but with far less computation.
+  \item[Speeding up simulations:] In much of the natural sciences, the theory in play is a computation or simulation or digital twin.
+  These simulations tend to be very computationally expensive, since they often span large ranges of spatial or time scales.
+  ML regressions can be trained to emulate the simulations, or patch up low-resolution simulations to higher resolutions.
+  HOGG: THERE ARE MANY different approaches to emulation; summarize here?
+  Below (\sectionname~\ref{sec:bad}) we argue that the introduction of ML-based emulators can introduce unwanted confirmation bias.
+  \item[Modeling nuisances:] In most natural-science domains, the quantities of interest are not directly observable, but rather the model or prediction for the data is a combination of the theory of relevance plus auxilliary theories of foregrounds, backgrounds, instrumentation, and noise sources.
+  When the goal is for these nuisances to be effectively modeled but not necessarily understood in detail, ML approaches can be effective and (as we argue in \sectionname~\ref{sec:good}) even conservative.
+  \item[Outlier detection:] In many of the sciences, discovery of new or previously unknown phenomena or objects can be of great importance.
+  A recent example in astronomy is the discovery of fast radio bursts \cite{grbs}, which started as nuisances in imaging projects and have turned out to be interesting astrophysical objects.
+  Since expressive unsupervised ML methods can describe accurately a complex distribution of data, they can also be used to find rare data points that are unlike any elements of the training data.
+  \item[Making discoveries?] There is a new hope in the sciences that sufficiently trained or constrained models might lead to new insights about scientific theories or might effectively make fundamental discoveries.
+  There are approaches along the direction of symbolic regression \cite{symbolic1, symbolic2}, and there are approaches along the lines of foundation models \cite{foundation1, foundation2}.
 \end{description}
@@
 In Bayesian language, this problem is closely related to the point that when data are to be combined, they should be combined at the likelihood level, not the posterior level.
-Likelihoods are not affected by population-level biases, and potentially produce unbiased estimates.
+Likelihoods are not affected by population-level biases.
+Likelihoods can produce unbiased point estimates.
 Likelihood-based estimators are not usually the lowest-risk estimators, but they can be low- or zero-bias estimates.
@@
 Distribution shifts just tend to make these kinds of problems worse.
+
+HOGG: DO WE CLEARLY SAY: ML estimates are always biased, maximum-likelihood estimates sometimes aren't. That's important if you are releasing data for downstream re-use.
 
@@
 \section{When is machine learning good for natural science?}\label{sec:good}
+HOGG: BE SURE to say what is in common across all of these examples. They are joined by a thread!
+
+\paragraph{Execution and operations:}
 
@@
 \section{Population-level biases in regression outputs}\label{app:toy}
-A machine-learning (ML) regression learns the best set of parameters (weights and thresholds, for example) $\hat{\theta}$ of a flexible function $f(x;\theta)$ from a training set of labeled data.
+A machine-learning (ML) regression learns the best set of parameters (weights and thresholds, for example) $\hat{\theta}$ of an expressive function $f(x;\theta)$ from a training set of labeled data.
 The learned function $f(x;\hat{\theta})$ delivers a predicted or estimated label $\hat{y}$ for any data point with a complete set of features $x$.
@@
     \item A floating-point value of a latent parameter called ``age'' $\eta$ is generated from a Gaussian with mean $14 - r$ and variance 4.
-    \item The point is discarded if the age is outside the range $0<\eta<13$. If the point is not discarded:
+    \item The point is discarded if the age is outside the range $0<\eta<14$. If the point is not discarded:
     \item A $K$-dimensional latent vector $\zeta$ with $K=14$ is generated from a Gaussian with zero mean and variance 1.
@@
 This relationship is shown in \figurename~\ref{fig:regression}.
-It isn't precisely linear because of the censoring of data points with latent ages outside the range $0<\eta<13$.
+It isn't precisely linear because of the removal of data points with latent ages $\eta$ outside the range $0<\eta<14$.
 
@@
 The model is validated on a validation set of 2048 data points with features $x$ and labels $y$.
-The test set is a much larger set of 105,864 data points with features $x$ and no labels.
+The test set is a much larger set of $10^5$ data points with features $x$ and no labels.
 The model is used to make estimated labels $\hat{y}$ in the full test set.
@@
 The regression-estimated labels $\hat{y}$ are related to the measured labels $y$ in the held-out validation set with a linear relationship with unit slope and zero intercept.
-They are not precisely estimated to be sure, but there is no evidence for strong bias at the individual-object level.
+The comparison to the validation data (bottom-left panel of \figurename~\ref{fig:regression}) shows that they are not precisely estimated, but there is no evidence for strong bias at the individual-object level.
 
@@
 The empirical relationship between label $y$ and known parameter $r$ in the training-set data is shown in \figurename~\ref{fig:regression} with open black squares.
-These two relationships are statistically consistent.
+These two relationships are statistically consistent, as expected.
 The empirical relationship in the test-set data between the estimated label $\hat{y}$ and the known parameter $r$ in the test-set data is shown in \figurename~\ref{fig:regression} with black X-shaped symbols.
 The test-set relationship is strongly biased, deviating from the true relationship at the edges of the $r$ range.
-Given the large size of the test set, the precision of the result is high: This bias in the estimated relationship is larger than 30-sigma in the worst data points.
+
+HOGG: TAKE A PARAGRAPH TO EXPLAIN THIS CLAIM: Given the large size of the test set, the precision of the result is high: This bias in the estimated relationship is larger than 30-sigma in the worst data points.
 

### 2023-12-26 paper/main.tex
tweaks to the appendix and etc

@@
   We also show that the there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
-  For one, when expressive models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
+  For one, when ML models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
   For another, when expressive regressions are used to label datasets, those labels cannot be used in downstream joint analyses without taking on uncontrolled biases.
@@
   Adrian Price-Whelan (Flatiron),
+  Hans-Walter Rix (MPIA),
   Sam Roweis (deceased),
@@
   Soledad Villar (JHU)
-for conversations over the years related to these arguments.
+for conversations over the years related to these arguments, and
+  Vedant Chandra (Harvard)
+for discussions of the bias in regression outputs.
 
@@
 \section{Population-level biases in regression outputs}\label{app:toy}
-A machine-learning (ML) regression learns the best set of parameters (weights and thresholds, for example) $\hat{\theta}$ of an expressive function $f(x;\theta)$ from a training set of labeled data.
+A machine-learning (ML) regression learns the best (that it can find) value of a list or vector $\hat{\theta}$ of parameters (weights and thresholds, for example) of an expressive function $f(x;\theta)$ that takes as input a feature list or vector $x$ and outputs a label estimate $y$.
+The parameters $\hat{\theta}$ are learned from a training set of labeled data, which is a set of $(x, y)$ pairs.
 The learned function $f(x;\hat{\theta})$ delivers a predicted or estimated label $\hat{y}$ for any data point with a complete set of features $x$.
-Thus an estimated label $\hat{y}_\ast$ for a data point with features $x_\ast$ has taken information \emph{both} from the features $x_\ast$ and \emph{also} from the features and labels of all the data points in the training set used to set the parameters $\hat{\theta}$.
+Thus a particular estimated label $\hat{y}_\ast$ for a particular data point with features $x_\ast$ obtains information \emph{both} from the features $x_\ast$ and \emph{also} from the features and labels of all the data points in the training set used to set the parameters $\hat{\theta}$.
 In applications in which regression-estimated labels are used, it matters \emph{how much information} is coming from the individual data-point's features and how much from the original training set.
@@
 
+HOGG: COMMENT OR SHOW that the bias goes down as the information in the features goes up (relative to the information in the training-set distribution (the prior)).
+
 In this case, what should the investigator do---the investigator who wants to know the relationship between $y$ (age) and $r$ (guiding radius)?

### 2023-12-29 paper/main.tex
lots of content, especially discussion

@@
 % - Be very careful about the SUBJECT of sentences. Is it ML that is bad? Is it the practice that is bad? Is it the practitioner that is bad?
+% - Should we use NS for natural science?
+% - How many authors are there? Should we switch to first-person singular?
 % - [more here]
@@
 
-\paragraph{Ontology:}
+\paragraph{What is natural science?}
+For the purposes of this \documentname, we will employ a restrictive definition of natural science.
+For us, the natural sciences are the sciences that study the natural world, with the primary aim of \emph{understanding}.
+Examples include physics, chemistry, biology, Earth science, ecology, and so on.
+For our purposes, the natural sciences are about understanding observed phenomena, unifying knowledge, making predictions for new experiments and observations.
+Natural-science questions include things like: How did our Solar System form? Or how do cells differentiate as the embryo grows? Or what causes the jet stream?
+
+For the purposes of this \documentname{} we will exclude more engineering-oriented questions such as: What protein sequences might be important for the treatment of diabetes?
+Or where in Canada might there be new uranium deposits?
+Or what kinds of catalytic surfaces might help turn biomass into biofuel?
+This is a bit unfair!
+We are excluding these kinds of questions in part because it is obvious that ML methods \emph{can} serve important purposes in such problems.
+That will re-emerge as a theme, below.
+
+\paragraph{Ontologies:}
 Unsupervised ML methods deliver descriptions or compressions of data.
@@
 
-Models are deemed stably optimized and useful not if the latent space is stable, but rather if the predictions for held-out data are stable.
+In ML, models are deemed stably optimized and useful not if the latent space---the joint values of all the weights and biases---is stable, but rather if the predictions for held-out data are stable.
 Thus it is the data that exist in ML contexts, not the latent parameters.
 
-\paragraph{Epistemology:}
+In contrast, the ontologies of the natural sciences contain far more things than just the data.
+In physics, for example, 
+not only do the data exist, but so do forces, energies, momenta, charges, spacetime, wave functions, virtual particles, and much more.
+These entities are judged to exist in part because they are involved in the latent structure of the successful theories; almost none of them are direct observables.
+The most important discoveries in the natural sciences are discoveries of latent structure: Natural selection as an explanation for the differentiated properties of species \cite{natural_selection}, for example, or the elemental composition of the Sun \cite{sun_composition}, or the quarks and gluons that make up the proton \cite{proton_substructure}.
+
+\paragraph{Epistemologies:}
 A trained ML model is deemed successful or correct if \emph{it performs well on held-out training data}.
@@
 
-A critic of our epistemological claim could point to the large literature on out-of-sample generalization, transfer learning \cite{transfer}, or the more grand contemporary idea of foundation models \cite{foundation}.
+A critic of our ML epistemology claim could point to the large literature on out-of-sample generalization, transfer learning \cite{transfer}, or the more grand contemporary idea of foundation models \cite{foundation}.
 But even in these contexts, models are deemed successful when they explain new or held-out data; the latent structure of the models is not a primary consideration when the validity of the model is in question.
 
-\paragraph{Natural sciences:}
-Ontologies and epistemologies are somewhat different in different scientific fields---and among different practitioners---but in almost no natural-scientific context do the ontology and epistemology match those of ML.
+Another critic might point to the literatures on interpretable ML \cite{interpretable} and explainable ML \cite{explainable}.
+In these ML subfields, in some cases, attempts are made to understand the internal structure or the behavior of the ML model.
+There are not enormous successes here, but if there were, these might be subfields in which the epistemology of ML gets more complex; maybe in the future some ML models will have to perform well on data \emph{and} be explainable in some important sense.
+At present this isn't part of any ML epistemology.
 
-In physics, for example, the ontology is much broader than it is in ML.
-Not only do the data exist, but so do forces, energies, momenta, charges, spacetime, wave functions, virtual particles, and much more.
-These entities are judged to exist in part because they are involved in the latent structure of the successful theories; almost none of them are direct observables.
-
-In physics the epistemology is much more restrictive than it is in ML.
-A physics model---which, as we note, is almost always a model of latent structure---is judged to be good or strongly confirmed if it explains some aspects of different data in multiple domains, and if it connects in natural ways to other theories or principles (such as conservation laws and invariances) that are strongly confirmed themselves.
-It is worthy of note here that a successful theory is usually a combination of a simple theory of the latents, and a less simple observation model, which explains how the latents appear in the data space.
+In the natural sciences, in contrast, the epistemologies are much more restrictive and much more demanding than they are in ML.
+The detailed epistemological framework depends on the natural-science field, and even on the practitioner within the field.
+However, in all cases, it is more demainding:
+A theory or explanation has to do much more than just explain the data in order to be accepted as true.
+In physics for example, a model---which, as we note, is almost always a model of latent structure---is judged to be good or strongly confirmed not just if it explains observed data.
+It must explain data in multiple domains, and it must also connect in natural ways to other theories or principles (such as conservation laws and invariances) that are strongly confirmed themselves.
+(It is worthy of note here that a successful theory in the natural sciences is usually a combination of a more fundamental theory of the latents, and a less fundamental or auxiliary observation model, which explains how the latents appear in or affect the observable data \cite{auxiliary}.)
+General relativity \cite{gr} was adopted not primarily because it explained anomalous data (although it did explain some); it was adopted because it had a beautiful structure, it resolved conceptual paradoxes in the pre-existing theory of gravity, and it was consistent with emerging ideas of field theory and geometry.
 
@@
   Below (\sectionname~\ref{sec:bad}) we argue that this is not an advisable use of ML in the natural sciences if the expectation is that the regression-generated labels are going to be used in downstream scientific projects.
-  \item[Classification:] 
+  \item[Classification:] This is the same use case as label transfer, but in the specific case in which the labels are drawn from a finite (small, actually) set of discrete values.
   \item[Speeding up decisions:] Many scientific projects must make decisions very fast in real time.
-  The most extreme examples of this are in particle physics, where detectors (like \textsl{Atlas} \cite{atlas}) must decide whether to trigger a data-saving event in a tiny fraction of second.
+  The most extreme examples of this are in particle physics, where detectors (such as the \textsl{Atlas} Experiment \cite{atlas} at the \textsl{Large Hadron Collider} \cite{lhc}) must decide whether to trigger a data-saving event in a tiny fraction of second.
   It is often the case that trained ML classifiers can reproduce the selection boundaries as well---or nearly as well---as first-principles models, but with far less computation.
@@
 The short summary is that we need ML in the natural sciences; we can't live without it.
-The question here is how we will use it safely?
+The question here is---given the restrictive ontology and loose epistemology---how we will use it safely?
 
 \section{When is machine learning bad for natural science?}\label{sec:bad}
-In this \sectionname{} we elucidate two statistical biases that can be introduced when ML methods are introduced into a natural-science project.
-Neither of these biases can be easily corrected or removed.
+In this \sectionname{} we elucidate two statistical biases that can be introduced into a natural-sciences project when ML methods are introduced.
+Neither of these biases can be easily corrected or removed, to our knowledge.
 
@@
 
-Hello World!
+We hope we have clearly answered the question in the title with the response ``Both!''
+Machine learning (ML) is a reality for the natural sciences:
+We need it for speed and scale in many contexts in many contemporary scientific projects, where instrument bandwidths, data volumes, assay parallelization, experimental automation, and scientific ambitions are all growing.
+It is also the case that PhD students and practitioners in the natural sciences want to---or even need to---learn ML practices to have useful and transferable skills.
+ML is here, and it is here to stay.
 
-HOGG: DROP THIS?
-All differences aside, there is an argument to make that maybe the natural sciences could be better off if they moved at least slightly closer to the ML points of view on ontology and epistemology.
+That said, there is a certain amount of hype in this area, and we are seeing scientists ``move fast and break things'' \cite{zuck} all over the place.
+If anything, this \documentname{} represents a call to slow down and think more.
+ML tools are incredibly powerful, if we put them to use in the right places in the natural sciences.
+
+There is a famous ancient quotation about ``the unreasonable effectiveness of mathematics in physics'' \cite{weiner_maybe}.
+Mathematics has been pursued (in many cases) for abstract reasons of beauty and technical challenge, but it has ended up creating a language for many of the natural sciences.
+Similarly, we could coin a phrase about \emph{the unreasonable effectiveness of machine learning in the natural sciences}.
+It's unreasonable because ML methods were designed---and, importantly, developed, engineered, and optimized---to solve problems for commerce.
+The most remarkable ML systems have been driven by critical industrial questions such as: Which advertisement should I display on this web search result?
+What direction should I steer this automobile?
+How could I make a chair that looks like an avocado?
+And which online videos contain kittens?
+In some sense it is remarkable that there is \emph{any} overlap between these technologies and the things we need in scientific domains.
+
+But that said, the industrial successes of ML can also be misleading.
+The fact that big ML models can solve countless problems in business does not mean that all areas of the natural sciences will be helped by the introduction of ML.
+You do not have to \emph{understand} your customers (nor their videos) in order to make plenty of revenue off of them; in the natural sciences, it is understanding, not revenue, that we seek.
+
+Finally, we make a philosophical comment:
+In physics---and in most of the natural sciences we believe---the same phenomena can be explained with many qualitatively different theories \cite{philosphy_of_physics}; mathematical theories are not usually unique.
+In the same way (as we noted above), ML models contain massive degeneracies, and even different architectures can lead to near-identical predictions after training on the same training data.
+Does this lead to a philosophical connection between ML and the natural sciences?
+We have argued elsewhere \cite{hogg} that the degeneracies in the natural sciences undermine \emph{realism} or the belief in the literal truth of the latent theories or objects of physics.
+There is an argument to be made that maybe the natural sciences could be better off if they moved at least slightly closer to the ML points of view on ontology and epistemology.
 In the long run, the data are more stable than the fundamental theories of physics:
@@
 At each change, the ontology changes, while the data are stable.
-We are not recommending this, but noting that a robust philosophy of science might be more positively oriented towards the data than standard practice.
+We are not recommending anything (yet), we are merely noting that a robust philosophy of natural science might be more positively oriented towards the data than current standard practice.
 

### 2023-12-30 paper/main.tex
zeroth draft

@@
 % - Publish on the arXiv.
-% - format for ICML Position Paper and submit.
+% - set formatting and anonymization switches to go between arXiv and ICML Position Paper
+% - submit to ICML.
 % - Wait for the call from Stockholm.
@@
 % - How many authors are there? Should we switch to first-person singular?
+% - Audit for flexible, expressive, high in capacity, and so on.
 % - [more here]
 
-\documentclass[11pt]{article}
+\documentclass[10pt]{article}
 \usepackage[utf8]{inputenc}
@@
 % Hogg typesetting issues
-\setlength{\textwidth}{6.00in}
-\setlength{\oddsidemargin}{0.25in}
+\setlength{\textwidth}{5.50in}
+\setlength{\oddsidemargin}{0.00in}
 \setlength{\oddsidemargin}{0.00in}
@@
 \newcommand{\documentname}{\textsl{Position Paper}}
+\newcommand{\draftstatus}{\textcolor{red}{Draft version 2023-12-30}}
 \newcommand{\sectionname}{Section}
@@
 \pagestyle{myheadings}
-\markboth{foo}{\textcolor{gray}{\textsf{Hogg / Is machine learning good or bad for the natural sciences?}}}
+\markboth{foo}{\textcolor{gray}{\textsf{Hogg / Is machine learning good or bad for the natural sciences? \draftstatus}}}
 \linespread{1.08}
@@
 
-\title{\documentname: Is machine learning good or bad for the natural sciences?}
+\title{Is machine learning good or bad for the natural sciences?}
+\noindent
+\draftstatus
 
@@
 Many of these places are in the operational parts of scientific projects.
-  \item Further, we demonstrate that there are contexts inside a natural-scientific analysis in which using the most expressive ML method delivers \emph{the most conservative approach} to the scientific problem.
-An example is in causal contexts, where providing immense flexibility to (say) a confounder model can strengthen conclusions about the causal channels of interest.
+  \item Further, we argue that there are contexts inside a natural-scientific analysis in which using the most expressive ML method delivers \emph{the most conservative approach} to the scientific problem.
+An example is in causal contexts, where making a confounder model more expressive can strengthen conclusions about the causal channels of interest.
 \end{itemize}
@@
 
-\section{Why do we need machine learning in the natural sciences?}
+\section{Why do we need machine learning in the natural sciences?}\label{sec:why}
 In this \sectionname{} we give some idea of how and why ML has had such a big impact in the natural sciences, and how and why many scientific projects require ML technologies.
@@
 \end{description}
-The short summary is that we need ML in the natural sciences; we can't live without it.
+This list isn't exhaustive!
+For example, it doesn't mention the use of ML to find summary statistics for the comparison of simulations and data in (say) simulation-based inferences \cite{sbi}.
+It doesn't mention using ML to emulate likelihoods or likelihood ratios \cite{biwei, kyle}.
+The uses of ML in the natural sciences is legion.
+
+The short summary of all this is that we need ML in the natural sciences; we can't live without it.
 The question here is---given the restrictive ontology and loose epistemology---how we will use it safely?
@@
 
-HOGG: DO WE CLEARLY SAY: ML estimates are always biased, maximum-likelihood estimates sometimes aren't. That's important if you are releasing data for downstream re-use.
-
 \paragraph{Emulator-induced confirmation bias:}
@@
 \section{When is machine learning good for natural science?}\label{sec:good}
-HOGG: BE SURE to say what is in common across all of these examples. They are joined by a thread!
+In \sectionname~\ref{sec:intro} we re-phrased the question in the title as
+``Where, in the natural sciences, can you use a model you don't understand?''
+The answer is: You can use it for the parts of your scientific project or experiment or analysis that \emph{you don't need to understand}.
+We have colleagues who believe that there is no such part.
+We disagree: For example in astronomy, we use charge-coupled devices to make extremely precise measurements of stars, without knowing in detail the solid-state physics that makes them efficient, linear detectors of photons.
+There are many parts of a scientific project that must work well, but which do not need to be carefully controlled or understood (in advance, anyway).
 
-\paragraph{Execution and operations:}
+Here we give a few examples.
+These are not intended to be exhaustive; they are just examples of places within a scientific project ML can be used to the benefit---and not detriment---of the natural-science goals.
+
+\paragraph{Real-time execution and operations:}
+In many scientific projects, experiments are adaptive to real-time outcomes, or data are taken selectively, or limited resources are assigned to particular targets.
+When bandwidths are high, or experimental timescales are short, sometimes the decisions that need to be made in real time cannot be made with the first-principles decision-making that the investigator would like.
+An example mentioned above is the selection of events for storage at the \textsl{LHC Atlas} Experiment \cite{atlas}.
+An example from astronomy is the upcoming \textsl{Rubin Observatory LSST} project \cite{rubin}, which will generate some $10^5$ events of possible interest for follow-up observing every night \cite{lsst_events}.
+These decisions can be made with very fast classifiers, trained on well-studied training data or simulations \cite{lsst_broker, atlas_trigger}.
+
+Is it unsafe to operate a project using ML decision-making systems?
+Even if these systems do not have to be extremely well understood in real time, but they do need to be \emph{versioned} and preserved for study later.
+Statistical projects making use of the data from systems like this will need to understand the statistical biases created by the ML selection procedures.
+This can be done after the fact, provided that the decision-making system is preserved in a precisely reproducible state, such that it can be run offline on counterfactual, simulated, and subsequent data.
+
+\paragraph{Discoveries of outliers and rare objects:}
+As mentioned in \sectionname~\ref{sec:why}, ML is well suited to outlier detection.
+Since ML methods are expressive, and trained to fit the data, new data that \emph{aren't} well described by a trained ML model are possibly interesting.
+Outlier-detection systems can be used to identify time intervals in which the equipment is malfunctioning.
+They can also be used to identify objects that are rare or unusual.
+Many discoveries in astrophysics are discoveries of new kinds of objects, originally found as data outliers \cite{quasars, vorweerp}.
+So far, ML has not been involved in any big discoveries in astronomy, but we expect this to be a productive avenue for ML in the natural sciences and astronomy in particular.
 
 \paragraph{Foreground, background, and confounder models:}
+We mentioned in \sectionname~\ref{sec:philosophy} that a natural-science theory is a combination of a fundamental, latent theory with a (usually complicated) auxiliary observation model that explains the details of the observations.
+The simplest part of this observation model is the model of the backgrounds, foregrounds, and confounders.
+For example, any maps of the cosmic microwave background (by the ESA \textsl{Planck} spacecraft \cite{planck_maps}, for example) will have emission from the intervening Milky Way and from radio galaxies, adding to, and distorting the maps.
+When the goal is to understand the cosmic microwave background, the investigator often doesn't need to understand these foregrounds \emph{physically}; an \emph{effective model} is sufficient.
+And indeed, all the bleeding-edge methods for this problem currently are ML methods \cite{cmb_foregrounds}.
+It doesn't matter to the questions of physical cosmology what those foregrounds \emph{mean}; and indeed with ML-based foreground removal, cosmic microwave background experiments have delivered the most precise measurements of the parameters of the cosmological model with sub-percent accuracy \cite{planck_parameters}.
 
-\paragraph{Outlier detection:}
+This is all true: For foregrounds and backgrounds, understanding is not always necessary.
+An even stronger statement can be made here, however, in the realm of causal inference:
+Sometimes the goal is to argue that an effect is being caused by some particular cause, and there are confounding causal inputs that might be distorting or influencing the data.
+The more expressive the model used to model the confounders, the stronger the conclusions that can be made about the causation \cite{causal_inference}.
+That is, the huge capacity of ML methods can made a causal claim \emph{very conservative}, and hence very robust.
+One way to see this is that if the confounder model has been given every chance it possibly can have to model the effect of interest, and it cannot, then it is a stronger conclusion that the effect is caused by the cause of interest.
+This is the strategy in contemporary causal-inference projects, which make use of ML methods for these reasons \cite{bart}.
+
+A key place for this kind of causal inference has been in instrument calibration.
+Data taken by an instrument have signals imprinted from the physical (or biological or chemical) processes of interest, and also from the details of the measurement hardware, such as sensitivity and bias variations, and time-dependent distortions of internal mappings.
+In the end, if the results of the investigation are to have the instrument calibrated out, it makes sense to give the instrument model a lot of expressive power, and judge that model in terms of its ability to explain the data.
+Hence, instrument models and calibration are important places for ML methods, and places where the introduction of ML methods can made scientific projects more conservative and more accurate.
+One very early example of this in astronomy was the calibration of the \textsl{Sloan Digital Sky Survey} imaging data \cite{ubercalibration}.
 
@@
 Similarly, we could coin a phrase about \emph{the unreasonable effectiveness of machine learning in the natural sciences}.
-It's unreasonable because ML methods were designed---and, importantly, developed, engineered, and optimized---to solve problems for commerce.
-The most remarkable ML systems have been driven by critical industrial questions such as: Which advertisement should I display on this web search result?
+It's unreasonable because ML methods were developed, engineered, and optimized to solve problems for commerce.
+The most remarkable ML systems have been driven by critical industrial questions such as:
+Which advertisement should I display on this web search result?
 What direction should I steer this automobile?

### 2023-12-30 paper/main.tex
references

@@
 % - Change it to be about PHYSICS rather than natural science?
-% - Fill in the outline.
+% - Get feedback from Yann, BS, MSV, JH, others
+% - put in all references
+% - audit bibliography; make sure arXiv references in the reference list are consistent?
 % - Publish on the arXiv.
@@
 
-In support of this position we could point to the literature on adversarial attacks \cite{adversarial1, adversarial2}.
+In support of this position we could point to the literature on adversarial attacks (eg, \cite{adversarial1}).
 This literature shows us, in dramatic ways, that ML methods are not doing what we (na\"ively) believe that humans or scientists or scientific theories do in comparable circumstances.
@@
   \item[Speeding up decisions:] Many scientific projects must make decisions very fast in real time.
-  The most extreme examples of this are in particle physics, where detectors (such as the \textsl{Atlas} Experiment \cite{atlas} at the \textsl{Large Hadron Collider} \cite{lhc}) must decide whether to trigger a data-saving event in a tiny fraction of second.
+  The most extreme examples of this are in particle physics, where detectors (such as the \textsl{Atlas} Experiment \cite{atlas} at the \textsl{Large Hadron Collider}) must decide whether to trigger a data-saving event in a tiny fraction of second.
   It is often the case that trained ML classifiers can reproduce the selection boundaries as well---or nearly as well---as first-principles models, but with far less computation.
@@
 They can also be used to identify objects that are rare or unusual.
-Many discoveries in astrophysics are discoveries of new kinds of objects, originally found as data outliers \cite{quasars, vorweerp}.
+Many discoveries in astrophysics are discoveries of new kinds of objects, originally found as data outliers \cite{quasars, voorwerp}.
 So far, ML has not been involved in any big discoveries in astronomy, but we expect this to be a productive avenue for ML in the natural sciences and astronomy in particular.
@@
 
-{\raggedright
+{\raggedright\footnotesize
 \bibliography{ml_in_astro}

### 2023-12-31 paper/main.tex
references, tweaks

@@
 % - Change it to be about PHYSICS rather than natural science?
-% - Get feedback from Yann, BS, MSV, JH, others
+% - Get feedback from BS, MSV, JH, others
+% - Update appendix with feedback from Vedant Chandra (email).
 % - put in all references
-% - audit bibliography; make sure arXiv references in the reference list are consistent?
+% - audit bibliography; make sure arXiv references in the reference list are consistent? Also capitalization!
 % - Publish on the arXiv.
@@
 
-% Hogg typesetting issues
-\setlength{\textwidth}{5.50in}
-\setlength{\oddsidemargin}{0.00in}
-\setlength{\oddsidemargin}{0.00in}
-\setlength{\textheight}{9.25in}
-\setlength{\topmargin}{-0.50in}
-\setlength{\parindent}{1.2\baselineskip} % seriously
-\renewcommand{\title}[1]{\section*{\raggedright #1}}
-\renewcommand{\author}[1]{\medskip\par\noindent\textbf{#1}}
-\newcommand{\affil}[1]{{\footnotesize\par\noindent #1 \par}}
-\renewenvironment{abstract}{\paragraph{Abstract:}}{}
+% Hogg typesetting issues that won't conflict with ICML
+\frenchspacing
 \newcommand{\documentname}{\textsl{Position Paper}}
-\newcommand{\draftstatus}{\textcolor{red}{Draft version 2023-12-30}}
+\newcommand{\draftstatus}{\textcolor{red}{Draft version 2023-12-31}}
 \newcommand{\sectionname}{Section}
+\newcommand{\secref}[1]{\sectionname~\ref{#1}}
+\newcommand{\figref}[1]{\figurename~\ref{#1}}
 \newenvironment{hoggnumerate}
@@
 {\end{list}}
+
+% Hogg typesetting issues that will conflict with ICML
+\setlength{\textwidth}{5.50in}
+\setlength{\oddsidemargin}{-0.25in}
+\setlength{\textheight}{9.25in}
+\setlength{\topmargin}{-0.50in}
+\setlength{\parindent}{1.2\baselineskip} % seriously
+\renewcommand{\title}[1]{\section*{\raggedright #1}}
+\renewcommand{\author}[1]{\medskip\par\noindent\textbf{#1}}
+\newcommand{\affil}[1]{{\footnotesize\par\noindent #1 \par}}
+\renewenvironment{abstract}{\paragraph{Abstract:}}{}
 \pagestyle{myheadings}
@@
 \linespread{1.08}
-\frenchspacing\raggedbottom\sloppy\sloppypar
+\raggedbottom\sloppy\sloppypar
 
@@
 Another is a more standard estimator bias that is (possibly enormously) amplified when elements of datasets produced by ML regressions are used in combination or joint analyses.
+These biases can't be easily corrected.
   \item We show that there are many safe places for the use of ML methods in current natural-science practices, and places where (in the contemporary context) the use of ML methods is effectively \emph{required}.
 Many of these places are in the operational parts of scientific projects.
-  \item Further, we argue that there are contexts inside a natural-scientific analysis in which using the most expressive ML method delivers \emph{the most conservative approach} to the scientific problem.
-An example is in causal contexts, where making a confounder model more expressive can strengthen conclusions about the causal channels of interest.
+  \item We argue that in causal contexts, in which the ML method is used to model foregrounds, backgrounds, calibration parameters, or other confounders, using the most expressive ML method can deliver \emph{the most conservative approach} to the scientific problem.
 \end{itemize}
@@
   In this case, if it is extremely expensive to label more objects, a regression can be trained on the few labeled data points and then the trained regression used to label all the rest.
-  Below (\sectionname~\ref{sec:bad}) we argue that this is not an advisable use of ML in the natural sciences if the expectation is that the regression-generated labels are going to be used in downstream scientific projects.
+  Below (\secref{sec:bad}) we argue that this is not an advisable use of ML in the natural sciences if the expectation is that the regression-generated labels are going to be used in downstream scientific projects.
   \item[Classification:] This is the same use case as label transfer, but in the specific case in which the labels are drawn from a finite (small, actually) set of discrete values.
+  It is subject to the same kinds of biases described below (\secref{sec:bad}).
   \item[Speeding up decisions:] Many scientific projects must make decisions very fast in real time.
@@
   It is often the case that trained ML classifiers can reproduce the selection boundaries as well---or nearly as well---as first-principles models, but with far less computation.
+  It is critical, if ML methods are used for real-time operations, that the methods and all their latent parameters (weights) be preserved for analysis, operations simulations, and conterfactual exercises, often performed long after the data are taken.
   \item[Speeding up simulations:] In much of the natural sciences, the theory in play is a computation or simulation or digital twin.
@@
   ML regressions can be trained to emulate the simulations, or patch up low-resolution simulations to higher resolutions.
-  HOGG: THERE ARE MANY different approaches to emulation; summarize here?
-  Below (\sectionname~\ref{sec:bad}) we argue that the introduction of ML-based emulators can introduce unwanted confirmation bias.
+  Below (\secref{sec:bad}) we argue that the introduction of ML emulators can introduce an unwanted, strong confirmation bias.
   \item[Modeling nuisances:] In most natural-science domains, the quantities of interest are not directly observable, but rather the model or prediction for the data is a combination of the theory of relevance plus auxilliary theories of foregrounds, backgrounds, instrumentation, and noise sources.
-  When the goal is for these nuisances to be effectively modeled but not necessarily understood in detail, ML approaches can be effective and (as we argue in \sectionname~\ref{sec:good}) even conservative.
+  When the goal is for these nuisances to be effectively modeled but not necessarily understood in detail, ML approaches can be effective and (as we argue in \secref{sec:good}) even conservative.
   \item[Outlier detection:] In many of the sciences, discovery of new or previously unknown phenomena or objects can be of great importance.
@@
   There are approaches along the direction of symbolic regression \cite{symbolic1, symbolic2}, and there are approaches along the lines of foundation models \cite{foundation1, foundation2}.
+  So far this hope has borne no important new discoveries, but discoveries are not inconceivable.
 \end{description}
@@
 For example, it doesn't mention the use of ML to find summary statistics for the comparison of simulations and data in (say) simulation-based inferences \cite{sbi}.
-It doesn't mention using ML to emulate likelihoods or likelihood ratios \cite{biwei, kyle}.
+It doesn't mention using ML to emulate likelihoods or likelihood ratios \cite{biwei, likelihood_ratio}.
 The uses of ML in the natural sciences is legion.
@@
 Most ML regressions are not capable of generating or emulating likelihoods.
-There are interesting approaches in development to replace regressions with generative models that can produce likelihoods HOGG CITE THINGS.
+There are interesting approaches in development to replace standard regressions with generative models that can produce likelihoods \cite{cannon, biwei, likelihood_ratio}.
 
@@
 In many fields in the natural sciences, the fundamental theoretical model is \emph{computational}, meaning that theoretical predictions are made with large computer simulations.
-These simulations are usually limited in the range of length scales and time scales which they can simultaneously compute.
-Comparison of data with simulations usually involves running enormous numbers of simulations \cite{sbi}, because simulations have to span ranges of fundamental parameters and initial conditions and so on.
+These simulations are usually being asked to handle large ranges of length scales and time scales, so they are generally getting larger and more computationally demanding each year.
+Quantitative comparison of data with simulations usually involves running enormous numbers of simulations \cite{abc}, because simulations have to span ranges of fundamental parameters and initial conditions and so on.
 Thus these simulation requirements are getting very large for contemporary research projects.
-For example, in cosmology contemporary experiments could easily consume the total computing capacity of the United States.
+For example, in cosmology, contemporary experiments could easily consume the total computing capacity of the United States.
 Hence we turn to emulators---trained ML regressions that have learned the input-output relationship of a simulation, or the relationship between an inexpensive low-resolution simulation and an expensive high-resolution experiment.
@@
 Recall that because these emulators are ML methods, they generally have uninterpretable internal parameters and weights, and they generally have been validated by comparison with held-out training data.
-Like all high-capacity ML methods, they generally are vulnerable to adversarial attacks.
+Like all high-capacity ML methods, they generally are vulnerable to adversarial attacks \cite{adversarial1}.
 
@@
 Experiment B finds something extremely surprising about the mass matrix for the neutrinos, in conflict with other measurements and our expectations.
-The comment is made, when the results of Experiment B come out, that perhaps the anomaly comes from something very slightly wrong or off in the trained ML emulator.
+Is it possible that the anomaly discovered by Experiment B isn't real; it just comes from something very slightly wrong or off in the trained ML emulator?
 
@@
 \section{When is machine learning good for natural science?}\label{sec:good}
-In \sectionname~\ref{sec:intro} we re-phrased the question in the title as
+In \secref{sec:intro} we re-phrased the question in the title as
 ``Where, in the natural sciences, can you use a model you don't understand?''
@@
 \paragraph{Discoveries of outliers and rare objects:}
-As mentioned in \sectionname~\ref{sec:why}, ML is well suited to outlier detection.
+As mentioned in \secref{sec:why}, ML is well suited to outlier detection.
 Since ML methods are expressive, and trained to fit the data, new data that \emph{aren't} well described by a trained ML model are possibly interesting.
@@
 \paragraph{Foreground, background, and confounder models:}
-We mentioned in \sectionname~\ref{sec:philosophy} that a natural-science theory is a combination of a fundamental, latent theory with a (usually complicated) auxiliary observation model that explains the details of the observations.
+We mentioned in \secref{sec:philosophy} that a natural-science theory is a combination of a fundamental, latent theory with a (usually complicated) auxiliary observation model that explains the details of the observations.
 The simplest part of this observation model is the model of the backgrounds, foregrounds, and confounders.
@@
 It doesn't matter to the questions of physical cosmology what those foregrounds \emph{mean}; and indeed with ML-based foreground removal, cosmic microwave background experiments have delivered the most precise measurements of the parameters of the cosmological model with sub-percent accuracy \cite{planck_parameters}.
+For foregrounds and backgrounds, understanding is not always necessary.
 
-This is all true: For foregrounds and backgrounds, understanding is not always necessary.
 An even stronger statement can be made here, however, in the realm of causal inference:
@@
 
-That said, there is a certain amount of hype in this area, and we are seeing scientists ``move fast and break things'' \cite{zuck} all over the place.
+That said, there is a certain amount of hype in this area, and we are seeing scientists ``move fast and break things'' all over the place.
 If anything, this \documentname{} represents a call to slow down and think more.
 ML tools are incredibly powerful, if we put them to use in the right places in the natural sciences.
+In particular, the biases we highlight in \secref{sec:bad}
 
-There is a famous ancient quotation about ``the unreasonable effectiveness of mathematics in physics'' \cite{weiner_maybe}.
+There is a famous article by Wigner about ``the unreasonable effectiveness of mathematics in the natural sciences'' \cite{wigner}.
 Mathematics has been pursued (in many cases) for abstract reasons of beauty and technical challenge, but it has ended up creating a language for many of the natural sciences.
@@
 \end{hoggnumerate}
-The data generated this way are shown in the top panels of \figurename~\ref{fig:regression}.
+The data generated this way are shown in the top panels of \figref{fig:regression}.
 The data contain a non-trivial relationship between label $y$ (measured age) $y$ and known parameter $r$ (guiding radius).
-This relationship is shown in \figurename~\ref{fig:regression}.
+This relationship is shown in \figref{fig:regression}.
 It isn't precisely linear because of the removal of data points with latent ages $\eta$ outside the range $0<\eta<14$.
@@
 The model is used to make estimated labels $\hat{y}$ in the full test set.
-The validation and test labels are shown in the bottom panels of \figurename~\ref{fig:regression}.
+The validation and test labels are shown in the bottom panels of \figref{fig:regression}.
 Importantly, \emph{the training, validation, and test samples are drawn from exactly the same process}, with the same distribution in all properties.
@@
 The regression-estimated labels $\hat{y}$ are related to the measured labels $y$ in the held-out validation set with a linear relationship with unit slope and zero intercept.
-The comparison to the validation data (bottom-left panel of \figurename~\ref{fig:regression}) shows that they are not precisely estimated, but there is no evidence for strong bias at the individual-object level.
+The comparison to the validation data (bottom-left panel of \figref{fig:regression}) shows that they are not precisely estimated, but there is no evidence for strong bias at the individual-object level.
 
 The true relationship between label $y$ (measured age) and known parameter $r$ (guiding radius) is estimated in a large generated data set by taking means of $y$ in bins of $r$.
-This relationship is shown in \figurename~\ref{fig:regression} with solid red circles.
-The empirical relationship between label $y$ and known parameter $r$ in the training-set data is shown in \figurename~\ref{fig:regression} with open black squares.
+This relationship is shown in \figref{fig:regression} with solid red circles.
+The empirical relationship between label $y$ and known parameter $r$ in the training-set data is shown in \figref{fig:regression} with open black squares.
 These two relationships are statistically consistent, as expected.
-The empirical relationship in the test-set data between the estimated label $\hat{y}$ and the known parameter $r$ in the test-set data is shown in \figurename~\ref{fig:regression} with black X-shaped symbols.
+The empirical relationship in the test-set data between the estimated label $\hat{y}$ and the known parameter $r$ in the test-set data is shown in \figref{fig:regression} with black X-shaped symbols.
 The test-set relationship is strongly biased, deviating from the true relationship at the edges of the $r$ range.
@@
 Transferring labels to the test set did not help, even though the test set is far larger than the training set.
-The test-set results are very precise (error bars on the X-shaped symbols cannot be visibly shown in \figurename~\ref{fig:regression}), but they are very wrong.
+The test-set results are very precise (error bars on the X-shaped symbols cannot be visibly shown in \figref{fig:regression}), but they are very wrong.
 The training-set results are less precise but unbiased; they are much closer to the truth.
@@
 And yet the population-level inference is biased.
-The averaging of points shown by the X-shaped symbols in \figurename~\ref{fig:regression} is a simple misuse of the regression outputs; that point is trivial.
+The averaging of points shown by the X-shaped symbols in \figref{fig:regression} is a simple misuse of the regression outputs; that point is trivial.
 However, it is a misuse that is occurring repeatedly in astrophysics with catalogs generated by ML regressions HOGG CITE.

### 2024-01-02 paper/main.tex
references, notes to self

@@
 \newcommand{\documentname}{\textsl{Position Paper}}
-\newcommand{\draftstatus}{\textcolor{red}{Draft version 2023-12-31}}
+\newcommand{\draftstatus}{\textcolor{red}{Draft version 2024-01-02}}
 \newcommand{\sectionname}{Section}
@@
 \raggedbottom\sloppy\sloppypar
-
-% Gross
-\newcommand{\apj}{The Astrophysical Journal}
-\newcommand{\araa}{Annual Reviews in Astronomy and Astrophysics}
 
@@
   So far this hope has borne no important new discoveries, but discoveries are not inconceivable.
+  Indeed, there is a very real sense in which the motivations underlying foundation models---finding generally useful latent representations for qualitatively diverse data---are similar to the motivations underlying theoretical physics!
 \end{description}
 This list isn't exhaustive!
-For example, it doesn't mention the use of ML to find summary statistics for the comparison of simulations and data in (say) simulation-based inferences \cite{sbi}.
+For example, it doesn't mention the use of ML to find summary statistics for the comparison of simulations and data in (say) simulation-based inferences \cite{sbi, simbig}.
 It doesn't mention using ML to emulate likelihoods or likelihood ratios \cite{biwei, likelihood_ratio}.
@@
 This is a straightforward point of statistics---this is not news---but it isn't currently informing most of the practice of ML in the natural sciences.
-For example, HOGG CITES.
+For example, HOGG CITES. HOGG: NOTE that we are self-citing here; we are guilty ourselves!
 
@@
 If that is what's required to avoid the bias, then we should just analyze everything with simulations in the first place, and never use the emulators at any stage.
+
+HOGG: Note that you could solve this with fully explainable methods.
+
+HOGG: Note that you can't solve this by doing finitely many tests of the system, by, say, simulating fake data sets.
 
@@
 
-A key place for this kind of causal inference has been in instrument calibration.
+A key place for this kind of causal inference in astrophysics has been in instrument calibration.
 Data taken by an instrument have signals imprinted from the physical (or biological or chemical) processes of interest, and also from the details of the measurement hardware, such as sensitivity and bias variations, and time-dependent distortions of internal mappings.
@@
 ML tools are incredibly powerful, if we put them to use in the right places in the natural sciences.
-In particular, the biases we highlight in \secref{sec:bad}
+In particular, the biases we highlight in \secref{sec:bad} do not have fixes; in our view ML simply cannot be used in those contexts without leading to substantial scientific mistakes.
 
@@
 And which online videos contain kittens?
-In some sense it is remarkable that there is \emph{any} overlap between these technologies and the things we need in scientific domains.
+In some sense it is remarkable that there is \emph{any} overlap between the technologies that solve these problems and the technologies we need in scientific domains.
 
@@
 
-HOGG: TAKE A PARAGRAPH TO EXPLAIN THIS CLAIM: Given the large size of the test set, the precision of the result is high: This bias in the estimated relationship is larger than 30-sigma in the worst data points.
+HOGG: TAKE A PARAGRAPH TO EXPLAIN THIS CLAIM, and make a figure: Given the large size of the test set, the precision of the result is high: This bias in the estimated relationship is larger than 30-sigma in the worst data points.
 
-HOGG: COMMENT OR SHOW that the bias goes down as the information in the features goes up (relative to the information in the training-set distribution (the prior)).
+HOGG: COMMENT AND SHOW that the bias goes down as the information in the features goes up (relative to the information in the training-set distribution (the prior)). Add this to the new figure.
+
+HOGG COMMENT AND SHOW that nothing here depends on the ML model complexity. Add this to the new figure.
 

### 2024-02-01 paper/main.tex
error-free submittable version

@@
+% This document is Copyright 2024 the author(s).
+% It is part of the BadForScience project https://github.com/davidwhogg/BadForScience .
+
 % To do:
 % ------
-% - Change it to be about PHYSICS rather than natural science?
-% - Get feedback from BS, MSV, JH, others
-% - Update appendix with feedback from Vedant Chandra (email).
-% - put in all references
+% - consider taking one non-trivial but simple-to-understand example and thread it through
+% - Take feedback from Margossian (email 2024-01-03).
+% - Take feedback from Rix comments (email 2024-01-04-ish).
+% - Take feedback from KSF comments (email 2024-01-19-ish).
+% - Take feedback from excellent Contardo comments (email 2024-01-28).
+% - Take feedback from Foit email (2024-01-29)
+% - submit to ICML.
+
+% To do after ICML submission:
+% ----------------------------
 % - audit bibliography; make sure arXiv references in the reference list are consistent? Also capitalization!
+% - Update appendix with feedback from Vedant Chandra (email). In particular, modify model complexity and information content.
+% - Should there be another BAD thing, which is about statistics for comparison of simulations and data? This is a great one; consider after ICML submission.
 % - Publish on the arXiv.
-% - set formatting and anonymization switches to go between arXiv and ICML Position Paper
-% - submit to ICML.
-% - Wait for the call from Stockholm.
 
@@
 
+% HOGG CONDITIONAL
+\newif\ifarxiv
+\arxivfalse % set to true for arXiv submission and do something way more complicated for final ICML camera ready.
+
+% NOW it's GO TIME
+\ifarxiv
 \documentclass[10pt]{article}
-\usepackage[utf8]{inputenc}
+\else
+\documentclass{article}
+\fi
+
+% Recommended, but optional, packages for figures and better typesetting:
 \usepackage{microtype}
-\usepackage[letterpaper]{geometry}
-\usepackage{xcolor}
 \usepackage{graphicx}
-\usepackage{url}
+\usepackage{subfigure}
+\usepackage{booktabs} % for professional tables
+
+% hyperref makes hyperlinks in the resulting PDF.
+% If your build breaks (sometimes temporarily if a hyperlink spans a page)
+% please comment out the following usepackage line and replace
+% \usepackage{icml2024} with \usepackage[nohyperref]{icml2024} above.
+\usepackage{hyperref}
+
+% Attempt to make hyperref and algorithmic work together better:
+\newcommand{\theHalgorithm}{\arabic{algorithm}}
+
+% Use the following line for the initial blind version submitted for review:
+\usepackage{icml2024}
+% If accepted, instead use the following line for the camera-ready submission:
+% \usepackage[accepted]{icml2024}
 
@@
 \newcommand{\documentname}{\textsl{Position Paper}}
-\newcommand{\draftstatus}{\textcolor{red}{Draft version 2024-01-02}}
+\newcommand{\draftstatus}{\textcolor{red}{\sffamily\slshape Draft version 2024-01-11}}
 \newcommand{\sectionname}{Section}
@@
 % Hogg typesetting issues that will conflict with ICML
+\ifarxiv
 \setlength{\textwidth}{5.50in}
@@
 \raggedbottom\sloppy\sloppypar
+\else
+\fi
 
@@
 
-\begin{document}\thispagestyle{empty}
+\icmltitlerunning{Is machine learning good or bad for the natural sciences?}
+\begin{document}
+\twocolumn[
+\icmltitle{Is machine learning good or bad for the natural sciences?}
 
-\title{Is machine learning good or bad for the natural sciences?}
-\noindent
-\draftstatus
+\begin{icmlauthorlist}
+\icmlauthor{David W. Hogg}{ccpp,cds,mpia,flatiron}
+\end{icmlauthorlist}
+\icmlaffiliation{ccpp}{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726~Broadway, New~York,~NY 10003, USA}
+\icmlaffiliation{cds}{Center for Data Science, New York University, 60 Fifth Avenue, New York, NY 10011, USA}
+\icmlaffiliation{mpia}{Max-Planck-Institut f{\"u}r Astronomie, K{\"o}nigstuhl 17, D-69117 Heidelberg, Germany}
+\icmlaffiliation{flatiron}{Flatiron Institute, a division of the Simons Foundation, 162 Fifth Avenue, New~York,~NY 10010, USA}
+\icmlcorrespondingauthor{David W. Hogg}{david.hogg@nyu.edu}
 
-\author{David W. Hogg}
-\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726~Broadway, New~York,~NY 10003, USA}
-\affil{Center for Data Science, New York University, 60 Fifth Avenue, New York, NY 10011, USA}
-\affil{Max-Planck-Institut f{\"u}r Astronomie, K{\"o}nigstuhl 17, D-69117 Heidelberg, Germany}
-\affil{Flatiron Institute, a division of the Simons Foundation, 162 Fifth Avenue, New~York,~NY 10010, USA}
+\icmlkeywords{Machine Learning, physics, science, philosophy, bias}
+
+\vskip 0.3in
+]
 
@@
   For example, when an expressive machine-learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can sometimes make the results more trustworthy.
-  We also show that the there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
+  We also show that there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
   For one, when ML models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
   For another, when expressive regressions are used to label datasets, those labels cannot be used in downstream joint analyses without taking on uncontrolled biases.
-  These remarks are intended to apply to all of the natural sciences, but most of the specific examples are in the astrophysics domain.
+  The question of the title is being asked of all of the natural sciences, but the answers here are from the perspective of physics.
 \end{abstract}
@@
 However, when we ask what scientific breakthroughs have been enabled by this influx of new tools and methods, there isn't a long list.
-The success of the AlphaFold projects in protein structure \cite{alphafold} are often raised.
+The success of the \textsl{AlphaFold} projects in protein structure \cite{alphafold} are often raised.
 But these are successes in a very specific challenge-problem setting in which \emph{performance} is valued over \emph{understanding}.
@@
 A ML method is considered successful if it performs well on held-out training data, even if the latent structure of the model is generic and the internals are impossible to interpret.
-In ML, the considerations are all at the level of the data, and we are happy to use models where we have litlle or no understanding of the meanings or values of the latent parameters or weights.
+In ML, the considerations are all at the level of the data, and we are happy to use models where we have little or no understanding of the meanings or values of the latent parameters or weights.
 Another way to ask our question is:
@@
 The main points of this \documentname{} are \textsl{(1)}~that ML has places where it is very valuable in the contemporary practice of science, and \textsl{(2)}~that it also has places where it will create problems for science.
-The answer to the question in the title is ``Both.''
+That said, we can't definitively answer the question in the title of this \documentname{} for all natural-science domains.
+This paper is a call to natural-science communities to think about it, and answer it for themselves.
+From the perspective of the physical sciences, our answer is ``Both.''
 
@@
 Another is a more standard estimator bias that is (possibly enormously) amplified when elements of datasets produced by ML regressions are used in combination or joint analyses.
-These biases can't be easily corrected.
+Neither of these two biases can be easily corrected.
   \item We show that there are many safe places for the use of ML methods in current natural-science practices, and places where (in the contemporary context) the use of ML methods is effectively \emph{required}.
@@
 This definition is broad:
-In addition to methods like convolutional neural networks \cite{cnn}, multi-layer perceptrons \cite{mlp}, and transformers \cite{transformer}, it includes large linear regressions \cite{linearregression}, gaussian processes \cite{gp}, support-vector machines \cite{svm}, principal components analysis \cite{pca}, kernel density estimates \cite{kde}, and even some kinds of multilayer or hierarchical models \cite{multilevel}.
+In addition to methods like convolutional neural networks \cite{cnn}, multi-layer perceptrons, and transformers \cite{transformer}, it includes large linear regressions, gaussian processes \cite{gp}, support-vector machines \cite{svm}, principal components analysis \cite{pca}, kernel density estimates \cite{kde}, and even some kinds of multilayer or hierarchical models \cite{multilevel}; indeed almost any models in any current ML textbooks (eg, \citealt{ml_book1, ml_book2}).
 Feel free to personally take any more restrictive definition.
@@
 \paragraph{Ontologies:}
+The ontology of a domain is (for our purposes) the set of things that exist within that domain.
+What is the ontology of ML?
+
 Unsupervised ML methods deliver descriptions or compressions of data.
@@
 Contemporary optimization schemes are stochastic \cite{stochastic} and most models are non-convex.
-The fact that it is impossible to find the global optimum---and the fact that in most contemporary models that isn't even the goal \cite{not_global}---shows that the latent parameters are not important.
-The use of early stopping \cite{early_stop} and drop-outs \cite{dropout} to stabilize optimization strengthens this point.
+The fact that it is impossible to find the global optimum---and the fact that in most contemporary models that isn't even the goal---shows that the latent parameters are not important.
+The stabilization of optimizations with early stopping \cite{early_stop} and drop-outs \cite{dropout} demonstrate that there isn't even a goal of getting precisely to \emph{any optimum}.
 Any practitioner using a 40-layer network can trivially (and without analysis) switch to a 42-layer, despite the fact that (from a functional point of view), this substantially changes the model capacity.
@@
 \paragraph{Epistemologies:}
+The epistemology of a domain is (for our purposes) the method or standard by which something is judged to be true or correct or known.
+What is the epistemology of ML?
+
 A trained ML model is deemed successful or correct if \emph{it performs well on held-out training data}.
@@
 
-In support of this position we could point to the literature on adversarial attacks (eg, \cite{adversarial1}).
+In support of this position we could point to the literature on adversarial attacks (eg, \citealt{adversarial1}).
 This literature shows us, in dramatic ways, that ML methods are not doing what we (na\"ively) believe that humans or scientists or scientific theories do in comparable circumstances.
@@
 It must explain data in multiple domains, and it must also connect in natural ways to other theories or principles (such as conservation laws and invariances) that are strongly confirmed themselves.
-(It is worthy of note here that a successful theory in the natural sciences is usually a combination of a more fundamental theory of the latents, and a less fundamental or auxiliary observation model, which explains how the latents appear in or affect the observable data \cite{auxiliary}.)
+(It is worthy of note here that a successful theory in the natural sciences is usually a combination of a more fundamental theory of the latents, and a less fundamental or auxiliary observation model, which explains how the latents appear in or affect the observable data.)
 General relativity \cite{gr} was adopted not primarily because it explained anomalous data (although it did explain some); it was adopted because it had a beautiful structure, it resolved conceptual paradoxes in the pre-existing theory of gravity, and it was consistent with emerging ideas of field theory and geometry.
@@
   \item[Outlier detection:] In many of the sciences, discovery of new or previously unknown phenomena or objects can be of great importance.
-  A recent example in astronomy is the discovery of fast radio bursts \cite{grbs}, which started as nuisances in imaging projects and have turned out to be interesting astrophysical objects.
+  A recent example in astronomy is the discovery of fast radio bursts \cite{frbs}, which started as outliers in the data taken for imaging projects and have turned out to be interesting astrophysical objects.
   Since expressive unsupervised ML methods can describe accurately a complex distribution of data, they can also be used to find rare data points that are unlike any elements of the training data.
+  \item[Information questions:] If a machine-learning method is correctly trained to predict labels $y$ from data $x$, and it succeeds (better than random, say), then it shows that data $x$ contains information within it about labels $y$.
+  Regressions thus can be used to make one-sided measurements of quantities of information-theoretic interest.
   \item[Making discoveries?] There is a new hope in the sciences that sufficiently trained or constrained models might lead to new insights about scientific theories or might effectively make fundamental discoveries.
-  There are approaches along the direction of symbolic regression \cite{symbolic1, symbolic2}, and there are approaches along the lines of foundation models \cite{foundation1, foundation2}.
+  There are approaches along the direction of symbolic regression \cite{symbolic1, symbolic2}, and there are approaches along the lines of foundation models \cite{foundation1}.
   So far this hope has borne no important new discoveries, but discoveries are not inconceivable.
@@
 For example, it doesn't mention the use of ML to find summary statistics for the comparison of simulations and data in (say) simulation-based inferences \cite{sbi, simbig}.
-It doesn't mention using ML to emulate likelihoods or likelihood ratios \cite{biwei, likelihood_ratio}.
+It also doesn't mention using ML to emulate likelihoods or likelihood ratios \cite{biwei, likelihood_ratio}.
 The uses of ML in the natural sciences is legion.
@@
 In general, when multiple data point estimates are used jointly, the variance-induced offsets of the estimators average out but the bias-induced offsets remain fixed.
-This is a straightforward point of statistics---this is not news---but it isn't currently informing most of the practice of ML in the natural sciences.
-For example, HOGG CITES. HOGG: NOTE that we are self-citing here; we are guilty ourselves!
+This is a straightforward point of statistics---this is not news---but it isn't currently informing most of the practice of ML in the natural sciences, where many catalogs and public data products are being produced with ML regressions.
+% AFTER ICML submission: For example, HOGG CITES. HOGG: NOTE that we are self-citing here; we are guilty ourselves!
 
@@
 
-HOGG: Note that you could solve this with fully explainable methods.
+It is worthy of note that this problem could be solved, or greatly alleviated, if the inferences were performed with fully explainable methods, from beginning to end.
+Right now there are no hopes for this, although as emulation methods are made to look more and more like physical models (see, eg, \citealt{scalars}) in their internal structure, maybe explainability will become possible.
+This problem cannot be solved in general by making finitely many tests of the emulators.
+First of all, such tests are expensive.
+Second of all, in many cases (cosmology especially), the phase space of random initial conditions is so large that it is impossible to span or cover it in any finite set of tests.
 
-HOGG: Note that you can't solve this by doing finitely many tests of the system, by, say, simulating fake data sets.
+Finally, we note that one objection to this confirmation-bias point is that perhaps it exists just as strongly for any simulation-based inference?
+After all, simulations are expensive, so the full suite of them probably can't be affordably repeated.
+There are two responses to this objection.
+The first is that if the emulators work---that is, if they greatly reduce the cost of the inference---then the emulators have enormously increased the cost difference (or cost factor) between the inference and any re-analysis using physics-based simulations.
+The second is that emulators are not doing precisely what the simulations did.
+One way to see this is that they are subject to adversarial attacks \cite{melchior}.
+Therefore we know that the emulator is not doing precisely what the physics-based simulation is doing.
+That is, verification is not just more expensive, but also more important, when an emulator has been introduced.
 
@@
 An example from astronomy is the upcoming \textsl{Rubin Observatory LSST} project \cite{rubin}, which will generate some $10^5$ events of possible interest for follow-up observing every night \cite{lsst_events}.
-These decisions can be made with very fast classifiers, trained on well-studied training data or simulations \cite{lsst_broker, atlas_trigger}.
+These decisions can be made with very fast classifiers, trained on well-studied training data or simulations (eg, \citealt{lsst_broker}).
 
@@
 
+One scientific-practice comment to make is that the ML literature has a big difference from the natural-science literatures.
+In ML, the publishability of a result is decided---or very strongly influenced---by performance on held-out data.
+It is almost impossible in ML to publish contributions that do not lead to the immediate improvement of some prediction.
+This has a positive side: There are standards!
+But it has many negative sides.
+For example, the path from one good method to another good method might necessarily pass through a valley of less-good methods.
+These non-local-optimization paths aren't available to the ML literature at present.
+Whether good or bad, it is interesting to us that publishing in ML is very different from publishing in the sciences.
+
 Finally, we make a philosophical comment:
-In physics---and in most of the natural sciences we believe---the same phenomena can be explained with many qualitatively different theories \cite{philosphy_of_physics}; mathematical theories are not usually unique.
+In the natural sciences, the same phenomena can usually be explained with many qualitatively different theories; theories are not usually unique (eg, \citealt{hogg, peebles}).
 In the same way (as we noted above), ML models contain massive degeneracies, and even different architectures can lead to near-identical predictions after training on the same training data.
 Does this lead to a philosophical connection between ML and the natural sciences?
-We have argued elsewhere \cite{hogg} that the degeneracies in the natural sciences undermine \emph{realism} or the belief in the literal truth of the latent theories or objects of physics.
+\ifarxiv
+We have
+\else
+It has been
+\fi
+argued elsewhere (op.cit) that the degeneracies in the natural sciences undermine \emph{realism} or the belief in the literal truth of the latent theories or objects of physics.
 There is an argument to be made that maybe the natural sciences could be better off if they moved at least slightly closer to the ML points of view on ontology and epistemology.
@@
 
+\ifarxiv
 \paragraph{Acknowledgements:}
@@
 for conversations over the years related to these arguments, and
-  Vedant Chandra (Harvard)
-for discussions of the bias in regression outputs.
+  Vedant Chandra (Harvard),
+  Contardo,
+  Valentino Foit (NYU),
+  Charles Margossian (Flatiron),
+  Rix,
+  Kate Storey-Fisher (DIPC)
+  Villar, and
+  Kaze Wong (Flatiron)
+for comments on this particular manuscript.
+\else
+\fi
 
+\ifarxiv
 {\raggedright\footnotesize
@@
 }
+\else
+\bibliography{ml_in_astro}
+\bibliographystyle{icml2024}
+\fi
 
@@
 Importantly, \emph{the training, validation, and test samples are drawn from exactly the same process}, with the same distribution in all properties.
-\begin{figure}[p!]
+\begin{figure*}[p!]
 \includegraphics[width=0.49\textwidth]{notebooks/data_examples.png}
@@
 \includegraphics[width=0.49\textwidth]{notebooks/test_data_results.png}
-\caption{\sffamily%
+\caption{
 Visualization of the toy regression.
@@
 The relationship shown by the Xs is very precise but biased far away from the true relationship, unlike the relationship shown by the open squares (measured in the training set alone).\label{fig:regression}}
-\end{figure}
+\end{figure*}
 
@@
 
-HOGG: TAKE A PARAGRAPH TO EXPLAIN THIS CLAIM, and make a figure: Given the large size of the test set, the precision of the result is high: This bias in the estimated relationship is larger than 30-sigma in the worst data points.
+In this example, not only is the test set providing a biased answer (to the question of interest), while the training set is not, it is also the case that the test-set answer is delivered with \emph{very high confidence}.
+Because of the number of objects involved in the test set, the mean-age values are computed at very high precision (the square root of $N$ is large).
+Thus it is not just the case that the test-set answer is biased, it is biased at the 30-sigma level away from the truth.
+The training set gives less precise answers (it is smaller), but the answers are consistent with the truth.
 
-HOGG: COMMENT AND SHOW that the bias goes down as the information in the features goes up (relative to the information in the training-set distribution (the prior)). Add this to the new figure.
+One objection to this example is that the information content in the features about the label is low.
+Indeed! This situation is common in stellar spectroscopy, where we care about subtle abundance and age effects.
+The bias does indeed go down as the information content in the features goes up.
 
-HOGG COMMENT AND SHOW that nothing here depends on the ML model complexity. Add this to the new figure.
+Another objection is that maybe this all depends on model capacity or expressivity.
+The data in this case are made so simply that the regression being used contains sufficient expressivity to model the data, so that issue isn't dominating this problem.
+However, in detail, the variance and bias of the ML regression estimator does depend on the effective model space, and sometimes in problems like this, simpler (less expressive) models can be less biased (see, eg, the Gauss-Markov theorem \citealt{gauss_markov}).
+Full tests of this are out of scope here.
 
@@
 The averaging of points shown by the X-shaped symbols in \figref{fig:regression} is a simple misuse of the regression outputs; that point is trivial.
-However, it is a misuse that is occurring repeatedly in astrophysics with catalogs generated by ML regressions HOGG CITE.
+However, it is a misuse that is occurring repeatedly in astrophysics with catalogs generated by ML regressions.
 
-All code (including the data-generating code) used in this toy are available under an open-source license at \url{https://github.com/davidwhogg/BadForScience}.
+All code (including the data-generating code) used in this toy are available under an open-source license at 
+\ifarxiv
+\url{https://github.com/davidwhogg/BadForScience}.
+\else
+[URL redacted].
+\fi
 

### 2024-05-23 paper/main.tex
nearly camera ready

@@
 % To do:
-% ------
-% - consider taking one non-trivial but simple-to-understand example and thread it through
-% - Take feedback from Margossian (email 2024-01-03).
-% - Take feedback from Rix comments (email 2024-01-04-ish).
-% - Take feedback from KSF comments (email 2024-01-19-ish).
-% - Take feedback from excellent Contardo comments (email 2024-01-28).
-% - Take feedback from Foit email (2024-01-29)
-% - submit to ICML.
-
-% To do after ICML submission:
 % ----------------------------
+% - Find the right place to cite the units paper.
 % - audit bibliography; make sure arXiv references in the reference list are consistent? Also capitalization!
@@
 % ------------
+% - We get an extra page for camera ready (9 pages + bibliography and appendices)
 % - Be very careful about the SUBJECT of sentences. Is it ML that is bad? Is it the practice that is bad? Is it the practitioner that is bad?
 % - Should we use NS for natural science?
-% - How many authors are there? Should we switch to first-person singular?
 % - Audit for flexible, expressive, high in capacity, and so on.
@@
 
-% HOGG CONDITIONAL
-\newif\ifarxiv
-\arxivfalse % set to true for arXiv submission and do something way more complicated for final ICML camera ready.
-
 % NOW it's GO TIME
-\ifarxiv
-\documentclass[10pt]{article}
-\else
 \documentclass{article}
-\fi
 
@@
 
+\usepackage[accepted]{icml2024} %% HACK HACK
+
 % Use the following line for the initial blind version submitted for review:
@@
 \newcommand{\documentname}{\textsl{Position Paper}}
-\newcommand{\draftstatus}{\textcolor{red}{\sffamily\slshape Draft version 2024-01-11}}
+\renewcommand{\paragraph}[1]{\noindent\par\textbf{#1}}
 \newcommand{\sectionname}{Section}
@@
 
-% Hogg typesetting issues that will conflict with ICML
-\ifarxiv
-\setlength{\textwidth}{5.50in}
-\setlength{\oddsidemargin}{-0.25in}
-\setlength{\textheight}{9.25in}
-\setlength{\topmargin}{-0.50in}
-\setlength{\parindent}{1.2\baselineskip} % seriously
-\renewcommand{\title}[1]{\section*{\raggedright #1}}
-\renewcommand{\author}[1]{\medskip\par\noindent\textbf{#1}}
-\newcommand{\affil}[1]{{\footnotesize\par\noindent #1 \par}}
-\renewenvironment{abstract}{\paragraph{Abstract:}}{}
-\pagestyle{myheadings}
-\markboth{foo}{\textcolor{gray}{\textsf{Hogg / Is machine learning good or bad for the natural sciences? \draftstatus}}}
-\linespread{1.08}
-\raggedbottom\sloppy\sloppypar
-\else
-\fi
-
 % Math issues
@@
 \begin{document}
+
 \twocolumn[
-\icmltitle{Is machine learning good or bad for the natural sciences?}
+\icmltitle{\documentname: Is machine learning good or bad for the natural sciences?}
 
 \begin{icmlauthorlist}
-\icmlauthor{David W. Hogg}{ccpp,cds,mpia,flatiron}
+\icmlauthor{David W. Hogg}{ccpp,mpia,flatiron}
+\icmlauthor{Soledad Villar}{flatiron,ams,minds}
 \end{icmlauthorlist}
-\icmlaffiliation{ccpp}{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726~Broadway, New~York,~NY 10003, USA}
-\icmlaffiliation{cds}{Center for Data Science, New York University, 60 Fifth Avenue, New York, NY 10011, USA}
-\icmlaffiliation{mpia}{Max-Planck-Institut f{\"u}r Astronomie, K{\"o}nigstuhl 17, D-69117 Heidelberg, Germany}
-\icmlaffiliation{flatiron}{Flatiron Institute, a division of the Simons Foundation, 162 Fifth Avenue, New~York,~NY 10010, USA}
+\icmlaffiliation{ccpp}{Center for Cosmology and Particle Physics, Department of Physics, New York University, USA}
+\icmlaffiliation{mpia}{Max-Planck-Institut f{\"u}r Astronomie, Heidelberg, Germany}
+\icmlaffiliation{flatiron}{Flatiron Institute, New~York, USA}
+\icmlaffiliation{ams}{Department of Applied Mathematics and Statistics, Johns Hopkins University,  USA}
+\icmlaffiliation{minds}{Mathematical Institute for Data Science, Johns Hopkins University, USA}
 \icmlcorrespondingauthor{David W. Hogg}{david.hogg@nyu.edu}
@@
 \vskip 0.3in
-]
+] % For the arXiv version, move this square bracket below the abstract
 
+\printAffiliationsAndNotice{}
 \begin{abstract}
@@
   These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
-  Here we identify locations for ML in the natural sciences where the ontology and epistemology are valuable.
-  For example, when an expressive machine-learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can sometimes make the results more trustworthy.
+  Here we identify some locations for ML in the natural sciences at which the ontology and epistemology are valuable.
+  For example, when an expressive machine-learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can make the results more trustworthy.
   We also show that there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
   For one, when ML models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
-  For another, when expressive regressions are used to label datasets, those labels cannot be used in downstream joint analyses without taking on uncontrolled biases.
-  The question of the title is being asked of all of the natural sciences, but the answers here are from the perspective of physics.
+  For another, when expressive regressions are used to label datasets, those labels cannot be used in downstream joint or ensemble analyses without taking on uncontrolled biases.
+  The question in the title is being asked of all of the natural sciences; that is, we are calling on the scientific communities to take a step back and consider the role and value of ML in their fields; the (partial) answers we give here come from the particular perspective of physics.
 \end{abstract}
@@
 A significant fraction of all scientific papers in the natural sciences now employ ML in part (or all) of their analyses.
+(We will define ML below in \secref{sec:philosophy}).
 However, when we ask what scientific breakthroughs have been enabled by this influx of new tools and methods, there isn't a long list.
@@
 A ML method is considered successful if it performs well on held-out training data, even if the latent structure of the model is generic and the internals are impossible to interpret.
-In ML, the considerations are all at the level of the data, and we are happy to use models where we have little or no understanding of the meanings or values of the latent parameters or weights.
-Another way to ask our question is:
-\emph{Where, in the natural sciences, can you use a model that you don't understand?}
+In ML, the considerations are almost all at the level of the data, and we are happy to use models in which we have little or no understanding of the meanings or values of the latent parameters or weights.
 
-In natural science the most important contributions and results are all at the level of the \emph{latents}:
+In natural science, on the other hand, the most important contributions and results are all at the level of the \emph{latents}:
 We use data to learn about the latent structure of the world or of the system we are studying.
-In astrophysics, this could be the interior structure of the Sun, or the statistics of planets around other stars, or the map of the dark matter surrounding the Milky Way.
+In astrophysics, this could be the interior structure of the Sun, or the processes that form planets around other stars, or the map of the dark matter surrounding the Milky Way.
 The things we care about are almost never directly observable; they are parameters (or hyper-parameters) of a physical (or chemical or biological) model that predicts the observables.
+Often the thing we care about is the model itself.
 
@@
 The discovery would not have been seen as important if Hubble and Humason had instead announced that they had trained a deep multilayer perceptron that could predict the Doppler shifts of held-out extragalactic nebulae.
+
+For another example, consider the discovery (by Kepler) that the paths of the planets are ellipses, with the Sun at one focus \cite{kepler}.
+This discovery led to extremely precise predictions for data.
+It was critical to this discovery that the data be well explained by the theory.
+But that wasn’t the primary consideration that made the scientific community prefer the Keplerian model.
+After all, the Ptolemaic model preceding Kepler made equally accurate predictions of held-out data.
+Kepler’s model was preferred because it fit in with other ideas being developed at the same time, most notably heliocentrism \cite{copernicus} and Newtonian gravity \cite{newton}.
 
@@
   \item We deliver a description of the fundamental \emph{ontology} and \emph{epistemology} of machine learning, and contrast these with the ontologies and epistemologies of the natural sciences.
-  \item We elucidate two important and strong statistical biases that are being introduced to the natural sciences by some of the uses of ML in those disciplines.
+  \item We elucidate two important and strong statistical biases that are being introduced to the natural sciences by some uses of ML.
 One is a confirmation bias that arises when simulations are replaced or augmented by emulators.
-Another is a more standard estimator bias that is (possibly enormously) amplified when elements of datasets produced by ML regressions are used in combination or joint analyses.
+Another is a more standard estimator bias that is (possibly enormously) amplified when elements of datasets produced by ML regressions are used in combination or joint or ensemble analyses.
 Neither of these two biases can be easily corrected.
@@
 
+As broad as this definition is, it does not include standard model fitting or parameter estimation.
+These tasks are usually performed in the context of a well-specified model, where the model does not qualitatively change as more data arrive, and the quantitative improvements to the parameter estimates only scale as something like the square root of the size of the data set.
+Standard model fitting and parameter estimation are not ML in our sense.
+
 \paragraph{What is natural science?}
@@
 We are excluding these kinds of questions in part because it is obvious that ML methods \emph{can} serve important purposes in such problems.
-That will re-emerge as a theme, below.
+
+In addition, we recognize that there are quite a few different attitudes that a scientist can take towards the concept of ``understanding''.
+Does a program that can predict and control a robot arm accurately ``understand'' the arm?
+Or does a system that can predict protein foldings ``understand'' protein folding?
+We are using a restricted definition of ``understanding'' that excludes these operational successes; mostly in the natural sciences we are looking for semantic rather than operational understandings of systems.
+But of course ML often \emph{can} provide extremely operationally capable systems.
 
@@
 
-Unsupervised ML methods deliver descriptions or compressions of data.
+Unsupervised and self-supervised ML methods deliver representations, descriptions, or compressions of data.
 Supervised ML methods find relationships between data (features) and data (labels).
 In both cases, the methods are judged on their capability to accurately describe the data.
-They are not judged on the details of their latent structure.
+They are (often) not judged on the details of their latent structure.
 In a very important sense, the ML ontology is that \emph{only the data exist}.
@@
 Deep-learning models have enormously large internal degeneracies (combinatorial degeneracies even).
-These degeneracies are not seen as a problem, since all they do is make the latent weights less interpretable.
+These degeneracies are not seen as a problem, since all they do is make the latent weights less interpretable without hurting performance \cite{belkin2019reconciling, bartlett2020benign}.
 Contemporary optimization schemes are stochastic \cite{stochastic} and most models are non-convex.
 The fact that it is impossible to find the global optimum---and the fact that in most contemporary models that isn't even the goal---shows that the latent parameters are not important.
-The stabilization of optimizations with early stopping \cite{early_stop} and drop-outs \cite{dropout} demonstrate that there isn't even a goal of getting precisely to \emph{any optimum}.
-Any practitioner using a 40-layer network can trivially (and without analysis) switch to a 42-layer, despite the fact that (from a functional point of view), this substantially changes the model capacity.
+The regularization of trained models with early stopping \cite{early_stop} and dropouts \cite{dropout} demonstrate that there isn't even a goal of getting precisely to \emph{any optimum}.
+Any practitioner using a 23-layer network can trivially (and without analysis) switch to a 42-layer, despite the fact that (from a functional point of view), this might substantially change the model capacity or expressivity.
 
 In ML, models are deemed stably optimized and useful not if the latent space---the joint values of all the weights and biases---is stable, but rather if the predictions for held-out data are stable.
-Thus it is the data that exist in ML contexts, not the latent parameters.
+This is used as an indicator that the \emph{learned function} (at least in the training, validation, and test set) is stable, even though there can be large subspaces of very different, unidentifiable latent parameters.
+Since the learned function is a model for the data, or something that predicts the data, it is the data that exist in ML contexts, not the latent parameters.
 
-In contrast, the ontologies of the natural sciences contain far more things than just the data.
+In ML there is sometimes additionally a notion of ``ground truth''; does this concept extend the ontology of ML?
+Some ML methods assume that the training (and validation) data constitute ground truth (the labels are correct and have knowledge of the world, say).
+In this case, ``ground truth'' is just another name for the training data.
+Other methods assume that the labels might be noisy or adversarial, and that the method discovers true or more true labels.
+This does perhaps extend the ontology, but only slightly, since the ground truth is in the same space as the data (and nearby to it, presumably).
+Another notion that sometimes arises in ML is the ``data-generating process'', especially in contexts of emulation.
+We don't see this as different from the data themselves, but technically this does extend the ontology marginally.
+
+In contrast to ML, the ontologies of the natural sciences contain far more things than just the data.
 In physics, for example, 
@@
 These entities are judged to exist in part because they are involved in the latent structure of the successful theories; almost none of them are direct observables.
-The most important discoveries in the natural sciences are discoveries of latent structure: Natural selection as an explanation for the differentiated properties of species \cite{natural_selection}, for example, or the elemental composition of the Sun \cite{sun_composition}, or the quarks and gluons that make up the proton \cite{proton_substructure}.
+The most important discoveries in the natural sciences are discoveries of latent structure: Natural selection as an explanation for the differentiated properties of species \cite{natural_selection}, for example, or the elemental composition of the Sun (see \citealt{sun_composition}), or the quarks and gluons that make up the proton \cite{proton_substructure}.
+
+The mismatch of ontology (and epistemology, below) between ML and natural sciences is not unique to ML;
+many tools used in the natural sciences have limited ontologies.
+Perhaps the most extreme case is that of mathematical tools, which are based (in some sense) only on a limited set of axioms.
+Mathematics has a very limited ontology.
+ML is useful in science in a way that is analogous to other methodological tools; the point made here is that the particular and specific use of each tool ought to be matched to its ontology and epistemology.
 
@@
 In these ML subfields, in some cases, attempts are made to understand the internal structure or the behavior of the ML model.
-There are not enormous successes here, but if there were, these might be subfields in which the epistemology of ML gets more complex; maybe in the future some ML models will have to perform well on data \emph{and} be explainable in some important sense.
-At present this isn't part of any ML epistemology.
+There is not consensus here, but if there were, it might create subfields in which the epistemology of ML gets more complex; maybe in the future some ML models will have to perform well on data \emph{and} be explainable in some important sense.
+That would have implications for the epistemology, and probably also the ontology of ML.
 
@@
 The detailed epistemological framework depends on the natural-science field, and even on the practitioner within the field.
-However, in all cases, it is more demainding:
-A theory or explanation has to do much more than just explain the data in order to be accepted as true.
-In physics for example, a model---which, as we note, is almost always a model of latent structure---is judged to be good or strongly confirmed not just if it explains observed data.
-It must explain data in multiple domains, and it must also connect in natural ways to other theories or principles (such as conservation laws and invariances) that are strongly confirmed themselves.
+However, in all cases, it is more demanding:
+A theory or explanation has to do much more than just explain the data in order to be widely accepted as true.
+In physics for example, a model---which, as we note, is almost always a model of latent structure---is judged to be good or strongly confirmed not only if it explains observed data.
+It ought to explain data in multiple domains, and it must connect in natural ways to other theories or principles (such as conservation laws and invariances) that are strongly confirmed themselves.
 (It is worthy of note here that a successful theory in the natural sciences is usually a combination of a more fundamental theory of the latents, and a less fundamental or auxiliary observation model, which explains how the latents appear in or affect the observable data.)
-General relativity \cite{gr} was adopted not primarily because it explained anomalous data (although it did explain some); it was adopted because it had a beautiful structure, it resolved conceptual paradoxes in the pre-existing theory of gravity, and it was consistent with emerging ideas of field theory and geometry.
+General relativity \cite{gr} was widely accepted by the community (and very quickly; \citealt{peebles}) not primarily because it explained anomalous data (although it did explain some); it was adopted because, in addition to explaining (a tiny bit of new) data, it also had good structure, it resolved conceptual paradoxes in the pre-existing theory of gravity, and it was consistent with emerging ideas of field theory and geometry.
 
@@
 In most cases, the ML is required on the engineering or execution side of projects that are large-scale in some sense.
-\begin{description}
-  \item[Label transfer:] Sometimes a project has informative data on a very large number of objects, but precise labels for only a few, maybe obtained through very careful analysis or external data.
+The descriptions here are non-exhaustive, they have a strong physics bias to them, and they do not address the many technical nuisances that make this discussion approximate and less specific than we would like.
+Our hope is that this \documentname{} is seen as a conversation starter and not as providing a complete answer to any of the posed questions.
+
+\paragraph{Label transfer:}
+Sometimes a project has informative data on a very large number of objects, but precise labels for only a few, maybe obtained through very careful analysis or external data.
 In this case, if it is extremely expensive to label more objects, a regression can be trained on the few labeled data points and then the trained regression used to label all the rest.
-  Below (\secref{sec:bad}) we argue that this is not an advisable use of ML in the natural sciences if the expectation is that the regression-generated labels are going to be used in downstream scientific projects.
-  \item[Classification:] This is the same use case as label transfer, but in the specific case in which the labels are drawn from a finite (small, actually) set of discrete values.
-  It is subject to the same kinds of biases described below (\secref{sec:bad}).
-  \item[Speeding up decisions:] Many scientific projects must make decisions very fast in real time.
+Below in \secref{sec:bad} we argue that this is not an advisable use of ML in the natural sciences if the expectation is that the regression-generated labels are going to be used certain straightforward ways in downstream scientific projects.
+
+\paragraph{Classification:}
+This is the same use case as label transfer, but in the specific case in which the labels are drawn from a finite (small, actually) set of discrete values.
+It is subject to the same kinds of issues as label transfer.
+
+\paragraph{Speeding up decisions:}
+Many scientific projects must make decisions very fast in real time.
 The most extreme examples of this are in particle physics, where detectors (such as the \textsl{Atlas} Experiment \cite{atlas} at the \textsl{Large Hadron Collider}) must decide whether to trigger a data-saving event in a tiny fraction of second.
@@
 It is critical, if ML methods are used for real-time operations, that the methods and all their latent parameters (weights) be preserved for analysis, operations simulations, and conterfactual exercises, often performed long after the data are taken.
-  \item[Speeding up simulations:] In much of the natural sciences, the theory in play is a computation or simulation or digital twin.
+% HOGG: KSF NOTES that this last point could be expanded upon here or in the discussion.
+
+\paragraph{Speeding up simulations:}
+In much of the natural sciences, the theory in play is a computation or simulation or digital twin.
 These simulations tend to be very computationally expensive, since they often span large ranges of spatial or time scales.
 ML regressions can be trained to emulate the simulations, or patch up low-resolution simulations to higher resolutions.
-  Below (\secref{sec:bad}) we argue that the introduction of ML emulators can introduce an unwanted, strong confirmation bias.
-  \item[Modeling nuisances:] In most natural-science domains, the quantities of interest are not directly observable, but rather the model or prediction for the data is a combination of the theory of relevance plus auxilliary theories of foregrounds, backgrounds, instrumentation, and noise sources.
-  When the goal is for these nuisances to be effectively modeled but not necessarily understood in detail, ML approaches can be effective and (as we argue in \secref{sec:good}) even conservative.
-  \item[Outlier detection:] In many of the sciences, discovery of new or previously unknown phenomena or objects can be of great importance.
+Below in \secref{sec:bad} we argue that the introduction of ML emulators can introduce an unwanted, strong confirmation bias.
+
+\paragraph{Modeling nuisances:}
+In most natural-science domains, the quantities of interest are not directly observable, but rather the model or prediction for the data is a combination of the theory of relevance plus auxilliary theories of foregrounds, backgrounds, instrumentation, and noise sources.
+When the goal is for these nuisances to be effectively modeled but not necessarily understood in detail, ML approaches can be effective and, as we argue in \secref{sec:good}, even conservative.
+
+\paragraph{Propose objects, materials, or interactions for follow-up:}
+In most scientific settings, discoveries are valid if they can be experimentally verified, regardless of the process used to produce the scientific hypothesis.
+That is, hypotheses can be generated by creative processes that don't need to be understood, provided that the hypotheses so generated can be tested by experimental methods that \emph{are} well understood.
+ML-based generative models can be used to conjecture the existence of objects or materials with certain properties, that can be later verified in lab assays.
+One example of this is ML-based drug discovery \cite{kang2018conditional}.
+
+\paragraph{Outlier detection:}
+In many of the sciences, discovery of new or previously unknown phenomena or objects can be of great importance.
 A recent example in astronomy is the discovery of fast radio bursts \cite{frbs}, which started as outliers in the data taken for imaging projects and have turned out to be interesting astrophysical objects.
-  Since expressive unsupervised ML methods can describe accurately a complex distribution of data, they can also be used to find rare data points that are unlike any elements of the training data.
-  \item[Information questions:] If a machine-learning method is correctly trained to predict labels $y$ from data $x$, and it succeeds (better than random, say), then it shows that data $x$ contains information within it about labels $y$.
+Since expressive unsupervised ML methods can (with some caveats) describe accurately a complex distribution of data, they can also be used to identify (some) rare data points that are unlike any elements of the training data.
+
+\paragraph{Information questions:}
+If a machine-learning method is correctly trained to predict labels $y$ from data $x$, and it succeeds (better than random, say), then it shows that data $x$ contains information within it about labels $y$.
 Regressions thus can be used to make one-sided measurements of quantities of information-theoretic interest.
-  \item[Making discoveries?] There is a new hope in the sciences that sufficiently trained or constrained models might lead to new insights about scientific theories or might effectively make fundamental discoveries.
+
+\paragraph{Making discoveries?}
+There is a new hope in the sciences that sufficiently trained or constrained models might lead to new insights about scientific theories or might effectively make fundamental discoveries.
 There are approaches along the direction of symbolic regression \cite{symbolic1, symbolic2}, and there are approaches along the lines of foundation models \cite{foundation1}.
 So far this hope has borne no important new discoveries, but discoveries are not inconceivable.
-  Indeed, there is a very real sense in which the motivations underlying foundation models---finding generally useful latent representations for qualitatively diverse data---are similar to the motivations underlying theoretical physics!
-\end{description}
+Indeed, there is a very real sense in which the motivations underlying foundation models---finding generally useful latent representations for qualitatively diverse data---are similar to the motivations underlying theoretical physics.
+
+\medskip
 This list isn't exhaustive!
@@
 It also doesn't mention using ML to emulate likelihoods or likelihood ratios \cite{biwei, likelihood_ratio}.
-The uses of ML in the natural sciences is legion.
+The uses of ML in the natural sciences are legion \cite{wang2023scientific,zhang2023artificial}.
 
@@
 This is good; the individual-data-point label estimates from the regression are the lowest ``risk'' (in the statistical sense) when they use non-zero bias in the bias--variance trade-off.
-However, the biases that are weak or manageable on an individual data-point basis become strong or unacceptable biases when outputs of regressions are used jointly or in combination to measure a population or sub-population property.
+However, the biases that are weak or manageable on an individual data-point basis become strong or unacceptable biases when outputs of regressions are used jointly or in combination to measure a population or sub-population or ensemble property.
 
@@
 It is worst when the regression is used to label a large data set or catalog or sample of data,
-and when elements of that data set are used in populations or joint analyses.
+and when elements of that data set are used in populations or joint or ensemble analyses.
 In general, when multiple data point estimates are used jointly, the variance-induced offsets of the estimators average out but the bias-induced offsets remain fixed.
@@
 
-In principle, there are fixes for this problem that involve de-biasing the estimates.
+In principle, there might be fixes for this problem that involve de-biasing the estimates.
 This is not possible in general, because there is not usually enough validation data to accurately assess the bias.
@@
 Likelihoods can produce unbiased point estimates.
-Likelihood-based estimators are not usually the lowest-risk estimators, but they can be low- or zero-bias estimates.
+Likelihood-based estimators are not usually the lowest-risk estimators, but they can be low- or zero-bias estimators.
 The outputs of regressions are more like posterior-based estimators, affected by the implicit prior set by the training set.
@@
 
-Finally we note---and demonstrate in \appendixname~\ref{app:toy}---that none of these problems arise from out-of-sample generalization or distribution-shift problems.
-Distribution shifts just tend to make these kinds of problems worse.
+Importantly, there is nothing unique to ML about low-risk estimators being biased, or that the use of biased estimators in populations analyses is problematic.
+These are standard properties of estimators.
+The unique thing to ML is that standard ML regressions almost always produce biased estimators.
+Given the objective functions (cost functions) in use in ML regressions, this is a feature of ML, not a bug.
+
+Finally we note---and demonstrate in \appendixname~\ref{app:toy}---that none of these problems of bias arise from out-of-sample generalization or distribution-shift problems, or even model mis-specification.
+Distribution shifts (where the test data are drawn from a different distribution than the training data) and model mis-specification (where the model doesn't have the expressivity to explain the data) just tend to make these biases even worse.
 
@@
 
-Now imagine that two cosmology experiments proceed to do enormous emulator-based inferences on large data sets.
-Imagine that these inferences are so large that it would have been exceedingly expensive to have done them with the original first-principles simulations instead of the emulators.
+Now imagine that two cosmology experiments proceed to do enormous inferences on large data sets, each of which requires a very large number of simulations---so many that a fast, trained ML emulator has to be used.
+Imagine that both of these inferences would have been exceedingly expensive to have been done with the original first-principles simulations instead of the emulators.
 Experiment A finds cosmological parameters very much in line with our expectations going in.
@@
 
-Since the money involved is astronomical, there is no simple fix for this unless we plan to re-analyze all experiments using first-principles simulations, no matter what they find.
-If that is what's required to avoid the bias, then we should just analyze everything with simulations in the first place, and never use the emulators at any stage.
+Since the resources required are substantial (or even enormous), there is no simple fix for this unless we plan to re-analyze all experiments using first-principles simulations, no matter what they find.
+If that is what's required to avoid the bias, then we should just analyze everything with the full simulations in the first place, and never use the emulators at any stage.
 
 It is worthy of note that this problem could be solved, or greatly alleviated, if the inferences were performed with fully explainable methods, from beginning to end.
-Right now there are no hopes for this, although as emulation methods are made to look more and more like physical models (see, eg, \citealt{scalars}) in their internal structure, maybe explainability will become possible.
+Right now there are no hopes for this, although as emulation methods are made to look more and more like physical models (see, eg, \citealt{scalars, villar2023dimensionless}) in their internal structure, maybe explainability will become possible.
 This problem cannot be solved in general by making finitely many tests of the emulators.
 First of all, such tests are expensive.
-Second of all, in many cases (cosmology especially), the phase space of random initial conditions is so large that it is impossible to span or cover it in any finite set of tests.
+Second of all, in many cases (and in cosmology especially), the phase space of random initial conditions is so large that it is impossible to span or cover it in any finite set of tests.
+Testing is good! But complete testing is impossible.
 
@@
 \section{When is machine learning good for natural science?}\label{sec:good}
-In \secref{sec:intro} we re-phrased the question in the title as
-``Where, in the natural sciences, can you use a model you don't understand?''
-The answer is: You can use it for the parts of your scientific project or experiment or analysis that \emph{you don't need to understand}.
-We have colleagues who believe that there is no such part.
-We disagree: For example in astronomy, we use charge-coupled devices to make extremely precise measurements of stars, without knowing in detail the solid-state physics that makes them efficient, linear detectors of photons.
-There are many parts of a scientific project that must work well, but which do not need to be carefully controlled or understood (in advance, anyway).
+Where, in the natural sciences, can you use models in which you don't fully understand the latent parameters and structures?
+The answer is: You can use them for the parts of your scientific project or experiment or analysis that \emph{you don't need to understand}.
+We have colleagues who believe that there is no such part; some scientists believe that you must understand deeply every part of your scientific toolkit.
+We disagree: For an example in physics, we use infrared HgCdTe detectors to make extremely precise measurements of intensity fields, without understanding in detail the solid-state physics that makes these devices efficient, linear detectors of photons.
+There are many parts of a scientific project that must work well, but which do not need to be carefully controlled or understood.
+They have to be testable and tested, but they don't have to be fully understood.
 
 Here we give a few examples.
-These are not intended to be exhaustive; they are just examples of places within a scientific project ML can be used to the benefit---and not detriment---of the natural-science goals.
+These are not intended to be exhaustive; they are just examples of places within a scientific project at which ML can be used to the benefit---and not detriment---of the natural-science goals.
 
@@
 Is it unsafe to operate a project using ML decision-making systems?
-Even if these systems do not have to be extremely well understood in real time, but they do need to be \emph{versioned} and preserved for study later.
+Even if these systems do not have to be extremely well understood in real time, they do need to be \emph{versioned} and preserved for study later.
 Statistical projects making use of the data from systems like this will need to understand the statistical biases created by the ML selection procedures.
@@
 Many discoveries in astrophysics are discoveries of new kinds of objects, originally found as data outliers \cite{quasars, voorwerp}.
-So far, ML has not been involved in any big discoveries in astronomy, but we expect this to be a productive avenue for ML in the natural sciences and astronomy in particular.
+So far, ML has not been involved in any big discoveries in astronomy, but we expect this to be a productive avenue for ML in the natural sciences.
 
@@
 The simplest part of this observation model is the model of the backgrounds, foregrounds, and confounders.
-For example, any maps of the cosmic microwave background (by the ESA \textsl{Planck} spacecraft \cite{planck_maps}, for example) will have emission from the intervening Milky Way and from radio galaxies, adding to, and distorting the maps.
+For example, any maps of the cosmic microwave background (by the ESA \textsl{Planck} spacecraft \cite{planck_maps}, for example) will have emission from the intervening Milky Way and from radio galaxies, which both add to and distort the maps.
 When the goal is to understand the cosmic microwave background, the investigator often doesn't need to understand these foregrounds \emph{physically}; an \emph{effective model} is sufficient.
 And indeed, all the bleeding-edge methods for this problem currently are ML methods \cite{cmb_foregrounds}.
-It doesn't matter to the questions of physical cosmology what those foregrounds \emph{mean}; and indeed with ML-based foreground removal, cosmic microwave background experiments have delivered the most precise measurements of the parameters of the cosmological model with sub-percent accuracy \cite{planck_parameters}.
+It doesn't matter to the questions of physical cosmology what those foregrounds \emph{mean}; and indeed with ML-based foreground removal, cosmic microwave background experiments have delivered measurements of the parameters of the cosmological model with sub-percent precision \cite{planck_parameters}.
 For foregrounds and backgrounds, understanding is not always necessary.
@@
 The more expressive the model used to model the confounders, the stronger the conclusions that can be made about the causation \cite{causal_inference}.
-That is, the huge capacity of ML methods can made a causal claim \emph{very conservative}, and hence very robust.
+That is, the huge capacity of ML methods can make a causal claim \emph{very conservative}, and hence very robust.
 One way to see this is that if the confounder model has been given every chance it possibly can have to model the effect of interest, and it cannot, then it is a stronger conclusion that the effect is caused by the cause of interest.
@@
 We hope we have clearly answered the question in the title with the response ``Both!''
-Machine learning (ML) is a reality for the natural sciences:
+Our more detailed answers come from a perspective of physics, but we expect the answers to be similar in many of the natural sciences.
+Machine learning (ML) is a reality for science:
 We need it for speed and scale in many contexts in many contemporary scientific projects, where instrument bandwidths, data volumes, assay parallelization, experimental automation, and scientific ambitions are all growing.
@@
 
-That said, there is a certain amount of hype in this area, and we are seeing scientists ``move fast and break things'' all over the place.
+Of course there is a certain amount of hype in this area, and we are seeing scientists ``move fast and break things'' all over the place.
 If anything, this \documentname{} represents a call to slow down and think more.
@@
 
+In some ways it is surprising that ML has been useful in the natural sciences, since much of the development of ML methods has been in commercial decision-making applications, such as delivering content to users.
+These commercial applications often show a reasonable match between the implicit philosophy of ML and the goals of the commercial ML systems (because, say, serving data to new users is like predicting held-out test data).
+That said, there are connections between commercial applications and scientific applications.
+For example, on the commerce side, there can be important and manageable (and even interpretable) distribution shifts, and, on the scientific side, there are contexts in which the system really will see more of the same (for example, trigger systems in particle-physics colliders).
+
 There is a famous article by Wigner about ``the unreasonable effectiveness of mathematics in the natural sciences'' \cite{wigner}.
@@
 Similarly, we could coin a phrase about \emph{the unreasonable effectiveness of machine learning in the natural sciences}.
-It's unreasonable because ML methods were developed, engineered, and optimized to solve problems for commerce.
+It's unreasonable because current ML methods are typically engineered and optimized to solve problems in commercial applications.
 The most remarkable ML systems have been driven by critical industrial questions such as:
@@
 
-One scientific-practice comment to make is that the ML literature has a big difference from the natural-science literatures.
-In ML, the publishability of a result is decided---or very strongly influenced---by performance on held-out data.
+One scientific-practice comment to make is that the ML literature is very different from the natural-science literatures.
+In ML, the publishability of a result is decided---or very strongly influenced---by its demonstrated performance on held-out data.
 It is almost impossible in ML to publish contributions that do not lead to the immediate improvement of some prediction.
@@
 
+Given everything we have said here, what are the ways forward for the natural sciences?
+ML is critical in many projects; how do we make sure it is doing what we want it to do?
+Our view is that the promising directions include imposing physically motivated mathematical structure into the design or regularization of the ML models.
+This includes symmetries, conservation laws, and constraints \cite{bronstein2021geometric, villar2023towards}, but also ODE/PDE solvers \cite{karniadakis2021physics}, dynamic programming modules \cite{xu2020can}, and other established mathematical tools.
+The research direction that looks towards explainability, interpretability \cite{interpretable}, and trustworthiness \cite{kearns2019ethical} is also fundamental to the development of ML methods for the physical sciences.
+These approaches can increase natural scientists’ confidence in ML tools, and also perhaps expand their ontologies and epistemologies.
+
 Finally, we make a philosophical comment:
@@
 Does this lead to a philosophical connection between ML and the natural sciences?
-\ifarxiv
-We have
-\else
-It has been
-\fi
-argued elsewhere (op.cit) that the degeneracies in the natural sciences undermine \emph{realism} or the belief in the literal truth of the latent theories or objects of physics.
+We have argued elsewhere (op.cit) that the degeneracies in the natural sciences undermine \emph{realism} or the belief in the literal truth of the latent theories or objects of physics.
 There is an argument to be made that maybe the natural sciences could be better off if they moved at least slightly closer to the ML points of view on ontology and epistemology.
@@
 
-\ifarxiv
+\paragraph{Broader societal and ethical impacts:}
+This \documentname{} is directly aimed at producing certain kinds of broader impacts, in the execution of natural-science investigations and projects.
+ML has many roles in the natural sciences.
+This work is intended to start a discussion which, in time, will support the implementation of ML in good roles in scientific projects, and criticize the implementation of ML in less good roles in scientific projects.
+It is also intended to help practitioners understand and respond to biases introduced by the use of ML in those less-good roles.
+The goal of this work is to improve the accuracy and success of the natural sciences.
+However, if the recommendations in this \documentname{} are adopted by scientific communities, the costs of some scientific projects may increase.
+
 \paragraph{Acknowledgements:}
@@
   Hans-Walter Rix (MPIA),
-  Sam Roweis (deceased),
-  Bernhard Sch\"olkopf (MPI-IS), and
-  Soledad Villar (JHU)
+  Sam Roweis (deceased), and
+  Bernhard Sch\"olkopf (MPI-IS)
 for conversations over the years related to these arguments, and
@@
   Valentino Foit (NYU),
+  Alex Jiang (CUNY),
   Charles Margossian (Flatiron),
   Rix,
-  Kate Storey-Fisher (DIPC)
-  Villar, and
-  Kaze Wong (Flatiron)
-for comments on this particular manuscript.
-\else
-\fi
+  Kate Storey-Fisher (DIPC),
+  Kaze Wong (Flatiron), and
+  our four anonymous ICML reviewers
+for valuable comments on this particular manuscript.
+SV was partially supported by ONR N00014-22-1-2126, NSF CCF 2212457, the NSF–Simons Research Collaboration
+on the Mathematical and Scientific Foundations of Deep Learning (MoDL) (NSF DMS 2031985), and NSF
+CAREER 2339682.
+The Flatiron Institute is a division of the Simons Foundation.
 
-\ifarxiv
-{\raggedright\footnotesize
-\bibliography{ml_in_astro}
-\bibliographystyle{plain}
-}
-\else
 \bibliography{ml_in_astro}
 \bibliographystyle{icml2024}
-\fi
 
@@
 \section{Population-level biases in regression outputs}\label{app:toy}
-A machine-learning (ML) regression learns the best (that it can find) value of a list or vector $\hat{\theta}$ of parameters (weights and thresholds, for example) of an expressive function $f(x;\theta)$ that takes as input a feature list or vector $x$ and outputs a label estimate $y$.
-The parameters $\hat{\theta}$ are learned from a training set of labeled data, which is a set of $(x, y)$ pairs.
+A standard machine-learning (ML) regression learns a function $f(x;\theta)$ that takes as input a feature list or vector $x$ and outputs a label estimate $y$.
+In detail, the method learns the value of a list or vector $\hat{\theta}$ of parameters (weights and thresholds, for example) of the function that leads to the best predictions of the labels in some training set; the training set is a set of labeled data, or a set of $(x, y)$ pairs.
 The learned function $f(x;\hat{\theta})$ delivers a predicted or estimated label $\hat{y}$ for any data point with a complete set of features $x$.
@@
     \item A $M$-dimensional feature vector $x$ called ``the data'' is generated from a Gaussian with mean $\tilde{\xi}$ and variance $0.0025$.
+(These feature vectors have been designed to look a little like stellar spectra.)
     \item A label $y$ called ``measured age'' is generated from a Gaussian with mean $\eta$ and variance 1.
@@
 Black X-shaped symbols show the mean relationship between $\hat{y}$ and $r$.
-The relationship shown by the Xs is very precise but biased far away from the true relationship, unlike the relationship shown by the open squares (measured in the training set alone).\label{fig:regression}}
+The relationship shown by the Xs is very precise (error bars are much smaller than the symbols; see the text) but biased far away from the true relationship, unlike the relationship shown by the open squares (measured in the training set alone).\label{fig:regression}}
 \end{figure*}
@@
 The data in this case are made so simply that the regression being used contains sufficient expressivity to model the data, so that issue isn't dominating this problem.
-However, in detail, the variance and bias of the ML regression estimator does depend on the effective model space, and sometimes in problems like this, simpler (less expressive) models can be less biased (see, eg, the Gauss-Markov theorem \citealt{gauss_markov}).
+However, in detail, the variance and bias of the ML regression estimator does depend on the effective model space, and sometimes in problems like this, simpler (less expressive) models can be less biased (see, eg, the Gauss-Markov theorem).
 Full tests of this are out of scope here.
@@
 The investigator should \emph{just use the training set}.
-Transferring labels to the test set did not help, even though the test set is far larger than the training set.
+Transferring labels to the test set did not help---indeed it actually hurt---even though the test set is far larger than the training set, and drawn from the same distribution.
 The test-set results are very precise (error bars on the X-shaped symbols cannot be visibly shown in \figref{fig:regression}), but they are very wrong.
@@
 And yet the population-level inference is biased.
-The averaging of points shown by the X-shaped symbols in \figref{fig:regression} is a simple misuse of the regression outputs; that point is trivial.
+The averaging of points shown by the X-shaped symbols in \figref{fig:regression} is a simple misuse of the regression outputs:
+The averaging of biased estimates amplifies the statistical significance of the estimator's biases.
 However, it is a misuse that is occurring repeatedly in astrophysics with catalogs generated by ML regressions.
 
+Another way to say it is:
+If you expect your downstream users to perform populations inferences with your outputs, you want your outputs to be unbiased.
+The methods you choose must be informed by the subjective needs and expectations of your users.
+ML regressions aren't appropriate to the use cases of many downstream users.
+
 All code (including the data-generating code) used in this toy are available under an open-source license at 
-\ifarxiv
 \url{https://github.com/davidwhogg/BadForScience}.
-\else
-[URL redacted].
-\fi
 

### 2024-05-24 paper/main.tex
camera-ready, according to hogg

@@
 
-\icmltitlerunning{Is machine learning good or bad for the natural sciences?}
+\icmltitlerunning{Hogg \& Villar: Is machine learning good or bad for the natural sciences?}
 \begin{document}
@@
 
-For another example, consider the discovery (by Kepler) that the paths of the planets are ellipses, with the Sun at one focus \cite{kepler}.
+For another example, consider the discovery that the paths of the planets are ellipses, with the Sun at one focus \cite{kepler}.
 This discovery led to extremely precise predictions for data.
 It was critical to this discovery that the data be well explained by the theory.
-But that wasn’t the primary consideration that made the scientific community prefer the Keplerian model.
+But that was not the primary consideration that made the nascent scientific community prefer the Keplerian model.
 After all, the Ptolemaic model preceding Kepler made equally accurate predictions of held-out data.
@@
 \paragraph{Our contributions:}
-\begin{itemize}
-  \item We deliver a description of the fundamental \emph{ontology} and \emph{epistemology} of machine learning, and contrast these with the ontologies and epistemologies of the natural sciences.
+\vspace{-1ex}\begin{itemize}
+  \item We deliver a description of the fundamental \emph{ontology} and \emph{epistemology} of ML, and contrast these with the ontologies and epistemologies of the natural sciences.
   \item We elucidate two important and strong statistical biases that are being introduced to the natural sciences by some uses of ML.
@@
 For our purposes, the natural sciences are about understanding observed phenomena, unifying knowledge, making predictions for new experiments and observations.
-Natural-science questions include things like: How did our Solar System form? Or how do cells differentiate as the embryo grows? Or what causes the jet stream?
+Natural-science questions include things like:
+How did our Solar System form?
+Or how do cells differentiate as the embryo grows?
+Or what causes variations in the jet stream?
 
@@
 These degeneracies are not seen as a problem, since all they do is make the latent weights less interpretable without hurting performance \cite{belkin2019reconciling, bartlett2020benign}.
-Contemporary optimization schemes are stochastic \cite{stochastic} and most models are non-convex.
+Contemporary optimization schemes are stochastic \cite{stochastic, adam} and most models are non-convex.
 The fact that it is impossible to find the global optimum---and the fact that in most contemporary models that isn't even the goal---shows that the latent parameters are not important.
@@
 many tools used in the natural sciences have limited ontologies.
-Perhaps the most extreme case is that of mathematical tools, which are based (in some sense) only on a limited set of axioms.
-Mathematics has a very limited ontology.
+Perhaps the most extreme case is that of mathematical tools, which are based (in some sense) only on a limited set of axioms; mathematics has a very limited ontology.
 ML is useful in science in a way that is analogous to other methodological tools; the point made here is that the particular and specific use of each tool ought to be matched to its ontology and epistemology.
@@
 A recent example in astronomy is the discovery of fast radio bursts \cite{frbs}, which started as outliers in the data taken for imaging projects and have turned out to be interesting astrophysical objects.
-Since expressive unsupervised ML methods can (with some caveats) describe accurately a complex distribution of data, they can also be used to identify (some) rare data points that are unlike any elements of the training data.
+Since expressive unsupervised ML methods can (with some caveats) describe accurately a complex distribution of data, they can also be used to identify (some) rare data points that are unlike any other elements of the training data.
 
 \paragraph{Information questions:}
-If a machine-learning method is correctly trained to predict labels $y$ from data $x$, and it succeeds (better than random, say), then it shows that data $x$ contains information within it about labels $y$.
+If a ML method is correctly trained to predict labels $y$ from data $x$, and it succeeds (better than random, say), then it shows that data $x$ contain information within it about labels $y$.
 Regressions thus can be used to make one-sided measurements of quantities of information-theoretic interest.
@@
 In general, when multiple data point estimates are used jointly, the variance-induced offsets of the estimators average out but the bias-induced offsets remain fixed.
-This is a straightforward point of statistics---this is not news---but it isn't currently informing most of the practice of ML in the natural sciences, where many catalogs and public data products are being produced with ML regressions.
-% AFTER ICML submission: For example, HOGG CITES. HOGG: NOTE that we are self-citing here; we are guilty ourselves!
+This is a straightforward point of statistics---this is not news---but it isn't currently informing most of the practice of ML in the natural sciences, where many catalogs and public data products are being produced with ML regressions \cite{spectrophotometric, leung, aspgap}, sometimes even by us.
 
@@
 Likelihood-based estimators are not usually the lowest-risk estimators, but they can be low- or zero-bias estimators.
-The outputs of regressions are more like posterior-based estimators, affected by the implicit prior set by the training set.
+The outputs of regressions, on the other hand, are more like posterior-based estimators, affected by the implicit prior set by the training set.
 When they are combined, the implicit prior gets amplified.
 Most ML regressions are not capable of generating or emulating likelihoods.
-There are interesting approaches in development to replace standard regressions with generative models that can produce likelihoods \cite{cannon, biwei, likelihood_ratio}.
-
-Importantly, there is nothing unique to ML about low-risk estimators being biased, or that the use of biased estimators in populations analyses is problematic.
-These are standard properties of estimators.
-The unique thing to ML is that standard ML regressions almost always produce biased estimators.
-Given the objective functions (cost functions) in use in ML regressions, this is a feature of ML, not a bug.
+There are interesting approaches in development to replace standard regressions with generative models that can produce likelihoods \cite{cannon, likelihood_ratio, biwei}.
 
@@
 Thus these simulation requirements are getting very large for contemporary research projects.
-For example, in cosmology, contemporary experiments could easily consume the total computing capacity of the United States.
+For example, in cosmology, contemporary experiments could easily consume the total scientific computing capacity of the United States.
 Hence we turn to emulators---trained ML regressions that have learned the input-output relationship of a simulation, or the relationship between an inexpensive low-resolution simulation and an expensive high-resolution experiment.
@@
 We have colleagues who believe that there is no such part; some scientists believe that you must understand deeply every part of your scientific toolkit.
-We disagree: For an example in physics, we use infrared HgCdTe detectors to make extremely precise measurements of intensity fields, without understanding in detail the solid-state physics that makes these devices efficient, linear detectors of photons.
+We disagree: For an example in physics, we use infrared HgCdTe detectors \cite{hgcdte} to make extremely precise measurements of intensity fields, without understanding in detail the solid-state physics that makes these devices efficient, linear detectors of photons.
 There are many parts of a scientific project that must work well, but which do not need to be carefully controlled or understood.
@@
 Many discoveries in astrophysics are discoveries of new kinds of objects, originally found as data outliers \cite{quasars, voorwerp}.
-So far, ML has not been involved in any big discoveries in astronomy, but we expect this to be a productive avenue for ML in the natural sciences.
+So far, ML has not been involved in any big discoveries in astronomy, but we expect this to be a productive avenue for ML in the natural sciences (but see \citealt{contardo}).
 
@@
 In the end, if the results of the investigation are to have the instrument calibrated out, it makes sense to give the instrument model a lot of expressive power, and judge that model in terms of its ability to explain the data.
-Hence, instrument models and calibration are important places for ML methods, and places where the introduction of ML methods can made scientific projects more conservative and more accurate.
+Hence, instrument models and calibration are important places for ML methods, places where the ontology and epistemology of ML are well matched to the objectives, and places where the introduction of ML can made scientific projects more conservative and more accurate.
 One very early example of this in astronomy was the calibration of the \textsl{Sloan Digital Sky Survey} imaging data \cite{ubercalibration}.
@@
 In particular, the biases we highlight in \secref{sec:bad} do not have fixes; in our view ML simply cannot be used in those contexts without leading to substantial scientific mistakes.
-
-In some ways it is surprising that ML has been useful in the natural sciences, since much of the development of ML methods has been in commercial decision-making applications, such as delivering content to users.
-These commercial applications often show a reasonable match between the implicit philosophy of ML and the goals of the commercial ML systems (because, say, serving data to new users is like predicting held-out test data).
-That said, there are connections between commercial applications and scientific applications.
-For example, on the commerce side, there can be important and manageable (and even interpretable) distribution shifts, and, on the scientific side, there are contexts in which the system really will see more of the same (for example, trigger systems in particle-physics colliders).
 

### 2024-05-28 paper/main.tex
camera ready, maybe

@@
-% This document is Copyright 2024 the author(s).
+% This document is Copyright 2024 the authors.
 % It is part of the BadForScience project https://github.com/davidwhogg/BadForScience .
@@
 % To do:
-% ----------------------------
-% - Find the right place to cite the units paper.
-% - audit bibliography; make sure arXiv references in the reference list are consistent? Also capitalization!
-% - Update appendix with feedback from Vedant Chandra (email). In particular, modify model complexity and information content.
-% - Should there be another BAD thing, which is about statistics for comparison of simulations and data? This is a great one; consider after ICML submission.
+% HOGG: address the comment that standard regression is not ML. (emphasized in yellow comment)
 % - Publish on the arXiv.
@@
   These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
-  Here we identify some locations for ML in the natural sciences at which the ontology and epistemology are valuable.
-  For example, when an expressive machine-learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can make the results more trustworthy.
+  Here, we identify some locations for ML in the natural sciences at which the ontology and epistemology are valuable.
+  For example, when an expressive machine learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can make the results more trustworthy.
   We also show that there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
@@
 
-As broad as this definition is, it does not include standard model fitting or parameter estimation.
-These tasks are usually performed in the context of a well-specified model, where the model does not qualitatively change as more data arrive, and the quantitative improvements to the parameter estimates only scale as something like the square root of the size of the data set.
-Standard model fitting and parameter estimation are not ML in our sense.
+As broad as this definition is, it does not include standard model fitting or parameter estimation with a well-specified (first-principles, say) model.
+These tasks are usually performed in the context of a rigid model, with far fewer parameters than data points, in which the model does not qualitatively change as more data arrive, and the quantitative improvements to the parameter estimates only scale as something like the square root of the size of the data set.
+Standard model fitting and parameter estimation in the enormously under-parameterized regime are not ML in our sense.
 
@@
 General relativity \cite{gr} was widely accepted by the community (and very quickly; \citealt{peebles}) not primarily because it explained anomalous data (although it did explain some); it was adopted because, in addition to explaining (a tiny bit of new) data, it also had good structure, it resolved conceptual paradoxes in the pre-existing theory of gravity, and it was consistent with emerging ideas of field theory and geometry.
+
+A recent position piece \citep{donoho2024data} points out that an important aspect of ML's epistemology might be its (aspirational) \emph{reproducibility}.
+Namely, many contributions to the field permit anyone to reproduce the results and build on top of them; the idea is that
+this aspect of ML is part of the reason that the field moves so fast.
+In contrast, it is often argued that the sciences face a reproducibility crisis \cite{baker2016reproducibility};
+the (only) positive spin to put on that is that reproducibility in the natural sciences is not nearly as straightforward as it is in ML.
+This reproducibility discussion is slightly out-of-scope here; we refer the reader to Donoho's piece and the follow-up commentaries (for example, \citealt{Freire2024Singularity}, \citealt{Milanfar2024Data}).
 
@@
 Does this lead to a philosophical connection between ML and the natural sciences?
-We have argued elsewhere (op.cit) that the degeneracies in the natural sciences undermine \emph{realism} or the belief in the literal truth of the latent theories or objects of physics.
+We have argued previously \cite{hogg} that the degeneracies in the natural sciences undermine \emph{realism} or the belief in the literal truth of the latent theories or objects of physics.
 There is an argument to be made that maybe the natural sciences could be better off if they moved at least slightly closer to the ML points of view on ontology and epistemology.

### 2024-05-28 paper/main.tex
passes format checker

@@
 \twocolumn[
-\icmltitle{\documentname: Is machine learning good or bad for the natural sciences?}
+\icmltitle{\textsl{Position:} Is machine learning good or bad for the natural sciences?}
 
@@
 
-\paragraph{Broader societal and ethical impacts:}
-This \documentname{} is directly aimed at producing certain kinds of broader impacts, in the execution of natural-science investigations and projects.
-ML has many roles in the natural sciences.
-This work is intended to start a discussion which, in time, will support the implementation of ML in good roles in scientific projects, and criticize the implementation of ML in less good roles in scientific projects.
-It is also intended to help practitioners understand and respond to biases introduced by the use of ML in those less-good roles.
-The goal of this work is to improve the accuracy and success of the natural sciences.
-However, if the recommendations in this \documentname{} are adopted by scientific communities, the costs of some scientific projects may increase.
-
 \paragraph{Acknowledgements:}
@@
 The Flatiron Institute is a division of the Simons Foundation.
+
+\section*{Impact statement}
+This \documentname{} is directly aimed at producing certain kinds of broader impacts, in the execution of natural-science investigations and projects.
+ML has many roles in the natural sciences.
+This work is intended to start a discussion which, in time, will support the implementation of ML in good roles in scientific projects, and criticize the implementation of ML in less good roles in scientific projects.
+It is also intended to help practitioners understand and respond to biases introduced by the use of ML in those less-good roles.
+The goal of this work is to improve the accuracy and success of the natural sciences.
+However, if the recommendations in this \documentname{} are adopted by scientific communities, the costs of some scientific projects may increase.
 

### 2024-05-28 paper/main.tex
Oh the horror. Finally met all requirements of ICML. Camera ready version.

@@
 \vskip 0.3in
-] % For the arXiv version, move this square bracket below the abstract
+] % For the arXiv version, comment out this square bracket
 
@@
 \end{abstract}
+% ] % For the arXiv version, uncomment this square bracket
 
@@
 We are writing this for an audience that is either developing an ML method for science applications or else bringing ML into a scientific domain.
-The main points of this \documentname{} are \textsl{(1)}~that ML has places where it is very valuable in the contemporary practice of science, and \textsl{(2)}~that it also has places where it will create problems for science.
-That said, we can't definitively answer the question in the title of this \documentname{} for all natural-science domains.
-This paper is a call to natural-science communities to think about it, and answer it for themselves.
+\textbf{
+The main points of this \documentname{} are \textsl{(1)}~that ML has places at which it is very valuable in the contemporary practice of science, \textsl{(2)}~that ML also has places at which it will create problems for science, and \textsl{(3)}~currently it is not obvious what is the epistemic role of ML within the broader goals of the natural sciences.
+%That said, we can't definitively answer the question in the title of this \documentname{} for all natural-science domains.
+This paper is a call to natural-science communities to think about it, and answer the question in the title for themselves.}
 From the perspective of the physical sciences, our answer is ``Both.''

### 2024-05-31 paper/main.tex
new camera-ready version, taking some comments

@@
 % Hogg typesetting issues that won't conflict with ICML
-\frenchspacing
+\raggedbottom\frenchspacing
 \newcommand{\documentname}{\textsl{Position Paper}}
@@
 
-\printAffiliationsAndNotice{}
 \begin{abstract}
@@
   These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
-  Here, we identify some locations for ML in the natural sciences at which the ontology and epistemology are valuable.
+  Here we identify some locations for ML in the natural sciences at which the ontology and epistemology are valuable.
   For example, when an expressive machine learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can make the results more trustworthy.
   We also show that there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
-  For one, when ML models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
+  For one, when ML models are used to emulate physical (or first-principles) simulations, they amplify confirmation biases.
   For another, when expressive regressions are used to label datasets, those labels cannot be used in downstream joint or ensemble analyses without taking on uncontrolled biases.
@@
 % ] % For the arXiv version, uncomment this square bracket
+\printAffiliationsAndNotice{}
 
@@
 
-This definition is broad:
-In addition to methods like convolutional neural networks \cite{cnn}, multi-layer perceptrons, and transformers \cite{transformer}, it includes large linear regressions, gaussian processes \cite{gp}, support-vector machines \cite{svm}, principal components analysis \cite{pca}, kernel density estimates \cite{kde}, and even some kinds of multilayer or hierarchical models \cite{multilevel}; indeed almost any models in any current ML textbooks (eg, \citealt{ml_book1, ml_book2}).
+This definition is broad, perhaps controversially broad:
+In addition to methods like convolutional neural networks \cite{cnn}, multi-layer perceptrons, and transformers \cite{transformer}, it includes large linear regressions, Gaussian processes \cite{gp}, support-vector machines \cite{svm}, principal components analysis \cite{pca}, kernel density estimates \cite{kde}, and even some kinds of multilayer or hierarchical models \cite{multilevel}; indeed almost any models in any current ML textbooks (eg, \citealt{ml_book1, ml_book2}).
 Feel free to personally take any more restrictive definition.
@@
 Most ML regressions are not capable of generating or emulating likelihoods.
-There are interesting approaches in development to replace standard regressions with generative models that can produce likelihoods \cite{cannon, likelihood_ratio, biwei}.
+There are interesting approaches in development to replace standard regressions with generative models that can produce likelihoods \cite{cannon, likelihood_ratio, sequential, biwei}.
 
@@
 
-Finally, we note that one objection to this confirmation-bias point is that perhaps it exists just as strongly for any simulation-based inference?
+Finally, we note that one objection to this confirmation-bias point is that perhaps it exists just as strongly for any simulation-based inference (such as approximate bayesian computation; \citealt{abc})?
 After all, simulations are expensive, so the full suite of them probably can't be affordably repeated.
@@
 Our view is that the promising directions include imposing physically motivated mathematical structure into the design or regularization of the ML models.
-This includes symmetries, conservation laws, and constraints \cite{bronstein2021geometric, villar2023towards}, but also ODE/PDE solvers \cite{karniadakis2021physics}, dynamic programming modules \cite{xu2020can}, and other established mathematical tools.
+This includes symmetries, conservation laws, and constraints \cite{bronstein2021geometric, villar2023towards}, but also ODE and PDE solvers \cite{karniadakis2021physics}, dynamic programming modules \cite{xu2020can}, and other established mathematical tools.
 The research direction that looks towards explainability, interpretability \cite{interpretable}, and trustworthiness \cite{kearns2019ethical} is also fundamental to the development of ML methods for the physical sciences.
@@
   Rix,
+  Ilya Shpitser (JHU),
   Kate Storey-Fisher (DIPC),
+  Jeremias Sulam (JHU),
+  Jesse Thaler (MIT),
+  Sebastian Wagner-Carena (Flatiron),
   Kaze Wong (Flatiron), and
@@
 The data generated this way are shown in the top panels of \figref{fig:regression}.
-The data contain a non-trivial relationship between label $y$ (measured age) $y$ and known parameter $r$ (guiding radius).
+The data contain a non-trivial relationship between label $y$ (measured age) and known parameter $r$ (guiding radius).
 This relationship is shown in \figref{fig:regression}.
